---
title: "OR Multi Disturbance"
author: "Michael Dietze"
date: "12/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("disturbance.R")
source("multi.ens.adj.R")
source("multi.analytical.ens.adj.R")
source("StatsAuxiliary.R")
library(BayesianTools)
library(magrittr)
```

## set up

```{r}
## Load disturbance rate, state&transition, disturbed biomass stats
load("STprior.RData")
load("extract.h03v04.RData")
load("dis.postprocessed.RData")
dtime <- 1985:2012
colnames(dextract) <- paste0("t",dtime)
dextract[is.na(dextract)] <- 0
d.prior["0"] = d.prior["0"] + d.prior[6]                         ## change at runtime (removes <NA>)
d.prior = d.prior[-6]

## Load Calibration
load("fit.RData")
Qsoil = Qs
```

## drivers

```{r}
latlon = points@coords # lat / lon of each site in dextract 
metFile = "daymet.RData"
if(file.exists(metFile)){
  load(metFile)
} else {
  dm = list()
  for(i in 1:nrow(latlon)){
    dm[[i]] = daymetr::download_daymet(site="h03v04",latlon[i,2],latlon[i,1],start=1985,end=2017,internal=TRUE,simplify=TRUE)
  }
  save(dm,file=metFile)
}

## extract Shortwave and convert to daily PAR
SW2PAR = 0.45
srad = subset(dm[[1]],measurement == "srad..W.m.2.")   
metTime <- as.POSIXct(paste(srad$year,srad$yday,sep="-"),format="%Y-%j")
rad <- matrix(NA,length(dm),length(metTime)) # [number of sites x number of measurements]
print(dim(rad))
for(i in seq_along(dm)){
  srad = subset(dm[[i]],measurement == "srad..W.m.2.")   
  rad[i,] = srad$value*1e-6*86400*SW2PAR  # -> MJ /m2 /day PAR (convert) # QUESTION -- should the changing day length, from the daymet day length values, be used (instead of a fixed 24-hr day length)?
}
plot(metTime,rad[1,],type='l')
## plot(date,PAR,col=2) ## calibration data
```

## functions

```{r,echo=FALSE,results='hide',message=FALS}
##' Single Site, Single Disturbance
##' @param Y observed data
##' @param Rparams 2 element list [slope, intercept] mapping data value to observation error (R) value
##' @param Q process error
##' @param Xic state initial condition matrix
##' @param THETA parameter matrix
##' @param dist  disturbance list: p.dist, mu0[3], V0[3,3], alloc soil
##' @param updateQ if true, conditionally update the process error Q based on observations
##' @param useAdjustment if true, use ensemble adjustment. If false, just sample from the posterior.
##' @param use_analytical_posterior if true use analytical posterior, if false use MCMC to sample from the posterior
SSSD.model <- function(Y,Rparams,Q,PAR,Xic,THETA,dist,updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE){
  
  ## init
  NT = length(PAR)/365
  Nf = Nmu = Na = Xic
  ne = nrow(Xic)
  muF <- array(NA, dim = c(dist$n + 1, 3))
  CF <- array(NA, dim = c(dist$n + 1, 3, 3))
  pF <- array(NA, dim = c(dist$n + 1, 3, 3))
  muA <- array(NA, dim = c(dist$n + 1, 3))
  CA <- array(NA, dim = c(dist$n + 1, 3, 3))
  wA <- array(NA, dim = c(dist$n + 1))
  if(!is.matrix(THETA)) THETA = matrix(THETA,nrow=ne,ncol=length(THETA),byrow = TRUE)
  Forecast = Analysis = Debug = list()
  ProcessErrors = list()
  
  AnalysisSSMDMP <- "
    model{
      ## Forecast priors
      for(i in 1:N){
        x[1:3, i] ~ dmnorm(muf[i,], pf[i,,])
      }
    
      ## --- Categorical disturbance selection without dmulti ---
      ## rho[1:N] is one-hot with probabilities p[1:N]
    
      # First category
      rho[1] ~ dbern(p[1])
    
      # Intermediate categories
      for(k in 2:(N-1)){
        z[k] ~ dbern(p[k] / (1 - sum(p[1:(k-1)])))
        rho[k] <- z[k] * (1 - sum(rho[1:(k-1)]))
      }
    
      # Last category (absorbs remaining probability)
      rho[N] <- 1 - sum(rho[1:(N-1)])
      
      ## --- Analysis --- 
      for(j in 1:3){ 
        X[j] <- sum(rho[1:N] * x[j,1:N]) 
      }
      
      # Model observation
      Y[3] ~ dmnorm(X[3], R)
    }
  "
  
  
  a_t = 1/0.1172475 # place a prior on process precision: (1/Q) ~ Gamma(a_0, b_0).
  b_t = 1
  ignoreyears <- c(1, 15)
  
  for(t in 1:NT){
    
    ## update IC
    THETA[,9:11]  <- Na    ## Multinomial
    
    # print("starting forecast...")
    
    ## Forecast 
    for(i in 1:ne){
      tsel    = (t-1)*365 + 1:365             ## Select days from driver files
      Nmu[i,] = VSEM(THETA[i,], PAR = PAR[tsel])[365,2:4] ## deterministic simulation
    }
    
    ## update process error (if we have observed data)
    ## assume Y ~ N(mu, Q). We will pick Q using the posterior: Q = 1/(E[p((1/Q) | y)]) = 1/E[Gamma(a_t, b_t)] = b_t / a_t
    if (!is.na(Y[t,3]) && updateQ && length(which(ignoreyears == t)) == 0) { # filter out disturbance years / start year (where ensemble is noisy )
      a_t <- a_t + 0.5
      b_t <- b_t + 0.5 * (Y[t,3] - colMeans(Nmu)[3])^2
      Q[3, 3] = b_t / a_t
    }
    ProcessErrors[[t]] <- Q[3, 3]
    
    ## Add process error
    for(i in 1:ne) {
      # Nf[i, ] <- Nmu[i, ]
      Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
    }
    
    
    ## Disturbance forecast model
    d.ens.p <- c(0.5, rep(0.5 / dist$n, dist$n))
    d.forecast <- rmultinom(ne, 1, d.ens.p)
    d.class <- apply(d.forecast,2,which.max) - 1
    sel.undisturbed <- !as.logical(d.class)
    
    for (d in 1:(dist$n)) {
      sel.dist = which(d.class == d)
      if(sum(!sel.undisturbed)>2) {
        for(j in sel.dist){
          Nf[j,] = disturbance.p(Nf[j,],
                      mu0=dist$vals[[d]]$mu0,
                      V0=dist$vals[[d]]$V0, 
                      alloc.soil=dist$alloc.soil)
        }
        muF[d+1,]  = colMeans(Nf[sel.dist,])
        CF[d+1,,] = var(Nf[sel.dist,])
        pF[d+1,,] = try(solve(CF[d+1,,]))
        # print(paste("precision, class: ", d, "value: ", pF[d+1,1,1]))
        # print(class(pF[d+1,1,1]))
      } 
      # backup values if the ensemble stats fail
      
      if(sum(!sel.undisturbed) <= 2 || "try-error" %in% class(pF[d+1,,]) || grepl("Error", pF[d+1, 1,1])) {
        muF[d+1,] = c(dist$vals[[d]]$mu0[1], mean(Nf[,2]),  dist$vals[[d]]$mu0[2])   
        CF[d+1,,] = diag(c(dist$vals[[d]]$V0[1,1],0.25,dist$vals[[d]]$V0[2,2]))
        pF[d+1,,] = solve(CF[d+1,,])
        print(paste("iteration: ", d, "class: ", class(pF[d+1,1,1]), "value: ", pF[d+1,1,1]))
      }
      
      pF <- array(as.numeric(pF), dim = dim(pF), dimnames = dimnames(pF))
      
      
      
    }
    Nf <- tobit(Nf)
    muF[1,] = colMeans(Nf[sel.undisturbed,])
    CF[1,,] = var(Nf[sel.undisturbed,])
    pF[1,,] = solve(CF[1,,])
    
    # save forecast stats
    mufd <- colMeans(Nf[!sel.undisturbed,])
    mufn <- colMeans(Nf[sel.undisturbed,])
    qs <- c(0.025,0.25,0.5,0.75,0.975)
    cid <- apply(Nf[!sel.undisturbed,], 2, quantile, qs)
    cin <- apply(Nf[sel.undisturbed,], 2, quantile, qs)
    Forecast[[t]] <- list(muD = mufd, muN = mufn, ciN = cin, ciD = cid)
    
    ## Analysis
    pD = as.numeric(dist$p[,t])
    H = which(!is.na(Y[t,]))
    
    if (length(H) == 1) { ## data observed, run filter
      run.adjustment = TRUE
      H = as.numeric(H[1])
      mat_H = matrix(0, 1, 3)
      mat_H[H] <- 1
      R_H <- Y[t,H] * Rparams[1] + Rparams[2] # compute heteroskedastic observation error
      
      if (use_analytical_posterior) { ## use analytical posterior
        for (d in 1:(dist$n + 1)) {
          muf <- muF[d,]
          Cf <- CF[d,,]
          pf <- pF[d,,]
          CA[d,,] <- solve(pf + t(mat_H) %*% ((1/R_H) * mat_H))
          muA[d,] <-  CA[d,,] %*% (pf %*% muf + (t(mat_H) * (1/R_H) * Y[t,H]))
          lik <- dnorm(Y[t,H], mean=mat_H %*% muf, sd = sqrt(mat_H %*% Cf %*% t(mat_H) + R_H))
          wA[d] <- lik * pD[d]
        }
        
        if (sum(wA) > 0) {
          wA <- wA / sum(wA) 
        } else {
          wA <- rep(1/"nclass", "nclass")
        }
        
        # -------------------------------
        # test: compute MCMC posterior approximation
        # -------------------------------
        
        ## number of disturbance types
        # N <- nrow(muF)
        # 
        # ## normalize disturbance probabilities
        # p.model <- dist$p[, t]
        # p.model <- p.model / sum(p.model)
        # 
        # ## data list passed to JAGS
        # update <- list(
        #   N   = N,
        #   p   = p.model,
        #   muf = muF,   # [3 x N]
        #   pf  = pF,    # [3 x 3 x N]
        #   Y   = Y[t,],
        #   R   = R_H
        # )
        # 
        # ## initialize and adapt model
        # mod <- rjags::jags.model(
        #   file      = textConnection(AnalysisSSMDMP),
        #   data      = update,
        #   n.adapt   = 1000,
        #   n.chains  = 3,
        #   quiet     = TRUE
        # )
        # 
        # ## draw posterior samples
        # jdat <- rjags::coda.samples(
        #   mod,
        #   variable.names = c("X", "rho", "x"),
        #   n.iter = 50000
        # )
        # 
        # save(jdat, file="debugging-or.RData")
        # 
        # ## convert to matrix
        # dat <- as.matrix(jdat)
        # 
        # ## extract posterior samples
        # Xsel   <- grep("X", colnames(dat))
        # rhosel <- grep("rho", colnames(dat))
        # 
        # Xa  <- dat[, Xsel]     # analysis state samples
        # rho <- dat[, rhosel]  # disturbance indicator samples
        # ca <- apply(rho,1,which.max)
        # 
        # muXa = colMeans(Xa)
        # print("overall mean MCMC...")
        # print(muXa)
        # 
        # print("undisturbed mean MCMC")
        # rows.undist <- dat[,"rho[1]"] > 0
        # print(colMeans(Xa[rows.undist,]))
        # 
        # print("disturbed mean MCMC")
        # rows.dist <- dat[,"rho[1]"] == 0
        # print(colMeans(Xa[rows.dist,]))
        # 
        # print("Y value...")
        # print(Y[t, 3])
        
        
      } else { ## run MCMC
        
        # 
        # print("computed disturbed mean from MCMC")
        # print(colMeans(Xa[ca==1,]))
        # print("computed undisturbed mean from MCMC")
        # print(colMeans(Xa[ca==0,]))
        # print("computed disturbed mean from MCMC (directly from muD samples)")
        # XDa = dat[,grep(pattern="^muD", x=colnames(dat))]
        # print(colMeans(XDa))
        # print("computed 50% quantile of muD samples")
        # print(apply(XDa,2,quantile,c(0.5)))
        # print("computed undisturbed mean from MCMC (directly from muN samples)")
        # XNa = dat[,grep(pattern="^muN", x=colnames(dat))]
        # print(colMeans(XNa))
        # print("computed 50% quantile of muN samples")
        # print(apply(XNa,2,quantile,c(0.5)))
        # print("mufD:")
        # print(mufD)
        # print("mufN")
        # print(mufN)
      }
    } else { # if we don't have observed data
      Na = Nf # analysis = forecast
      ca = 1
      run.adjustment = FALSE
    }
      ## draw new IC
    if(run.adjustment){
      if(useAdjustment && use_analytical_posterior) { 
        ## run ensemble adjustment
        Na <- multi.analytical.ens.adj(
            Xf   = Nf,
            cf   = d.class,
            mu.f = muF,
            Pf   = CF,  # covariance matrices for undisturbed / disturbed ensemble states
            Pp.mu.a = muA,
            Pp.Pa = CA,
            Pp.w = wA,
            uc = 0:dist$n)
        Ca <- Na[,1] ## updated class assignments
        Na <- tobit(Na[,-1])
      } else if (!useAdjustment && use_analytical_posterior) {
        ## sample directly from the posterior
        Ca <- rbinom(nrow(Na), size=1, prob=pD_star)
        Pp.mu.a <- t(cbind(m_starN, m_starD))
        Pp.Pa <- abind::abind(C_starN, C_starD, along=3)
        for (i in 1:ne) {
          distclass <- Ca[i] + 1
          Na[i,] <- mvtnorm::rmvnorm(1, mean=Pp.mu.a[distclass,], sigma=Pp.Pa[,,distclass])
        }
        Na <- tobit(Na)
      } else if (useAdjustment && !use_analytical_posterior) {
        Na <- multi.ens.adj(
          Xf = Nf,
          cf = d.forecast,
          mu.f = rbind(mufN, mufD),
          Pf   = abind::abind(CfN,
                   CfD,
                   along=3),
          Xa = Xa,
          ca = ca
        )
        Ca <- Na[,1]
        Na <- tobit(Na[,-1])
      } else {
        stop("useAdjustment = FALSE and use_analytical_posterior = FALSE is not supported")
      }
      
    } else {
      Na <- tobit(Na)
      Ca <- rep(0, ne) ## dummy data
    }
    
    ## Compute post-analysis means and CIs
    if (run.adjustment) { ## if we did run analysis (since we had data)
      if (use_analytical_posterior) { ## if we used analytical posterior
        AnalysisCIs <- matrix(NA, nrow=5, ncol=3)
        quantiles <- c(0.025,0.25,0.5,0.75,0.975)
        
        # compute analysis CIs from most likely component
        ci.idx <- which.max(wA)
        for (i in 1:3) {
          AnalysisCIs[,i] <- normal_quantile(quantiles, muA[ci.idx, i], CA[ci.idx, i, i])
        }
        
        # compute overall mean
        mu.post = rep(0, 3)
        for (i in 1:(dist$n + 1)) {
          mu.post <- mu.post + wA[i] * muA[i,]
        }
        
        # compute mean of just disturbed components
        mu.post.D = rep(0, 3)
        wA.sumdist <- sum(wA[2:dist$n+1])
        for (d in 2:(dist$n + 1)) {
          mu.post.D <- mu.post.D + (wA[d] / wA.sumdist) * muA[d,]
        }
        
        # mean of undisturbed
        mu.post.N = muA[1, ]

        prob.dist <- sum(wA[2:(dist$n + 1)])
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = AnalysisCIs,
                              mu = mu.post,
                              mu.D = mu.post.D,
                              mu.N = mu.post.N,
                              probD = prob.dist
                              )
        
      } else { ## if we used MCMC
        mu.post.D <- sum(ca == 1, na.rm = TRUE) / length(ca)
        # print(paste0("year ", t))
        # print("means of Xa")
        # print(colMeans(Xa))
        # print("0.5 quantiles of Xa")
        # print(apply(Xa,2,quantile,c(0.5)))
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = apply(Xa,2,quantile,c(0.025,0.25,0.5,0.75,0.975)),
                              m_starN = colMeans(Xa[ca==0,]),
                              m_starD = colMeans(Xa[ca==1,]),
                              muD = mu.post.D)
      }
    } else { ## if we didn't have data
      AnalysisCIs <- apply(Na,2,quantile,c(0.025,0.25,0.5,0.75,0.975))
      mu.post <- apply(Na,2,mean,na.rm=TRUE)
      mu.post.D <- 0.1
      if (use_analytical_posterior) {
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = AnalysisCIs,
                              mu = mu.post,
                              mu.D = 0.1,
                              probD = 0)
      } else {
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = AnalysisCIs,
                              mu = mu.post,
                              muD = mu.post.D)
      }
    }
  }
  return(list(Forecast=Forecast,Analysis=Analysis,Y=Y, ProcessErrors=ProcessErrors, Debug=Debug))
}
```

```{r}
load("debugging-or.RData")

# pf <- matrix(
#   as.numeric(pf),
#   nrow = nrow(pf),
#   ncol = ncol(pf)
# )
# solve(pf + t(mat_H) %*% ((1/R_H) * mat_H))

```

## Load disturbance parameters

```{r}
dist <- list()
dist_name <- c(
  "0"  = "none",
  "10" = "other",
  "20" = "cut",
  "30" = "road",
  "40" = "fire",
  "50" = "biotic"
)

dist.forecastparams <- list()

count <- 1
for (dclass in names(dist.priors)) {
  dist.name <- as.character(dist_name[dclass])
  nclust <- length(dist.priors[[dclass]]$means)
  
  for (i in 1:nclust) {
    dist.forecastparams[[count]] <- list()
    dist.forecastparams[[count]]$mu0 <- c(0.2, dist.priors[[dclass]]$means[i])
    dist.forecastparams[[count]]$V0 <- diag(c(0.04, dist.priors[[dclass]]$vars[i]))
    dist.forecastparams[[count]]$name <- paste0(dist.name, i)
    count <- count + 1
  }
}
```


## initial conditions & observations

```{r}
## BIOMASS
B2C = 0.5 ##biomass to carbon ratio
Yw <- bextract*B2C*1000*1e-4 ## convert Mg/ha (Mg per hectacre)  to  kg C [356 (num ensemble members) x 28 (num years observed)]
Rw <- (84*B2C*1000*1e-4)^2 ## Observation Error: Kennedy et al 2018 Environ. Res. Lett. 13 025004; Fig 2 RMSE
## units: Mg/ha -> kg/m2
print(dim(Yw))
## grab the distribution of AGB for the first time point from all pixels
AGBprior <- Yw[,1]
ne = length(AGBprior) ## number of ensemble members

## spin up
PARlong <- as.vector(t(rad)) # concatenate PAR time series for each site into one vector
spin = VSEM(theta,PARlong) # Question: why run VSEM on the concatenated PAR? What is theta?
thin = seq(nrow(spin)/2,nrow(spin),length=ne)
spinIC = spin[thin,2:4] # select ne samples from the spin-up 
#for(i in 2:4){
#  hist(spin[thin,i],type='l')
#}

## disturbance recovery trajectory
nNBG = 200 ## years
NBG <- array(NA,c(ne,nNBG,3))
for(i in 1:ne){
  param = theta
  param[9:11] <- spinIC[i,]*c(0.01,1,0.01) ## set leaf and wood to 1% (simulate disturbance)
  raw = VSEM(param,PARlong[1:(nNBG*365.25)]) # simulate for nNBG years
  NBG[i,,] = raw[round(seq(365,nNBG*365.25,length=nNBG)),2:4]  ## grab pools at the end of each year
}
  
## for each AGBprior, grab the matching leaf and soil
ICprior <- matrix(NA,ne,3)
age = rep(NA,ne)
for(i in 1:ne){
  age[i] <- which.min((AGBprior[i]-NBG[i,,3])^2) # age = year in the disturbance recover trajectory where the simulated site has similar AGB to the AGB prior information
  ICprior[i,] <- c(NBG[i,age[i],1:2],AGBprior[i]) # use AGBprior for the model ICs
}
hist(age)
colnames(ICprior) <- c("Cv","Cs","CR") 
pairs(ICprior)

```

### wrapper function to run data assimilation

```{r}
##' @description runDA: function to run data assimilation given R / Q / P multipliers
##' @param Rparams[2] -- A list of the form [m, b] containing the slope / intercept of observation error line. 
##'                      R = m * [measured wood carbon (kg / m^2)] + b
##' @param useDistProduct -- if TRUE, use incorporate disturbance product predictions into priors. If FALSE, use a                                uniform prior across all years
##' @param Qmult -- multiplier for process error, Q
##' @param Pmult -- multiplier for disturbance probability
##' @param updateQ if true, conditionally update the process error Q based on observations
##' @param useAdjustment
##' @param use_analytical_posterior
##' @param siteIndices indices of Oregon sites to run with
##' @param expName name of experiment to use in output file names
runDA <- function(Rparams, useDistProduct, updateQ, useAdjustment, use_analytical_posterior, siteIndices, expName) {
  for(i in siteIndices){
    ofile = paste0(
        "experiments/final/",
        expName, "_OR_SD_",
        formatC(i, width = 3, format = "d", flag = "0"),
        ".RData"
    )
    
    print(ofile)
    NT = floor(ncol(rad)/365)
    Yi = matrix(NA,NT,3)  ## observed data
    Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed (Question: where do Yi observations come from?)
    Q = diag(c(Ql,Qsoil,Qw)^2)  ## process error
    
    
    # dtype = which(names(d.prior) == dextract[i,"t2004"]) ## 0 = none, 10 = other, 20 = cut, 30 = road, 40 = fire, 50 = biotic
    # d.class.str = as.character(dextract[i,"t2004"])
    # d.class.mu0 = dist.forecastparams[[d.class.str]]$mu0
    # d.class.V0 = dist.forecastparams[[d.class.str]]$V0
    # if(length(dtype)==0) break
    # p.dist = rep(d.prior[dtype],NT) ## set background rate
    # if (useDistProduct) {
    #   users_accuracy = c(1-d.prior[1],.77,.92,.77,.78,.78) #Kennedy et al 2015 table 6
    #   has.dist = which(dextract[i,] > 0)
    #   p.dist[has.dist] = users_accuracy[dextract[i,has.dist]/10 + 1]
    # }
    
    # load disturbance class priors
    n.dist = length(dist.forecastparams)
    p.dist = array(NA, dim=c(n.dist + 1, NT))
    distname.toidx <- list("other"=2, "cut"=3, "fire"=4, "biotic"=5)
    dist.prior.probs <- sapply(dist.forecastparams, function(x) {
      name <- x$name
      for (d in names(distname.toidx)) {
        if (grepl(d, name)) {
          return(d.prior[distname.toidx[[d]]])
        }
      }
    })
    
    # print(length(dist.prior.probs))
    # print(nrow(p.dist))
    # print(dist.prior.probs)
    for (t in 1:NT) {
      p.dist[1:(n.dist+1),t] <- as.numeric(c(d.prior[1], unlist(dist.prior.probs)))
    }
    
    if (useDistProduct) {
      users_accuracy = c(1-d.prior[1],.77,.92,.77,.78,.78)
      for (d in seq(names(dist.priors))) {
        d.name <- names(dist.priors)[d]
        has.dist <- which(dextract[i,] == d.name)
        p.dist[d+1, has.dist] <- users_accuracy[dextract[i,has.dist]/10 + 1]
      }
    }
    
    disturb.params <- list(              
      p = p.dist,
      n = n.dist,
      vals = dist.forecastparams,
      alloc.soil = 0.05        ## made up  
    )
    test = SSSD.model(Y = Yi,Rparams=Rparams,Q=Q,
                    PAR=rad[i,],Xic=ICprior,THETA=theta[1:11],
                    dist = disturb.params, useAdjustment=useAdjustment, updateQ=updateQ, 
                    use_analytical_posterior = use_analytical_posterior)
    save(test,file=ofile)
  }
}
```

## Compare results with GEDI Mean / Landtrendr mean, with and without observation error heteroskedasticity

```{r}
od = c(83:176,264:356)
fullRun = c(224:356)
debug = c(90, 91, 94, 104, 106, 125, 152, 153, 165, 175) # high abs dist, low dist prob
testRun = seq(551, ne, by=10)

RGEDImean <- 0.8848249 # kg / m^2
RLandTrendrmean <- 3.79986 # kg / m^2
RparamsGEDIfixed <- c(0, RGEDImean)
RparamsLandTrendrfixed <- c(0, RLandTrendrmean)
RparamsGEDIheteroskedastic <- c(0.01331, 0.09761) # from regular regression (m_G, b_G)

# 0.5 quantile regression line: c(0.01159707, 0.04660497)

# compute intercept for Landtrendr heteroskedastic params
Cwmean <- (RGEDImean - RparamsGEDIheteroskedastic[2]) / RparamsGEDIheteroskedastic[1] # m_G * Cwmean + b_G = RGEDImean
Landtrendrslope <- RparamsGEDIheteroskedastic[1] # set m_L = m_G
Landtrendrintercept <- RLandTrendrmean - Landtrendrslope * Cwmean # m_L * Cwmean + b_L = RLandTrendrmean
RparamsLandtrendrheteroskedastic <- c(Landtrendrslope, Landtrendrintercept) # (m_L, b_G)

# # experiment Rgedi_Hfalse_Dfalse
# runDA(Rparams=RparamsGEDIfixed, useDistProduct = FALSE, updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rgedi_Hfalse_Dfalse") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)
# 
# # experiment Rlandtrendr_Hfalse_Dfalse
# runDA(Rparams=RparamsLandTrendrfixed, useDistProduct = FALSE, updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rlandtrendr_Hfalse_Dfalse") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)
# 
# # experiment Rgedi_Htrue_Dfalse
# runDA(Rparams=RparamsGEDIheteroskedastic, useDistProduct = FALSE, updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rgedi_Htrue_Dfalse") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)

# experiment Rlandtrendr_Htrue_Dfalse
# runDA(Rparams=RparamsLandtrendrheteroskedastic, useDistProduct = FALSE, updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rlandtrendr_Htrue_Dfalse") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)

# experiment Rgedi_Hfalse_Dtrue
# runDA(Rparams=RparamsGEDIfixed, updateQ=TRUE, useDistProduct = TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rgedi_Hfalse_Dtrue") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)

# experiment Rlandtrendr_Hfalse_Dtrue
# runDA(Rparams=RparamsLandTrendrfixed, updateQ=TRUE, useDistProduct = TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rlandtrendr_Hfalse_Dtrue") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)

# experiment Rgedi_Htrue_Dtrue
runDA(Rparams=RparamsGEDIheteroskedastic, updateQ=TRUE, useDistProduct=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rgedi_Htrue_Dtrue") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)

# experiment Rlandtrendr_Htrue_Dtrue
# runDA(Rparams=RparamsLandtrendrheteroskedastic, useDistProduct = TRUE, updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=fullRun, expName="Rlandtrendr_Htrue_Dtrue") # R = (gedi, landtrendr), H = heteroskedastic = (true, false), D = disturbance product = (true, false)

```

```{r}
pool.names = c("Cl, Cs, Cw")
for (site in 100:356) {
  runfile <- paste0("Rgedi_Htrue_Dtrue_OR_SD_", sprintf("%03d", site), ".RData")
  load(paste0("./experiments/final/", runfile))
  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
      run.meanOverall = sapply(X = test$Analysis, function(x) {
        y = x$mu
        y[i]
      })
      run.muN = sapply(X = test$Analysis, function(x) {
        y = x$mu.N
        y[i]
      })
      run.muD = sapply(X = test$Analysis, function(x) {
        y = x$mu.D
        y[i]
      })
      run.mufD = sapply(X = test$Forecast, function(x) {
        y = x$muD
        y[i]
      })
      run.ciD = sapply(X = test$Forecast, function(x) { 
        y = x$ciD
        y[,i]
      })
      run.mufN = sapply(X = test$Forecast, function(x) {
        y = x$muN
        y[i]
      })
      run.ciN = sapply(X = test$Forecast, function(x) { 
        y = x$ciN
        y[,i]
      })
      run.probD = sapply(X = test$Analysis, function(x) { 
        y = x$probD
        y
      })
      # run.mufN = sapply(X = test$Forecast, function(x) {
      #   y = x$mufN
      #   y[i]
      # })
      # run.mufD = sapply(X = test$Forecast, function(x) {
      #   y = x$mufD
      #   y[i]
      # })
      # 
      # run.probD = sapply(X = test$Analysis, function(x) {
      #   y = x$muD
      # })
      # 
      # CI.NfD = sapply(X = test$Forecast, function(x) {
      #   y = x$CID
      #   y[,i]
      # })
      # 
      # CI.NfN = sapply(X = test$Forecast, function(x) {
      #   y = x$CIN
      #   y[,i]
      # })
      # 
      # Debug.muN = sapply(X = test$Debug, function(x) {
      #   y = x$mcmcMuN
      #   y[i]
      # })
      # 
      # Debug.muD = sapply(X = test$Debug, function(x) {
      #   y = x$mcmcMuD
      #   y[i]
      # })
      # 
      # Debug.pD = sapply(X = test$Debug, function(x) {
      #   x$mcmcPd
      # })


      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=pool.names[i],xlab="time", main=runfile)

    ecoforecastR::ciEnvelope(1:ncol(CI.Na),
                             CI.Na[1,], CI.Na[5,],
                             col=ecoforecastR::col.alpha('#b2df8a',0.7))

    # overall mean
    lines(unlist(run.meanOverall), col="#40826D", lwd=2)

    # μD line
    lines(unlist(run.muD), col="#002EFF", lwd=2)
    
    # undisturbed mean analysis
    lines(unlist(run.muN), col="#f38116", lwd=2)
    
    # forecast mean disturbed
    lines(unlist(run.mufD), col="#FF002E", lwd=2)
    
    # forecast mean undisturbed
    lines(unlist(run.mufN), col="#91CB8F", lwd=2)
    
    # print(Debug.muD)
    # mcmc muD
    # lines(unlist(Debug.muD), col="#FF002E", lwd=2)
    # mcmc muN
    # lines(unlist(Debug.muN), col="#002EFF", lwd=2)
    
    
    
    # forecast CID
    # ecoforecastR::ciEnvelope(1:ncol(run.ciD),
    #                          run.ciD[1,], run.ciD[5,],
    #                          col=ecoforecastR::col.alpha('#e86666',0.7))
    # 
    # # forecast CIN
    # ecoforecastR::ciEnvelope(1:ncol(run.ciN),
    #                          run.ciN[1,], run.ciN[5,],
    #                          col=ecoforecastR::col.alpha('#80bcf3',0.7))

    # Observations
    points(test$Y[,i], pch = 1)

    # === ADD LEGEND HERE ===
    legend("topright",
           legend = c("mean (overall)", "mean (disturbed)", "mean (undisturbed)", "forecast mean (disturbed)", "forecast mean (undisturbed)"),
           col    = c("#40826D", "#002EFF", "#f38116", "#FF002E", "#91CB8F"),
           lwd    = rep(2, 5),
           pch    = rep(NA, 5),
           bty    = "n")
    # ========================
    plot(run.probD, main=paste0(runfile, " disturbance probs"))
    
    # legend("topright",
    #        legend = c("MCMC", "Analytical"),
    #        col    = c("#FF002E", "#9D2933"),
    #        lwd    = c(2, 2),
    #        pch    = c(NA, NA),
    #        bty    = "n")
    # lines(run.probD, col="#9D2933", lwd=2)
    # lines(Debug.pD, col="#FF002E", lwd=2)
  }
}
```

## postprocessing

```{r}
## Detect set of experiments that have been run / sites that have been experimented with
runs <- dir(path="./experiments/final", pattern="OR_SD")
sites <- sapply(strsplit(runs, "_", fixed=TRUE), function(x) {
  site <- x[6]
  site <- as.numeric(sub(".RData", "", site))
})
exp = sapply(strsplit(runs,"_",fixed=TRUE),function(x){
  x <- paste0(x[1:3], collapse="_")
})
exps <- unique(exp)
print(exps) # the combos of P / Q / R multipliers experimented with
# load experiments that haven't yet been summarized
elist <- list()   ## run once
t04 = 15            ## time of disturbance assessement (2004)
tlo = 28            ## time of last observation
for(i in seq_along(exps)){
  if(is.null(elist[[exps[i]]])){
    esel <- which(exp==exps[i])
    distrate = rep(NA,nrow(dextract)) # 1 x (number of sites)
    divergence = predist = distsize = matrix(NA,nrow(dextract),6) ## pre and post disturbance biomass pools (w/uncertainty) nsite x 6
    for(j in seq_along(esel)){
      ## load file
      load(paste0("./experiments/final/", runs[esel[j]]))
      k = sites[esel[j]]
      #print(k)
      ## extract detection rate
      distrate[k]=test$Analysis[[t04]]$probD # posterior probability of disturbance

      ## capture of disturbance change
      predist[k,1]    = test$Analysis[[t04-1]]$mu[3]
      predist[k,2:6]  = test$Analysis[[t04-1]]$CI[,3]
      distsize[k,1]   = test$Analysis[[t04]]$mu[3]
      distsize[k,2:6] = test$Analysis[[t04]]$CI[,3]

      ## Filter divergence
      divergence[k,1]   = test$Analysis[[tlo]]$mu[3]
      divergence[k,2:6] = test$Analysis[[tlo]]$CI[,3]
    } ## end loop over individual runs

    elist[[exps[i]]] = list(distrate=distrate,predist=predist,
                            distsize=distsize,divergence=divergence)

  } ## end if not processed

} ## end loop over experiments

## grab matching data
obs.pre  = Yw[,t04-1] # 1 x nsite (observation before disturbance)
obs.dist = Yw[,t04] # observation during disturbance
obs.tlo  = Yw[,tlo] # observation during last timestep
obs.class= dextract[,"t2004"]/10 # 1 x nsite
save(elist,obs.pre,obs.dist,obs.tlo,obs.class,file="./experiments/final/elist.RData")

```

## analyses

```{r}
pdf("./experiments/final/exps_diagnostics.pdf")
trim <- function(x,lo=0.0001,hi=0.9999) {
  x[x<lo] <- lo
  x[x>hi] <- hi
  return(x)
}
betatrim <- function(y){
    n.obs <- sum(!is.na(y))
    (y * (n.obs - 1) + 0.5) / n.obs
}
stats <- list()

classnames <- c("other", "cut", "road" ,"fire", "biotic") # disturbance types
class_vals <- sort(unique(obs.class))
class_vals <- class_vals[class_vals != 0] # drop 0 (corresponds to no disturbance)
classes_in_data <- classnames[class_vals]


for(i in seq_along(exps)){
  ## divergence
  div = (elist[[i]]$divergence[,1] - obs.tlo)/
    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5) # 1 x nsite (compute filter divergence)
  div.reg    = lm(div~obs.tlo) # linear model for div in terms of obs.tlo
  div.median = median(div,na.rm = TRUE) # compute median divergence
  div.cov    = sum(div < 1 & div > -1,na.rm=TRUE)/sum(!is.na(div)) # % of sites with filter divergence outside of (-1, 1)

  
  # hist(div,main=exps[i])
  # abline(v=c(-1,1),lty=2)
  plot(obs.tlo,div,col=obs.class,main=exps[i], ylim=c(-2, 2))
  legend("topright",
       legend = classes_in_data,
       col = class_vals,
       pch = 21,
       bty = "n",
       title = "disturbance type")
  abline(h=c(-1,1),lty=2)
  abline(div.reg,col=2)
  
  ## Disturbance detection
  distrate <- elist[[i]]$distrate
  # hist(distrate,main=exps[i])

  ## Disturbance skill
  ord <- order(distrate)
  # plot(distrate[ord],col=obs.class[ord],main=exps[i])
  dist.bar   = mean(distrate,na.rm = TRUE)
  dist.class = tapply(distrate,obs.class,mean,na.rm=TRUE)

  ## vs predisturbance biomass
  distrate2 = betatrim(distrate)
  pred.seq = data.frame(obs.pre=seq(0,max(obs.pre,na.rm = TRUE),length=500),
                        distprop = seq(-1,1,length=500))
  dist.reg.pre  = lm(distrate ~ obs.pre)
  dist.beta.pre = betareg::betareg(distrate2 ~ obs.pre,
                                   na.action=na.omit,link="logit")
  # plot(obs.pre,distrate,col=obs.class,main=exps[i])
  # abline(dist.reg.pre,col=2)
  # lines(pred.seq$obs.pre,predict(dist.beta.pre,newdata=pred.seq),col="green")

  ## vs proportional disturbance size
  distprop = 1 - elist[[i]]$distsize[,1]/elist[[i]]$predist[,1] ## predicted
  distprop = 1 - obs.dist/obs.pre ## observed
  dist.beta.prop = betareg::betareg(distrate2 ~ distprop,
                                   na.action=na.omit,link="logit")
  # plot(distprop,distrate,col=obs.class,main=exps[i])
  # lines(pred.seq$distprop,predict(dist.beta.prop,newdata=pred.seq),col="green")

  ## vs ABSOLUTE disturbance size
  distabs = obs.pre-obs.dist
  pred.seq$distabs = seq(min(distabs, na.rm = TRUE),max(distabs, na.rm = TRUE),length=500)
  dist.beta.abs = betareg::betareg(distrate2 ~ distabs | distabs,
                                   na.action=na.omit,
                                   link="logit",
                                   link.phi = "log",
                                   start=c(0,5,0.5,0),
                                   type="BR")
  # print(distrate)
  # print(distabs)
  dist.qb.abs = glm(distrate2 ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
  dist.bin.abs = glm(round(distrate) ~ distabs,
                    na.action=na.omit,
                    family=binomial("logit")
  )
  plot(distabs,distrate,col=obs.class,main=exps[i])
  legend("topright",
       legend = classes_in_data,
       col = class_vals,
       pch = 21,
       bty = "n",
       title = "disturbance type")
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile"),col="green")
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.75),col="green",lty=2)
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.25),col="green",lty=2)
  lines(pred.seq$distabs,predict(dist.qb.abs,newdata=pred.seq,type="response"),col="purple")
  lines(pred.seq$distabs,predict(dist.bin.abs,newdata=pred.seq,type="response"),col="blue")

  ## vs divergence
  absdiv = abs(div)
  dist.beta.absdiv = betareg::betareg(distrate2 ~ absdiv,
                                   na.action=na.omit,link="logit")
  dist.bin.absdiv = glm(round(distrate) ~ absdiv,
                                   na.action=na.omit,
                        family = binomial("logit"))
  pred.seq$absdiv = seq(0,max(absdiv,na.rm=TRUE),length=500)
  # plot(absdiv,distrate,col=obs.class,main=exps[i])
  # abline(v=c(-1,1),lty=2,col=3)
  # abline(lm(distrate ~ absdiv),col=2)
  # lines(pred.seq$absdiv,predict(dist.beta.absdiv,newdata=pred.seq),col="green")
  # lines(pred.seq$absdiv,predict(dist.bin.absdiv,newdata=pred.seq,type="response"),col="blue")
  # 
  dist.qb.combo = glm(distrate2 ~ obs.pre + distprop +
                        distabs + absdiv + obs.class,
                      na.action=na.omit,
                      family=quasibinomial("logit"))
  dist.bin.combo = glm(round(distrate) ~ obs.pre + distprop +
                        distabs + absdiv + as.factor(obs.class),
                      na.action=na.omit,
                      family=binomial("logit"))

# dist.beta.combo = betareg::betareg(distrate2 ~ obs.pre + obs.class + distabs + absdiv,
#                                   na.action=na.omit,link="logit")

  stats[[i]] <- list(
    div.reg = div.reg,
    div.median = div.median,
    div.cov = div.cov,
    dist.bar = dist.bar,
    dist.class = dist.class,
    dist.pre =  dist.beta.pre,
    dist.prop = dist.beta.prop,
    dist.beta.abs =  dist.beta.abs,
    dist.bin.abs  = dist.bin.abs,
    dist.absdiv   = dist.beta.absdiv,
    dist.combo = dist.bin.combo
  )

}
dev.off()
save(stats,file="./experiments/final/stats.RData")
```

## plot Cw for sites with fires to see disturbance visually

```{r}
nsite <- dim(dextract)[1]
for (i in 1:nsite) {
  site_i <- dextract[i,]
  site_i_fire <- site_i[site_i == 40]
  if (length(site_i_fire) > 0) { # plot site if fire
    plot(Yw[i,], ylim=c(0, 20), type="l", main=paste("site", i))
  }
}
```

```{r}
nsite <- dim(dextract)[1]
for (i in 1:nsite) {
  site_i <- dextract[i,]
  site_i_fire <- site_i[site_i == 20]
  if (length(site_i_fire) > 0) { # plot site if cut
    plot(Yw[i,], ylim=c(0, 20), type="l", main=paste("site", i))
  }
}
```

## determine sites / years with disturbances using disturbance metric

```{r}
pdf("./experiments/exploration/disturbance-years-labelled.pdf")
nsite <- dim(Yw)[1]
nyear <- dim(dextract)[2]
distlist <- list()
disttypes <- seq(0, 50, by=10)

for (i in 1:nsite) {
  pctchanges <- as.numeric(Yw[i, 2:nyear] / Yw[i, 1:nyear-1])
  nchange <- length(pctchanges)
  changediffs <- c()
  for (t in 1:nchange) {
    # find average in 3 year window around t
    tmin <- max(1, t-3)
    tmax <- min(nchange, t+3)
    
    if (t < tmax) {
      changeavg <- mean(c(pctchanges[tmin:(t-1)], pctchanges[(t+1):tmax])) 
    } else {
      changeavg <- mean(pctchanges[tmin:(t-1)])
    }
    changediff <- abs(pctchanges[t] / changeavg)
    changediffs <- c(changediffs, changediff)
  }
  # print(changediffs)
  disidxs <- which(changediffs < 0.80) # get disturbance idxs
  disidxs <- Filter(function(idx) {
    yidx <- Yw[i, idx]
    dyidx <- Yw[i, idx + 1] - Yw[i, idx]
    return(dyidx < -0.75 & yidx > 0.75); # filter indices for which the decrease or starting biomass is too small
  }, disidxs)
  plot(Yw[i,], ylim=c(0, 20), type="l", main=paste("site", i))
  abline(v = disidxs, col = "red", lwd = 2, lty = 2)
  
  
  # for (d in disttypes) {
  #   if (length(which(dextract[i,] == d)) > 0) {
  #     distlist[[as.character(d)]] <- c(distlist[[as.character(d)]], pctchanges[disidxs]) # store disturbances for that site / dist type
  #   }
  # }
}
dev.off()
```

## make density plots for disturbances

```{r}
pdf("./experiments/exploration/disturbance-size-across-years-from-metric.pdf")
dist_name <- c(
  "0"  = "none",
  "10" = "other",
  "20" = "cut",
  "30" = "road",
  "40" = "fire",
  "50" = "biotic"
)
nyear <- dim(dextract)[2]
binwidth <- 0.025

for (disttype in seq(0, 50, by=10)) {
  distpct_all <- as.numeric(distlist[[as.character(disttype)]])
  hist(distpct_all, prob = TRUE, breaks=seq(0, 1.1, by=0.025), xlim=c(0, 1.1), ylim=c(0, 20), xlab="% Cw retained after decrease", main=paste(
    dist_name[as.character(disttype)]
  ))
}
dev.off()
```

## plot disturbances at sites with decreases in biomass (no disturbance product)

```{r}
pdf("./experiments/exploration/disturbance-distributions-decreases.pdf")
lastidx <- dim(Yw)[2]-1 # year 27
for (idx in 2:lastidx) {
  year <- dtime[idx]
  
  obs.pre  = Yw[,idx-1] 
  obs.dist = Yw[,idx]
  distabs = obs.pre - obs.dist
  distpct = 1 - distabs / obs.pre
  # hasdist = distpct < 1 # sites with decrease in biomass
  distpct2 = distpct[distpct >= 0 & distpct < 1.1]
  hist(distpct2, breaks=seq(0, 1.1, by=0.025), xlim=c(0, 1.1), xlab="% Cw retained after decrease", main=paste(year, "biomass % change"))
}
dev.off()
```

## plot disturbance % changes by disturbance type

```{r}
pdf("./experiments/exploration/disturbance-size-distribution-by-type.pdf")
lastidx <- dim(Yw)[2]-1 # year 27
nsite <- dim(Yw)[1]
dist_name <- c(
  "0"  = "none",
  "10" = "other",
  "20" = "cut",
  "30" = "road",
  "40" = "fire",
  "50" = "biotic"
)
for (idx in 2:lastidx) {
  year <- dtime[idx]
  
  obs.pre  = Yw[,idx-1] 
  obs.dist = Yw[,idx]
  disttypes <- unique(dextract[,idx])
  distabs = obs.pre - obs.dist
  distpct = 1 - distabs / obs.pre
  
  for (disttype in disttypes) {
    mask <- dextract[,idx] == disttype
    distpct2 = distpct[distpct >= 0 & distpct < 1.1 & mask]
    hist(distpct2, breaks=seq(0, 1.1, by=0.025), xlim=c(0, 1.1), ylim=c(0, 120), xlab="% Cw retained after decrease", main=paste(year, "biomass % change", "type =", dist_name[as.character(disttype)]))
  }
}
dev.off()
```

## plot % change across all years

```{r}
pdf("./experiments/exploration/disturbance-size-across-years.pdf")

nyear <- dim(dextract)[2]
binwidth <- 0.025

for (disttype in seq(0, 50, by=10)) {
  distpct_all <- c()
  for (y in 2:nyear) {
    obs.pre  = Yw[,y-1] 
    obs.dist = Yw[,y]
    distabs = obs.pre - obs.dist
    distpct = 1 - distabs / obs.pre
    mask <- dextract[,y] == disttype
    distpct2 <- distpct[distpct >= 0 & distpct < 1.1 & mask]
    distpct_all <- c(distpct_all, distpct2)
  }
  # ndata <- length(distpct_all)
  # distpct_all_freqs <- distpct_all / (ndata * binwidth) # frequency plot
  hist(distpct_all, prob = TRUE, breaks=seq(0, 1.1, by=0.025), xlim=c(0, 1.1), ylim=c(0, 20), xlab="% Cw retained after decrease", main=paste(
    dist_name[as.character(disttype)]
  ))
}

```

## plot percent change across all years, count both first / second year as disturbance year

```{r}
pdf("./experiments/exploration/disturbance-size-across-years-count-first-and-second.pdf")

nyear <- dim(dextract)[2]
binwidth <- 0.025

for (disttype in seq(0, 50, by=10)) {
  distpct_all <- c()
  for (y in 2:nyear) {
    obs.pre  = Yw[,y-1] 
    obs.dist = Yw[,y]
    distabs = obs.pre - obs.dist
    distpct = 1 - distabs / obs.pre
    mask <- dextract[,y] == disttype | dextract[,y-1] == disttype # counting both first and second
    distpct2 <- distpct[distpct >= 0 & distpct < 1.1 & mask]
    distpct_all <- c(distpct_all, distpct2)
  }
  # ndata <- length(distpct_all)
  # distpct_all_freqs <- distpct_all / (ndata * binwidth) # frequency plot
  hist(distpct_all, prob = TRUE, breaks=seq(0, 1.1, by=0.025), xlim=c(0, 1.1), ylim=c(0, 20), xlab="% Cw retained after decrease", main=paste(
    dist_name[as.character(disttype)]
  ))
}
```

## plot percent change across all years for all first years that are 3 years from measured disturbance

```{r}
pdf("./experiments/exploration/disturbance-size-across-years-3-year-window.pdf")

nsite <- dim(dextract)[1]
nyear <- dim(dextract)[2]
binwidth <- 0.025

for (disttype in seq(0, 50, by=10)) {
  
  distpct_all <- c()
  for (i in 1:nsite) {
    # get window of starting years
    distindices <- which(dextract[i,] == disttype)
    if (length(distindices) > 0) {
      min_start <- max(1, min(distindices) - 5) # min starting year / max starting year
      max_start <- min(nyear - 1, max(distindices) + 5)
      
      for (y in min_start:max_start) {
       
        pctchange <- as.numeric(Yw[i, y+1] / Yw[i, y])
        if (!is.nan(pctchange) & !is.null(pctchange)) {
          if (0 <= pctchange & pctchange <= 1.1) {
            distpct_all <- c(distpct_all, pctchange) 
          } 
        }
      } 
    }
  }
  
  # ndata <- length(distpct_all)
  # distpct_all_freqs <- distpct_all / (ndata * binwidth) # frequency plot
  print(distpct_all)
  if (length(distpct_all) > 0) {
    hist(distpct_all, prob = TRUE, breaks=seq(0, 1.1, by=0.025), xlim=c(0, 1.1), ylim=c(0, 20), xlab="% Cw retained after decrease", main=paste(
      dist_name[as.character(disttype)]
    )) 
  }
}
```

## explore distribution of % biomass lost in disturbance (in years marked by disturbance product)

```{r}
# plot histogram of number of disturbed sites per year

num.dist <- apply(dextract, 2, function(col.dists) {
  return(length(col.dists[col.dists > 0]))
})
dtime <- 1985:2012
t2004 <- 20
plot(dtime, num.dist, main="number of disturbance events each year")
```

## plot sites with high distabs, but low distrate

```{r}
load("./experiments/final/elist.RData")
load("./experiments/final/stats.RData")

# distabs > 10 but distrate < 0.1
distabs <- obs.pre-obs.dist
distrate <- elist[[1]]$distrate # Rgedi_Htrue_Dtrue
mask <- (distabs > 3) & (distrate < 0.4)
highabslowdist <- which(mask)
print(highabslowdist)

# plot analysis for these sites
pool.names=c("Cleaf","Csoil","Cstem")
runs <- dir(path="./experiments/final/", pattern="Rgedi_Htrue_Dtrue")



for (site in highabslowdist) {
  runfile <- paste0("Rgedi_Htrue_Dtrue_OR_SD_", sprintf("%03d", site), ".RData")
  load(paste0("./experiments/final/", runfile))
  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
      run.muN = sapply(X = test$Analysis, function(x) {
        y = x$m_starN
        y[i]
      })
      run.muD = sapply(X = test$Analysis, function(x) {
        y = x$m_starD
        y[i]
      })
      run.mufN = sapply(X = test$Forecast, function(x) {
        y = x$mufN
        y[i]
      })
      run.mufD = sapply(X = test$Forecast, function(x) {
        y = x$mufD
        y[i]
      })

      run.probD = sapply(X = test$Analysis, function(x) {
        y = x$muD
      })
      
      CI.NfD = sapply(X = test$Forecast, function(x) {
        y = x$CID
        y[,i]
      })
      
      CI.NfN = sapply(X = test$Forecast, function(x) {
        y = x$CIN
        y[,i]
      })


      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=pool.names[i],xlab="time", main=runfile)

    ecoforecastR::ciEnvelope(1:ncol(CI.Na),
                             CI.Na[1,], CI.Na[5,],
                             col=ecoforecastR::col.alpha('#b2df8a',0.7))

    # Median CI.Na line
    lines(CI.Na[3,], col="#b2df8a", lwd=2)

    # μN line
    lines(unlist(run.muN), col="#40826D", lwd=2)

    # μD line
    lines(unlist(run.muD), col="#9D2933", lwd=2)
    
    # forecast mean N
    lines(unlist(run.mufN), col="#002EFF", lwd=2)
    
    # forecast mean D
    lines(unlist(run.mufD), col="#FF002E", lwd=2)
    
    # forecast CID
    ecoforecastR::ciEnvelope(1:ncol(CI.NfD),
                             CI.NfD[1,], CI.NfD[5,],
                             col=ecoforecastR::col.alpha('#e86666',0.7))
    
    # forecast CIN
    ecoforecastR::ciEnvelope(1:ncol(CI.NfN),
                             CI.NfN[1,], CI.NfN[5,],
                             col=ecoforecastR::col.alpha('#80bcf3',0.7))

    # Observations
    points(test$Y[,i], pch = 1)

    # === ADD LEGEND HERE ===
    legend("topright",
           legend = c("CI.mean", "muN", "muD", "mufN", "mufD", "observed"),
           col    = c("#b2df8a", "#40826D", "#9D2933",  "#002EFF", "#FF002E", "black"),
           lwd    = c(2, 2, 2, 2, 2, NA),
           pch    = c(NA, NA, NA, NA, NA, 1),
           bty    = "n")
    # ========================
    plot(run.probD, main=paste0(runfile, " disturbance probs"))
    lines(run.probD, col="#9D2933", lwd=2)
    
    plot(unlist(test$ProcessErrors), main=paste0(runfile, " process errors"))
    lines(unlist(test$ProcessErrors), col="#FF005F")
    
    # lines(CI.D[3,])
  }
}
```

## Compare Analytical (UpdateQ = False) / MCMC Results

```{r}
## summary figure showing didn't work great at default error rate
m = matrix(1:6,nrow=3,byrow = TRUE)
layout(m)
options(repr.plot.width = 14, repr.plot.height = 14)

### Row 1: Example time-series ###
site = 185 # 188
exp.name = "P1Q1R0.001"
analytical_elist_idx = 1 # indices of these experiments in elist
mcmc_elist_idx = 5
exp.pretty.name = c("(a) Example site, ATrueQFalse, R=0.001","(c) Example site, MCMC (R = 0.001)")
pool.names=c("Leaf","Soil","AGB")
i = 3 ## Stem
time = 1985:2017

load(paste0("ATrueQFalse_OR_SD_",site,"_",exp.name,".RData"))
CI.Na = sapply(X = test$Analysis,function(x){
      y = x$CI
      y[,i]
    })
plot(time,test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main=exp.pretty.name[1])
ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('blue',0.5))
lines(time,CI.Na[3,],col="blue",lwd=2)
points(time,test$Y[,i],col="red",cex=1.5)
legend("bottomleft",legend=c("Assimilation","Data"),
     lwd=5,col=c("blue","red"),cex=0.7)


load(paste0("OR_SD_",site,"_",exp.name,".RData"))
print(paste0("OR_SD_",site,"_",exp.name,".RData"))
CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        sel.n = which(colnames(y) == paste0("n[",i,"]"))
        y[,sel.n]
      })
plot(time,test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main=exp.pretty.name[2])
ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('blue',0.5))
lines(time,CI.Na[3,],col="blue",lwd=2)
points(time,test$Y[,i],col="red",cex=1.5)

### Row 2: Divergence ###
load("ATrueQFalseStats.RData")
load("ATrueQFalseElist.RData")
disturb.names = c("Intact","Other","Cut","Fire","Pest")
disturb.col = c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a')
disturb.col.al <- tapply(disturb.col,1:5,FUN=ecoforecastR::col.alpha,0.7); disturb.col.al[1] = disturb.col[1]
div.pretty.name = c("(b) Divergence, Analytical","(e) Divergence, MCMC")
div.reg <- list()
div.class <- list()
obs.class.re = obs.class; obs.class.re[obs.class.re ==4] = 3; obs.class.re[obs.class.re ==5] = 4; obs.class.re = obs.class.re + 1

i <- analytical_elist_idx
div = (elist[[i]]$divergence[,1] - obs.tlo)/
  (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
div.reg[[i]]    = lm(div~obs.tlo)
div.class[[i]]    = lm(div~obs.tlo + as.factor(obs.class))

dh = density(na.omit(div))
#  abline(v=c(-1,1),lty=2)
plot(obs.tlo,div,col=disturb.col.al[obs.class.re],
     xlab="final AGB",ylab="Divergence",
     pch=16, main=div.pretty.name[1])
abline(h=c(-1,1),lty=2)
abline(div.reg[[i]],col="purple",lwd=3)
abline(v=0,lty=2)
lines(dh$y*10,dh$x,lwd=3)
legend("topright",legend=disturb.names,
                col=disturb.col,pch=16,cex=0.7,
                text.col=c(1,disturb.col[2:5]))

load("stats.RData")
load("elist.RData")

i <- mcmc_elist_idx
div = (elist[[i]]$divergence[,1] - obs.tlo)/
  (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
div.reg[[i]]    = lm(div~obs.tlo)
div.class[[i]]    = lm(div~obs.tlo + as.factor(obs.class))

dh = density(na.omit(div))
#  abline(v=c(-1,1),lty=2)
plot(obs.tlo,div,col=disturb.col.al[obs.class.re],
     xlab="final AGB",ylab="Divergence",
     pch=16, main=div.pretty.name[2])
abline(h=c(-1,1),lty=2)
abline(div.reg[[i]],col="purple",lwd=3)
abline(v=0,lty=2)
lines(dh$y*10,dh$x,lwd=3)


### Row 3: Disturbance magnitude vs P(detection) ###
load("ATrueQFalseStats.RData")
load("ATrueQFalseElist.RData")
pred.seq=list()
distabs = obs.pre-obs.dist
pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
dist.qb.abs <- list()
i <- analytical_elist_idx
distrate = distrate = elist[[i]]$distrate
dist.qb.abs[[i]] = glm(distrate ~ distabs,
                  na.action=na.omit,
                  family=quasibinomial("logit")
)
plot(distabs,distrate,col=disturb.col.al[obs.class.re],
     xlab="AGB disturbance magnitude",ylab="P(detect)",
     pch=16, main="(c) Probability of detection, Analytical",xlim=c(0,17), ylim=c(0, 1))
lines(pred.seq$distabs,predict(dist.qb.abs[[i]],newdata=pred.seq,type="response"),col="purple")

load("stats.RData")
load("elist.RData")
pred.seq=list()
distabs = obs.pre-obs.dist
pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
dist.qb.abs <- list()
i <- mcmc_elist_idx
distrate = distrate = elist[[i]]$distrate
dist.qb.abs[[i]] = glm(distrate ~ distabs,
                  na.action=na.omit,
                  family=quasibinomial("logit")
)
plot(distabs,distrate,col=disturb.col.al[obs.class.re],
     xlab="AGB disturbance magnitude",ylab="P(detect)",
     pch=16, main="(f) Probability of Detection, MCMC",xlim=c(0,17), ylim=c(0, 1))
lines(pred.seq$distabs,predict(dist.qb.abs[[i]],newdata=pred.seq,type="response"),col="purple")
if(i == 1) legend("topleft",legend=disturb.names,
                col=disturb.col,pch=16,cex=0.7,
                text.col=c(1,disturb.col[2:5]))
```

## Compare Analytical (UpdateQ = True) / MCMC Results

```{r}
## summary figure showing didn't work great at default error rate
m = matrix(1:6,nrow=3,byrow = TRUE)
layout(m)
options(repr.plot.width = 14, repr.plot.height = 14)

### Row 1: Example time-series ###
site = 185 # 188
exp.name = "P1Q1R0.001"
analytical_elist_idx = 1 # indices of these experiments in elist
mcmc_elist_idx = 5
exp.pretty.name = c("(a) Example site, Analytical / Updating Q, R=0.001","(c) Example site, MCMC (R = 0.001)")
pool.names=c("Leaf","Soil","AGB")
i = 3 ## Stem
time = 1985:2017

load(paste0("ATrueQTrue_OR_SD_",site,"_",exp.name,".RData"))
print(paste0("ATrueQTrue_OR_SD_",site,"_",exp.name,".RData"))
CI.Na = sapply(X = test$Analysis,function(x){
      y = x$CI
      y[,i]
    })
plot(time,test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main=exp.pretty.name[1])
ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('blue',0.5))
lines(time,CI.Na[3,],col="blue",lwd=2)
points(time,test$Y[,i],col="red",cex=1.5)
legend("bottomleft",legend=c("Assimilation","Data"),
     lwd=5,col=c("blue","red"),cex=0.7)


load(paste0("OR_SD_",site,"_",exp.name,".RData"))
print(paste0("OR_SD_",site,"_",exp.name,".RData"))
CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        sel.n = which(colnames(y) == paste0("n[",i,"]"))
        y[,sel.n]
      })
plot(time,test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main=exp.pretty.name[2])
ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('blue',0.5))
lines(time,CI.Na[3,],col="blue",lwd=2)
points(time,test$Y[,i],col="red",cex=1.5)

### Row 2: Divergence ###
load("AnalyticalStats.RData")
load("AnalyticalElist.RData")
disturb.names = c("Intact","Other","Cut","Fire","Pest")
disturb.col = c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a')
disturb.col.al <- tapply(disturb.col,1:5,FUN=ecoforecastR::col.alpha,0.7); disturb.col.al[1] = disturb.col[1]
div.pretty.name = c("(b) Divergence, Analytical","(e) Divergence, MCMC")
div.reg <- list()
div.class <- list()
obs.class.re = obs.class; obs.class.re[obs.class.re ==4] = 3; obs.class.re[obs.class.re ==5] = 4; obs.class.re = obs.class.re + 1

i <- analytical_elist_idx
div = (elist[[i]]$divergence[,1] - obs.tlo)/
  (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
div.reg[[i]]    = lm(div~obs.tlo)
div.class[[i]]    = lm(div~obs.tlo + as.factor(obs.class))

dh = density(na.omit(div))
#  abline(v=c(-1,1),lty=2)
plot(obs.tlo,div,col=disturb.col.al[obs.class.re],
     xlab="final AGB",ylab="Divergence",
     pch=16, main=div.pretty.name[1], ylim=c(-1.1, 1.1))
abline(h=c(-1,1),lty=2)
abline(div.reg[[i]],col="purple",lwd=3)
abline(v=0,lty=2)
# lines(dh$y*10,dh$x,lwd=3)
legend("topright",legend=disturb.names,
                col=disturb.col,pch=16,cex=0.7,
                text.col=c(1,disturb.col[2:5]))

load("stats.RData")
load("elist.RData")

i <- mcmc_elist_idx
div = (elist[[i]]$divergence[,1] - obs.tlo)/
  (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
div.reg[[i]]    = lm(div~obs.tlo)
div.class[[i]]    = lm(div~obs.tlo + as.factor(obs.class))

dh = density(na.omit(div))
#  abline(v=c(-1,1),lty=2)
plot(obs.tlo,div,col=disturb.col.al[obs.class.re],
     xlab="final AGB",ylab="Divergence",
     pch=16, main=div.pretty.name[2])
abline(h=c(-1,1),lty=2)
abline(div.reg[[i]],col="purple",lwd=3)
abline(v=0,lty=2)
lines(dh$y*10,dh$x,lwd=3)


### Row 3: Disturbance magnitude vs P(detection) ###
load("AnalyticalStats.RData")
load("AnalyticalElist.RData")
pred.seq=list()
distabs = obs.pre-obs.dist
pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
dist.qb.abs <- list()
i <- analytical_elist_idx
distrate = distrate = elist[[i]]$distrate
dist.qb.abs[[i]] = glm(distrate ~ distabs,
                  na.action=na.omit,
                  family=quasibinomial("logit")
)
plot(distabs,distrate,col=disturb.col.al[obs.class.re],
     xlab="AGB disturbance magnitude",ylab="P(detect)",
     pch=16, main="(c) Probability of detection, Analytical",xlim=c(0,17), ylim=c(0, 1))
lines(pred.seq$distabs,predict(dist.qb.abs[[i]],newdata=pred.seq,type="response"),col="purple")

load("stats.RData")
load("elist.RData")
pred.seq=list()
distabs = obs.pre-obs.dist
pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
dist.qb.abs <- list()
i <- mcmc_elist_idx
distrate = distrate = elist[[i]]$distrate
dist.qb.abs[[i]] = glm(distrate ~ distabs,
                  na.action=na.omit,
                  family=quasibinomial("logit")
)
plot(distabs,distrate,col=disturb.col.al[obs.class.re],
     xlab="AGB disturbance magnitude",ylab="P(detect)",
     pch=16, main="(f) Probability of Detection, MCMC",xlim=c(0,17), ylim=c(0, 1))
lines(pred.seq$distabs,predict(dist.qb.abs[[i]],newdata=pred.seq,type="response"),col="purple")
if(i == 1) legend("topleft",legend=disturb.names,
                col=disturb.col,pch=16,cex=0.7,
                text.col=c(1,disturb.col[2:5]))
```

## quick check: find and plot sites with high distabs, but distrate = 0

```{r}
print(exps)

# stats[[1]] = P1Q1R0.001
distabs <- obs.pre-obs.dist
distrate <- elist[[1]]$distrate

# get sites with distabs > 5 but distrate = 0
allsites <- 1:356
distratezero <- distrate == 0
distratezero[is.na(distratezero)] = FALSE
print(length(distabs))
print(length(distratezero))
print(distabs & distratezero)
testsites <- allsites[distabs > 5 & distratezero]

# plot DA for those sites
for (j in testsites) {
  pool.names=c("Cleaf","Csoil","Cstem")
# runs <- dir(pattern="*.RData")
# runs <- runs[grepl("^Analytical_OR",runs)]
# print(runs)
  runfile <- paste0("Analytical_OR_SD_", j, "_P1Q1R0.001.RData")
  load(runfile)
# runs <- runs[grepl("84",runs)]

  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
  
      
      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
         ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
         type='n',
         ylab=pool.names[i],xlab="time", main=runfile)
    # ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
    ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
    # ecoforecastR::ciEnvelope(1:ncol(CI.Na.2),CI.Na.2[1,],CI.Na.2[3,],col=ecoforecastR::col.alpha("red",0.7))
    lines(CI.Na[3,],col="#b2df8a",lwd=2)
    points(test$Y[,i])
    # lines(CI.D[3,])
}
}
```

factors affecting detection of disturbance

```{r}
Ps = as.numeric(substr(names(elist),2,2))
Qs = as.numeric(substr(names(elist),4,4))
Rs = as.numeric(substr(names(elist),6,6))

## P

## Q

## R

## Divergence
div.cov = sapply(stats,"[[","div.cov")
plot(Rmult[Rs],div.cov,col=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage")
legend("topright",legend=paste0(Qmult[1:2],"x"))

div.median = sapply(stats,"[[","div.median")
plot(Rmult[Rs],div.median,col=Qs,log="x",xlab="Variance Multiplier",ylab="Median Divergance")
legend("topleft",legend=paste0(Qmult[1:2],"x"))

div.slope = sapply(stats,function(x){coef(x$div.reg)[2]})
plot(Rmult[Rs],div.slope,col=Qs,log="x",xlab="Variance Multiplier",ylab="Divergance Slope")
## Disturbance @ default
summary(stats[[4]]$dist.combo)
## nothing matters

## Disturbance @ P1Q2R2 (intermediate)
summary(stats[[4]]$dist.combo)
## detection varies with divergence & disturbance type

## Disturbance magnitude: P1Q1R4
i = 3
summary(stats[[i]]$dist.combo)  ## absolute magnitude of divergence is best predictor
summary(stats[[i]]$dist.bin.abs)
beta = coef(stats[[i]]$dist.bin.abs)
## threshold mx+b = 0
thresh = -beta[1]/beta[2] 

## Disturbance type: P1Q1R4 (reduced divergence)
stats[[i]]$dist.class
tapply(distabs,obs.class,summary) ## obs dist size by class
thresh.prop = tapply(distabs,obs.class,function(x){sum(x>thresh,na.rm = TRUE)/sum(!is.na(x))}) ## prop > detection threshold
thresh.prop
stats[[i]]$dist.class
dclass = stats[[i]]$dist.class
#dclass[1] = 0
dcol = as.numeric(names(stats[[i]]$dist.class)) + 1
plot(thresh.prop,dclass,col=dcol,pch=19)
abline(0,1,lty=2,col=2)
legend("topleft",legend = dcol-1,col=dcol,pch=19)


## additional experiments to run (P1Q1R4)
```

false positives

false negatives

classification success

## preprocess timeseries example for visualization

```{r}
## visualization to help select a site to look at
plotly::plot_ly(data.frame(obs.pre=obs.pre,obs.post=obs.dist),x=~obs.pre,y=~obs.post)
which.min((obs.pre-13.3)^2+(obs.dist-5.1)^2) ## select a point

## settings
site = 188
exp.name = "P1Q1R4" #"P1Q1R1"  ## manually run for each experiment **

## data
NT = 33
Yi = matrix(NA,NT,3)  ## observed data
Yi[1:ncol(Yw),3] = Yw[site,]  ## only Cw observed
  
## model
load(paste0("OR_SD_",site,"_",exp.name,".RData"),envir = viz)
pool.names=c("Cleaf","Csoil","Cstem")
test = viz$test

for(i in 1:3){
  
    CI.Na = sapply(X = test$Analysis,function(x){
      y = x$CI
      sel.n = which(colnames(y) == paste0("n[",i,"]"))
      y[,sel.n]
    })
    
    CI.D = sapply(X = test$Analysis,function(x){
      y = x$CI
      sel.n = which(colnames(y) == paste0("D"))
      y[,sel.n]
    })

    CI.Nf = sapply(X = test$Forecast, function(x){
      mu = x$mufN[i]
      sd = 1/sqrt(x$pfN[i,i])
      return(mu + c(-1.96,0,1.96)*sd)
    })
    
  plot(Yi[,i],
       ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
       type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[3,],col="#b2df8a",lwd=2)
  points(Yi[,i])
  lines(CI.D[3,])
  
  
}
save(Yi,CI.Na,CI.D,file=paste0("OR_TS_",site,"_",exp.name,".RData"))
```

## summary info for filter divergence figures

```{r}
exp.name = "P1Q1R4" #"P1Q1R1"

for(i in seq_along(exps)){
  
  ## divergence
  div = (elist[[i]]$divergence[,1] - obs.tlo)/
    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
  div.reg    = lm(div~obs.tlo)
  div.median = median(div,na.rm = TRUE)
  div.cov    = sum(div < 1 & div > -1,na.rm=TRUE)/sum(!is.na(div))
  
}
```

## Next steps:

-   Run with multiple disturbance options and PFT switching





