---
title: "OR Multi Disturbance"
author: "Michael Dietze"
date: "12/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("disturbance.R")
source("multi.ens.adj.R")
source("multi.analytical.ens.adj.R")
source("StatsAuxiliary.R")
library(BayesianTools)
library(magrittr)
```

## set up
```{r}
## Load disturbance rate, state&transition, disturbed biomass stats
load("STprior.RData")
load("fit.RData")
load("extract.h03v04.RData")
# print(dtime)
# print(paste0("t", dtime))
# print(colnames(dextract))
dtime <- 1985:2012
colnames(dextract) <- paste0("t",dtime)
dextract[is.na(dextract)] <- 0
d.prior["UN"] = d.prior["UN"] + d.prior[6]                         ## change at runtime (removes <NA>)
d.prior = d.prior[-6]
# Question: what are the classes of disturbance in d.prior? 
print(dim(dextract))
print(colnames(dextract))
## Load Calibration
load("fit.RData")
Qsoil = Qs
```

## drivers
```{r}
latlon = points@coords # lat / lon of each site in dextract 
metFile = "daymet.RData"
if(file.exists(metFile)){
  load(metFile)
} else {
  dm = list()
  for(i in 1:nrow(latlon)){
    dm[[i]] = daymetr::download_daymet(site="h03v04",latlon[i,2],latlon[i,1],start=1985,end=2017,internal=TRUE,simplify=TRUE)
  }
  save(dm,file=metFile)
}

## extract Shortwave and convert to daily PAR
SW2PAR = 0.45
srad = subset(dm[[1]],measurement == "srad..W.m.2.")   
metTime <- as.POSIXct(paste(srad$year,srad$yday,sep="-"),format="%Y-%j")
rad <- matrix(NA,length(dm),length(metTime)) # [number of sites x number of measurements]
print(dim(rad))
for(i in seq_along(dm)){
  srad = subset(dm[[i]],measurement == "srad..W.m.2.")   
  rad[i,] = srad$value*1e-6*86400*SW2PAR  # -> MJ /m2 /day PAR (convert) # QUESTION -- should the changing day length, from the daymet day length values, be used (instead of a fixed 24-hr day length)?
}
plot(metTime,rad[1,],type='l')
## plot(date,PAR,col=2) ## calibration data
```

## functions
```{r,echo=FALSE,results='hide',message=FALS}
##' Single Site, Single Disturbance
##' @param Y observed data
##' @param R observation error
##' @param Q process error
##' @param Xic state initial condition matrix
##' @param THETA parameter matrix
##' @param dist  disturbance list: p.dist, mu0[3], V0[3,3], alloc soil
##' @param updateQ if true, conditionally update the process error Q based on observations
##' @param useAdjustment if true, use ensemble adjustment. If false, just sample from the posterior.
##' @param use_analytical_posterior if true use analytical posterior, if false use MCMC to sample from the posterior
SSSD.model <- function(Y,R,Q,PAR,Xic,THETA,dist,updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE){
  
  ## init
  NT = length(PAR)/365
  Nf = Nmu = Na = Xic
  ne = nrow(Xic)
  if(!is.matrix(THETA)) THETA = matrix(THETA,nrow=ne,ncol=length(THETA),byrow = TRUE)
  Forecast = Analysis = list()
  ProcessErrors = list()
  
  AnalysisSSSDMP <- "  ## SingleSite SingleDisturbance MultiPool
    model{
      ## prior
      muN ~ dmnorm(mufN,pfN)                ## forecast prior on Not disturbed
      muD ~ dmnorm(mufD,pfD)                ## forecast prior on Disturbed
      D ~ dbern(p)                          ## disturbance probability
      ## Analysis 
      n <- D*muD + (1-D)*muN                ## select between disturbed and undisturbed
      Y[H] ~ dmnorm(n[H],R[H,H])                       ## Observation error
  }"
  
  a_t = 1 # place a prior on process precision: (1/Q) ~ Gamma(a_0, b_0).
  b_t = 1
  
  for(t in 1:NT){
    
    ## update IC
    THETA[,9:11]  <- Na    ## Multinomial
    
    ## Forecast 
    for(i in 1:ne){
      tsel    = (t-1)*365 + 1:365             ## Select days from driver files
      Nmu[i,] = VSEM(THETA[i,], PAR = PAR[tsel])[365,2:4] ## deterministic simulation
    }
    
    ## update process error (if we have observed data)
    ## assume Y ~ N(mu, Q). We will pick Q using the posterior: Q = 1/(E[p((1/Q) | y)]) = 1/E[Gamma(a_t, b_t)] = b_t / a_t
    if (!is.na(Y[t,3]) && updateQ) {
      a_t <- a_t + 0.5
      b_t <- b_t + 0.5 * (Y[t,3] - colMeans(Nmu)[3])^2
      Q[3, 3] = b_t / a_t
      ProcessErrors[[t]] <- Q[3, 3]
    }
    
    ## Add process error
    for(i in 1:ne) {
      Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
    }
    
    ## Disturbance
    d.forecast <- rep(0,ne)
    d.forecast[1:ceiling(ne*0.1)] <- 1 ## prior on disturbance status (set ~10% of ensemble members to be disturbed) Question: Why not use dist$p.dist? Why not disturb random ensemble members
    # d.forecast <- rbinom(ne, 1, dist$p.dist)
    sel <- !as.logical(d.forecast)                ## select undisturbed ensemble members
    if(sum(!sel)>2){ # if there are more than 2 disturbed ensemble members
      sel.dist = which(!sel)
      for(j in sel.dist){                      ## run disturbance model
        Nf[j,]  = disturbance.p(Nf[j,],mu0=dist$mu0,V0 = dist$V0,alloc.soil = dist$alloc.soil) # disturb states of the selected ensemble members
      }
      mufD  = colMeans(Nf[!sel,]) # compute sample mean of disturbed ensemble members
      CfD = var(Nf[!sel,])
      pfD = try(solve(CfD)) # compute sample precision of disturbed ensemble members
    } 
    if(sum(!sel) <= 2 || "try-error" %in% class(pfD)) {   ## backup values if the ensemble stats fail
      mufD = c(dist$mu0[1], mean(Nf[,2]),  dist$mu0[2]) # use default disturbance mean            
      CfD = diag(c(0.15,0.5,0.25)^2)
      pfD  = solve(CfD) # use hard-coded precision
    }
    Nf <- tobit(Nf)
    
    ## estimate forecast parameters
    pD = 0.1 # pD = as.numeric(dist$p.dist) ## disturbance rate
    mufN = colMeans(Nf[sel,])    ## Not disturbed forecast means
    CfN = var(Nf[sel,])
    pfN = solve(CfN)  ## Not disturbed forecast precision
    H = which(!is.na(Y[t,]))
    if (length(H) == 1) { ## data observed, run filter
      run.adjustment = TRUE
      if (use_analytical_posterior) { ## use analytical posterior
        ## compute posterior parameters
        H = as.numeric(H[1])
        mat_H = matrix(0, 1, 3)
        mat_H[H] <- 1
        R_H = R[H, H]
        C_starD = solve(pfD + t(mat_H) %*% ((1/R_H) * mat_H))
        m_starD = C_starD %*% (pfD %*% mufD + (t(mat_H) * (1/R_H) * Y[t,H]))
        C_starN = solve(pfN + t(mat_H) %*% ((1/R_H) * mat_H))
        m_starN = C_starN %*% (pfN %*% mufN + (t(mat_H) * (1/R_H) * Y[t,H]))
        
        # posterior mixture weights
        LD_star = dnorm(Y[t,H], mean=mat_H %*% mufD, sd = sqrt(mat_H %*% CfD %*% t(mat_H) + R_H)) # posterior probability of disturbance
        LN_star = dnorm(Y[t,H], mean=mat_H %*% mufN, sd = sqrt(mat_H %*% CfN %*% t(mat_H) + R_H)) # posterior probability of not being disturbed
        total = pD * LD_star + (1-pD) * LN_star
        if (total > 0) {
          pD_star = (pD * LD_star) / total
          pN_star = ((1-pD) * LN_star) / total
        } else {
          pD_star = 0.5
          pN_star = 0.5 ## set them to be equal
        }
        # test: compute MCMC posterior approximation
        priors <- list(p    = pD,         ## disturbance rate
                     mufN = mufN,    ## Not disturbed forecast means
                     mufD = mufD,   ## Disturbed forecast means
                     pfN  = pfN,  ## Not disturbed forecast precision
                     pfD  = pfD  ## Disturbed forecast precision
        )
        # update = c(list(Y=Y[t,],R=solve(R)),priors,H = which(!is.na(Y[t,])))
        # mod <- rjags::jags.model(file=textConnection(AnalysisSSSDMP),
        #                          data=update,
        #                          n.adapt=1000,n.chains=3,quiet=TRUE)
        # jdat <- rjags::coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000)
        # dat = as.matrix(jdat)
        # nsel <- grep(pattern="^n",x=colnames(dat))
        # Xa = dat[,nsel] # analysis samples
        # ca = dat[,"D"]
        # muXa = colMeans(Xa)
        # print("Computed mean from MCMC posterior")
        # print(muXa)
        # print("computed cov from MCMC posterior")
        # print(cov(Xa))
        # print("data Y[t]")
        # print(Y[t,])
        # print("computed mean from analytical posterior")
        # print(pN_star * m_starN + pD_star * m_starD)
        # print("computed disturbed cov from MCMC")
        # print(cov(Xa[ca == 1,]))
        # print("computed undisturbed cov from MCMC")
        # print(cov(Xa[ca == 0,]))
        # print("computed disturbed cov from analytical posterior")
        # print(C_starD)
        # print("computed undisturbed cov from analytical posterior")
        # print(C_starN)
        # print("computed disturbed mean from MCMC")
        # print(colMeans(Xa[ca==1,]))
        # print("computed undisturbed mean from MCMC")
        # print(colMeans(Xa[ca==0,]))
        # print("computed disturbed mean from analytical posterior")
        # print(m_starD)
        # print("computed undisturbed mean from analytical posterior")
        # print(m_starN)
        # print("")
        # print(paste0("year = ", t))
        # print("computed prob dist from MCMC")
        # print(sum(ca) / length(ca))
        # print("computed prob dist from analytical posterior")
        # print(pD_star)
        # print("posterior mixture weights unnormalized")
        # print(paste0("LD_star: ", LD_star))
        # print(paste0("LN_star: ", LN_star))
        
      } else { ## run MCMC
        # test: compute MCMC posterior approximation
        priors <- list(p    = pD,         ## disturbance rate
                     mufN = mufN,    ## Not disturbed forecast means
                     mufD = mufD,   ## Disturbed forecast means
                     pfN  = pfN,  ## Not disturbed forecast precision
                     pfD  = pfD  ## Disturbed forecast precision
        )
        update = c(list(Y=Y[t,],R=solve(R)),priors,H = which(!is.na(Y[t,])))
        mod <- rjags::jags.model(file=textConnection(AnalysisSSSDMP),
                                 data=update,
                                 n.adapt=1000,n.chains=3,quiet=TRUE)
        jdat <- rjags::coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000)
        dat = as.matrix(jdat)
        nsel <- grep(pattern="^n",x=colnames(dat))
        Xa = dat[,nsel] # analysis samples
        ca = dat[,"D"]
        # print("unique dist classes")
        # print(unique(ca))
        # muXa = colMeans(Xa)
        # print("computed disturbed mean from MCMC")
        # print(colMeans(Xa[ca==1,]))
        # print("computed undisturbed mean from MCMC")
        # print(colMeans(Xa[ca==0,]))
        # print("computed disturbed mean from MCMC (directly from muD samples)")
        # XDa = dat[,grep(pattern="^muD", x=colnames(dat))]
        # print(colMeans(XDa))
        # print("computed 50% quantile of muD samples")
        # print(apply(XDa,2,quantile,c(0.5)))
        # print("computed undisturbed mean from MCMC (directly from muN samples)")
        # XNa = dat[,grep(pattern="^muN", x=colnames(dat))]
        # print(colMeans(XNa))
        # print("computed 50% quantile of muN samples")
        # print(apply(XNa,2,quantile,c(0.5)))
        # print("mufD:")
        # print(mufD)
        # print("mufN")
        # print(mufN)
      }
    } else { # if we don't have observed data
      Na = Nf # analysis = forecast
      ca = 1
      run.adjustment = FALSE
    }
      ## draw new IC
    if(run.adjustment){
      if(useAdjustment && use_analytical_posterior) { 
        ## run ensemble adjustment
        Na <- multi.analytical.ens.adj(
            Xf   = Nf,
            cf   = d.forecast,
            mu.f = rbind(mufN,mufD),
            Pf   = abind::abind(CfN,
                   CfD,
                   along=3),  # covariance matrices for undisturbed / disturbed ensemble states
            Pp.mu.a = cbind(m_starN, m_starD),
            Pp.Pa = abind::abind(C_starN, C_starD, along=3),
            Pp.w = c(pN_star, pD_star),
            uc = c(0, 1))
        Ca <- Na[,1] ## updated class assignments
        Na <- tobit(Na[,-1])
      } else if (!useAdjustment && use_analytical_posterior) {
        ## sample directly from the posterior
        Ca <- rbinom(nrow(Na), size=1, prob=pD_star)
        Pp.mu.a <- t(cbind(m_starN, m_starD))
        Pp.Pa <- abind::abind(C_starN, C_starD, along=3)
        for (i in 1:ne) {
          distclass <- Ca[i] + 1
          Na[i,] <- mvtnorm::rmvnorm(1, mean=Pp.mu.a[distclass,], sigma=Pp.Pa[,,distclass])
        }
        Na <- tobit(Na)
      } else if (useAdjustment && !use_analytical_posterior) {
        Na <- multi.ens.adj(
          Xf = Nf,
          cf = d.forecast,
          mu.f = rbind(mufN, mufD),
          Pf   = abind::abind(CfN,
                   CfD,
                   along=3),
          Xa = Xa,
          ca = ca
        )
        Ca <- Na[,1]
        Na <- tobit(Na[,-1])
      } else {
        stop("useAdjustment = FALSE and use_analytical_posterior = FALSE is not supported")
      }
      
    } else {
      Na <- tobit(Na)
      Ca <- rep(0, ne) ## dummy data
    }
    
    ## Compute post-analysis means and CIs
    if (run.adjustment) { ## if we did run analysis (since we had data)
      if (use_analytical_posterior) { ## if we used analytical posterior
        AnalysisCIs <- matrix(NA, nrow=5, ncol=3)
        quantiles <- c(0.025,0.25,0.5,0.75,0.975)
        if (pD_star > pN_star) { # compute analysis CIs from most likely component
          for (i in 1:3) {
            AnalysisCIs[,i] <- normal_quantile(quantiles, m_starD[i], C_starD[i, i])
          }
        } else {
          for (i in 1:3) {
            AnalysisCIs[,i] <- normal_quantile(quantiles, m_starN[i], C_starN[i, i])
          }
        }
        mu.post <- pN_star * m_starN + pD_star * m_starD
        mu.post.D <- sum(Ca == 1, na.rm = TRUE) / length(Ca)
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = AnalysisCIs,
                              m_starN = m_starN,
                              m_starD = m_starD,
                              mu = mu.post,
                              muD = pD_star)
      } else { ## if we used MCMC
        mu.post.D <- sum(ca == 1, na.rm = TRUE) / length(ca)
        # print(paste0("year ", t))
        # print("means of Xa")
        # print(colMeans(Xa))
        # print("0.5 quantiles of Xa")
        # print(apply(Xa,2,quantile,c(0.5)))
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = apply(Xa,2,quantile,c(0.025,0.25,0.5,0.75,0.975)),
                              m_starN = colMeans(Xa[ca==0,]),
                              m_starD = colMeans(Xa[ca==1,]),
                              muD = mu.post.D)
      }
    } else { ## if we didn't have data
      AnalysisCIs <- apply(Na,2,quantile,c(0.025,0.25,0.5,0.75,0.975)) # apply quantile directly to forecasts
      mu.post <- apply(Na,2,mean,na.rm=TRUE)
      mu.post.D <- 0.1
      if (use_analytical_posterior) {
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = AnalysisCIs,
                              m_starN = m_starN,
                              m_starD = m_starD,
                              mu = mu.post,
                              muD = 0.1)
      } else {
        Analysis[[t]] <- list(Na = Na,
                              Ca = Ca,
                              CI = AnalysisCIs,
                              mu = mu.post,
                              muD = mu.post.D)
      }
    }
  }
  return(list(Forecast=Forecast,Analysis=Analysis,Y=Y, ProcessErrors=ProcessErrors))
}
```




## initial conditions & observations
```{r}
## BIOMASS
B2C = 0.5 ##biomass to carbon ratio
Yw <- bextract*B2C*1000*1e-4 ## convert Mg/ha (Mg per hectacre)  to  kg C [356 (num ensemble members) x 28 (num years observed)]
Rw <- (84*B2C*1000*1e-4)^2 ## Observation Error: Kennedy et al 2018 Environ. Res. Lett. 13 025004; Fig 2 RMSE
## units: Mg/ha -> kg/m2
print(dim(Yw))
## grab the distribution of AGB for the first time point from all pixels
AGBprior <- Yw[,1]
ne = length(AGBprior) ## number of ensemble members

## spin up
PARlong <- as.vector(t(rad)) # concatenate PAR time series for each site into one vector
spin = VSEM(theta,PARlong) # Question: why run VSEM on the concatenated PAR? What is theta?
thin = seq(nrow(spin)/2,nrow(spin),length=ne)
spinIC = spin[thin,2:4] # select ne samples from the spin-up 
#for(i in 2:4){
#  hist(spin[thin,i],type='l')
#}

## disturbance recovery trajectory
nNBG = 200 ## years
NBG <- array(NA,c(ne,nNBG,3))
for(i in 1:ne){
  param = theta
  param[9:11] <- spinIC[i,]*c(0.01,1,0.01) ## set leaf and wood to 1% (simulate disturbance)
  raw = VSEM(param,PARlong[1:(nNBG*365.25)]) # simulate for nNBG years
  NBG[i,,] = raw[round(seq(365,nNBG*365.25,length=nNBG)),2:4]  ## grab pools at the end of each year
}
  
## for each AGBprior, grab the matching leaf and soil
ICprior <- matrix(NA,ne,3)
age = rep(NA,ne)
for(i in 1:ne){
  age[i] <- which.min((AGBprior[i]-NBG[i,,3])^2) # age = year in the disturbance recover trajectory where the simulated site has similar AGB to the AGB prior information
  ICprior[i,] <- c(NBG[i,age[i],1:2],AGBprior[i]) # use AGBprior for the model ICs
}
hist(age)
colnames(ICprior) <- c("Cv","Cs","CR") 
pairs(ICprior)

```


### wrapper function to run data assimilation
```{r}
##' @description runDA: function to run data assimilation given R / Q / P multipliers
##' @param Rmult -- multiplier for observation error, R
##' @param Qmult -- multiplier for process error, Q
##' @param Pmult -- multiplier for disturbance probability
##' @param updateQ if true, conditionally update the process error Q based on observations
##' @param useAdjustment
##' @param use_analytical_posterior
##' @param siteIndices indices of Oregon sites to run with
##' @param expName name of experiment to use in output file names
runDA <- function(Rmult, Qmult, Pmult, updateQ, useAdjustment, use_analytical_posterior, siteIndices, expName) {
  for(i in siteIndices){
    ofile = paste0(expName,"_OR_SD_",
                   formatC(i, width = 3, format = "d", flag = "0"),
                   "_P",Pmult,"Q",Qmult,"R",Rmult,".RData")
    print(ofile)
    NT = floor(ncol(rad)/365)
    Yi = matrix(NA,NT,3)  ## observed data
    Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed (Question: where do Yi observations come from?)
    R = diag(rep(Rw,3))*Rmult    ## observation error
    Q = diag(c(Ql,Qsoil,Qw)^2)*Qmult  ## process error
    dtype = which(names(d.prior) == dextract[i,"t2004"])
    if(length(dtype)==0) break
    dist <- list(              ## disturbance parameters
      p.dist = d.prior[dtype]*Pmult,     ## disturbance probability (Question: where do disturbance probabilities come from?)
      mu0 = c(0.2,PBBAR[4,dtype]),   ## % retention [leaf,agb], first parameter made up, should include soil
      V0 =  diag(c(0.15,PBVAR[4,dtype])^2), #variance in retention Question(where do the woody biomass mean / variance parameters come from?)
      alloc.soil = 0.05        ## made up  
    )
    test = SSSD.model(Y = Yi,R=R,Q=Q,
                    PAR=rad[i,],Xic=ICprior,THETA=theta[1:11],
                    dist = dist, useAdjustment=useAdjustment, updateQ=updateQ, 
                    use_analytical_posterior = use_analytical_posterior)
    save(test,file=ofile)
  }
}
```


## experiment: use_analytical_posterior=FALSE / updateQ=FALSE
```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1)
Pmult = c(1)
#i = 181  # 177:263 = #fire

# site indices
od = c(83:176,264:356)
fullRun = c(83:356)
testRun = fullRun[seq(1, length(fullRun), by = 6)]
testHighAbsDist = c(107, 125, 137, 149, 161) # sites with large disturbances
testOneSite = c(107)
expGrid <- expand.grid(Rmult=Rmult, Qmult=Qmult, Pmult=Pmult)
for (i in seq_len(nrow(expGrid))) {
  row <- expGrid[i, ]  # extract a combo of R / Q / P
  runDA(Rmult=row$Rmult, Qmult=row$Qmult, Pmult=row$Pmult, updateQ=FALSE, useAdjustment=TRUE, use_analytical_posterior=FALSE, siteIndices=testOneSite, expName="AFalseQFalse") # run data assimilation experiment
}
```

```{r}
library(stringr)

i = 181
pool.names=c("Cleaf","Csoil","Cstem")
runs <- dir(pattern="*.RData")
runs <- runs[grepl("^AFalseQFalse_OR",runs)]
print(runs)

#load("Analytical_OR_SD_263_P1Q1R0.001.RData")
# runs <- runs[grepl("84",runs)]
for (runfile in runs) {
  load(runfile)
  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
      run.muN = sapply(X = test$Analysis, function(x) {
        y = x$m_starN
        y[i]
      })
      run.muD = sapply(X = test$Analysis, function(x) {
        y = x$m_starD
        y[i]
      })
      
      run.probD = sapply(X = test$Analysis, function(x) {
        y = x$muD
      }) 
  
      
      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=pool.names[i],xlab="time", main=runfile)

    ecoforecastR::ciEnvelope(1:ncol(CI.Na),
                             CI.Na[1,], CI.Na[5,],
                             col=ecoforecastR::col.alpha('#b2df8a',0.7))
    
    # Median CI.Na line
    lines(CI.Na[3,], col="#b2df8a", lwd=2)
    
    # μN line
    lines(unlist(run.muN), col="#40826D", lwd=2)
    
    # μD line
    lines(unlist(run.muD), col="#9D2933", lwd=2)
    
    # Observations
    points(test$Y[,i], pch = 1)
    
    # === ADD LEGEND HERE ===
    legend("topright",
           legend = c("CI.mean", "muN", "muD", "observed"),
           col    = c("#b2df8a", "#40826D", "#9D2933", "black"),
           lwd    = c(2, 2, 2, NA),
           pch    = c(NA, NA, NA, 1),
           bty    = "n")
    # ========================
    plot(run.probD, main=paste0(runfile, " disturbance probs"))
    lines(run.probD, col="#9D2933", lwd=2)
    # lines(CI.D[3,])
  }
  
}
```


## experiment use_analytical_posterior == FALSE, updateQ==TRUE
```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1)
Pmult = c(1)
#i = 181  # 177:263 = #fire

# site indices
od = c(83:176,264:356)
fullRun = c(83:356)
testRun = fullRun[seq(1, length(fullRun), by = 6)]
testHighAbsDist = c(107, 125, 137, 149, 161) # sites with large disturbances
testOneSite = c(107)
expGrid <- expand.grid(Rmult=Rmult, Qmult=Qmult, Pmult=Pmult)
for (i in seq_len(nrow(expGrid))) {
  row <- expGrid[i, ]  # extract a combo of R / Q / P
  runDA(Rmult=row$Rmult, Qmult=row$Qmult, Pmult=row$Pmult, updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=FALSE, siteIndices=testOneSite, expName="AFalseQTrue") # run data assimilation experiment
}
```


```{r}
library(stringr)

i = 181
pool.names=c("Cleaf","Csoil","Cstem")
runs <- dir(pattern="*.RData")
runs <- runs[grepl("^AFalseQTrue_OR",runs)]
print(runs)

#load("Analytical_OR_SD_263_P1Q1R0.001.RData")
# runs <- runs[grepl("84",runs)]
for (runfile in runs) {
  load(runfile)
  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
      run.muN = sapply(X = test$Analysis, function(x) {
        y = x$m_starN
        y[i]
      })
      run.muD = sapply(X = test$Analysis, function(x) {
        y = x$m_starD
        y[i]
      })
      
      run.probD = sapply(X = test$Analysis, function(x) {
        y = x$muD
      }) 
      
      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=pool.names[i],xlab="time", main=runfile)

    ecoforecastR::ciEnvelope(1:ncol(CI.Na),
                             CI.Na[1,], CI.Na[5,],
                             col=ecoforecastR::col.alpha('#b2df8a',0.7))
    
    # Median CI.Na line
    lines(CI.Na[3,], col="#b2df8a", lwd=2)
    
    # μN line
    lines(unlist(run.muN), col="#40826D", lwd=2)
    
    # μD line
    lines(unlist(run.muD), col="#9D2933", lwd=2)
    
    # Observations
    points(test$Y[,i], pch = 1)
    
    # === ADD LEGEND HERE ===
    legend("topright",
           legend = c("CI.mean", "muN", "muD", "observed"),
           col    = c("#b2df8a", "#40826D", "#9D2933", "black"),
           lwd    = c(2, 2, 2, NA),
           pch    = c(NA, NA, NA, 1),
           bty    = "n")
    # ========================
    plot(run.probD, main=paste0(runfile, " disturbance probs"))
    lines(run.probD, col="#9D2933", lwd=2)
    
    plot(unlist(test$ProcessErrors), main=paste0(runfile, " process errors"))
    lines(unlist(test$ProcessErrors), col="#FF005F")
    # lines(CI.D[3,])
  }
  
}
```

## use_analytical_posterior=TRUE / updateQ = FALSE

```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1)
Pmult = c(1)
#i = 181  # 177:263 = #fire

# site indices
od = c(83:176,264:356)
fullRun = c(83:356)
testRun = fullRun[seq(1, length(fullRun), by = 6)]
testHighAbsDist = c(107, 125, 137, 149, 161) # sites with large disturbances
testOneSite = c(107)
expGrid <- expand.grid(Rmult=Rmult, Qmult=Qmult, Pmult=Pmult)
for (i in seq_len(nrow(expGrid))) {
  row <- expGrid[i, ]  # extract a combo of R / Q / P
  runDA(Rmult=row$Rmult, Qmult=row$Qmult, Pmult=row$Pmult, updateQ=FALSE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=testOneSite, expName="ATrueQFalse") # run data assimilation experiment
}
```


```{r}
library(stringr)

i = 181
pool.names=c("Cleaf","Csoil","Cstem")
runs <- dir(pattern="*.RData")
runs <- runs[grepl("^ATrueQFalse_OR",runs)]
print(runs)

#load("Analytical_OR_SD_263_P1Q1R0.001.RData")
# runs <- runs[grepl("84",runs)]
for (runfile in runs) {
  load(runfile)
  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
      run.muN = sapply(X = test$Analysis, function(x) {
        x$m_starN[i]
      })
      run.muD = sapply(X = test$Analysis, function(x) {
        x$m_starD[i]
      })
      run.probD = sapply(X = test$Analysis, function(x) {
        x$muD
      }) 
  
      
      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=pool.names[i],xlab="time", main=runfile)

    ecoforecastR::ciEnvelope(1:ncol(CI.Na),
                             CI.Na[1,], CI.Na[5,],
                             col=ecoforecastR::col.alpha('#b2df8a',0.7))
    
    # Median CI.Na line
    lines(CI.Na[3,], col="#b2df8a", lwd=2)
    
    # μN line
    lines(run.muN, col="#40826D", lwd=2)
    
    # μD line
    lines(run.muD, col="#9D2933", lwd=2)
    
    # Observations
    points(test$Y[,i], pch = 1)
    
    # === ADD LEGEND HERE ===
    legend("topright",
           legend = c("CI.mean", "muN", "muD", "observed"),
           col    = c("#b2df8a", "#40826D", "#9D2933", "black"),
           lwd    = c(2, 2, 2, NA),
           pch    = c(NA, NA, NA, 1),
           bty    = "n")
    # ========================
    plot(run.probD, main=paste0(runfile, " disturbance probs"))
    lines(run.probD, col="#9D2933", lwd=2)
    # lines(CI.D[3,])
  }
  
}
```

## use_analytical_posterior=TRUE, updateQ=TRUE
```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1)
Pmult = c(1)
#i = 181  # 177:263 = #fire

# site indices
od = c(83:176,264:356)
fullRun = c(83:356)
testRun = fullRun[seq(1, length(fullRun), by = 6)]
testHighAbsDist = c(107, 125, 137, 149, 161) # sites with large disturbances
testOneSite = c(107)
expGrid <- expand.grid(Rmult=Rmult, Qmult=Qmult, Pmult=Pmult)
for (i in seq_len(nrow(expGrid))) {
  row <- expGrid[i, ]  # extract a combo of R / Q / P
  runDA(Rmult=row$Rmult, Qmult=row$Qmult, Pmult=row$Pmult, updateQ=TRUE, useAdjustment=TRUE, use_analytical_posterior=TRUE, siteIndices=testOneSite, expName="ATrueQTrue") # run data assimilation experiment
}
```

```{r}
library(stringr)

i = 181
pool.names=c("Cleaf","Csoil","Cstem")
runs <- dir(pattern="*.RData")
runs <- runs[grepl("^ATrueQTrue_OR",runs)]
print(runs)

#load("Analytical_OR_SD_263_P1Q1R0.001.RData")
# runs <- runs[grepl("84",runs)]
for (runfile in runs) {
  load(runfile)
  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
      run.muN = sapply(X = test$Analysis, function(x) {
        x$m_starN[i]
      })
      run.muD = sapply(X = test$Analysis, function(x) {
        x$m_starD[i]
      })
      run.probD = sapply(X = test$Analysis, function(x) {
        x$muD
      }) 
  
      
      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=pool.names[i],xlab="time", main=runfile)

    ecoforecastR::ciEnvelope(1:ncol(CI.Na),
                             CI.Na[1,], CI.Na[5,],
                             col=ecoforecastR::col.alpha('#b2df8a',0.7))
    
    # Median CI.Na line
    lines(CI.Na[3,], col="#b2df8a", lwd=2)
    
    # μN line
    lines(run.muN, col="#40826D", lwd=2)
    
    # μD line
    lines(run.muD, col="#9D2933", lwd=2)
    
    # Observations
    points(test$Y[,i], pch = 1)
    
    # === ADD LEGEND HERE ===
    legend("topright",
           legend = c("CI.mean", "muN", "muD", "observed"),
           col    = c("#b2df8a", "#40826D", "#9D2933", "black"),
           lwd    = c(2, 2, 2, NA),
           pch    = c(NA, NA, NA, 1),
           bty    = "n")
    # ========================
    plot(run.probD, main=paste0(runfile, " disturbance probs"))
    lines(run.probD, col="#9D2933", lwd=2)
    
    plot(unlist(test$ProcessErrors), main=paste0(runfile, " process errors"))
    lines(unlist(test$ProcessErrors), col="#FF005F")
  }
  
}
```

## analyze sites - just one disturbance
```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1)
Pmult = c(1)
#i = 181  # 177:263 = #fire

# site indices
od = c(83:176,264:356)
fullRun = c(83:356)
testRun = fullRun[seq(1, length(fullRun), by = 6)]
testHighAbsDist = c(107, 125, 137, 149, 161) # sites with large disturbances
expGrid <- expand.grid(Rmult=Rmult, Qmult=Qmult, Pmult=Pmult)
for (i in seq_len(nrow(expGrid))) {
  row <- expGrid[i, ]  # extract a combo of R / Q / P
  runDA(Rmult=row$Rmult, Qmult=row$Qmult, Pmult=row$Pmult, updateQ=TRUE, useAdjustment=FALSE, siteIndices=testHighAbsDist) # run data assimilation experiment
}
```

```{r}


# runDA(Rmult=0.001, Qmult=1, Pmult=1, updateQ=TRUE, siteIndices=263)

load("Analytical_OR_SD_107_P1Q1R0.001.RData")
print("Site 107 R0.001")
print(test$Analysis[[15]]$muD)
for (i in 1:28) {
  print(test$Analysis[[i]]$muD)
}

print("Site 107 R1")
load("Analytical_OR_SD_107_P1Q1R1.RData")
print(test$Analysis[[15]]$muD)
for (i in 1:28) {
  print(test$Analysis[[i]]$muD)
}

print("Site 125 R0.001")
load("Analytical_OR_SD_125_P1Q1R0.001.RData")
print(test$Analysis[[15]]$muD)
for (i in 1:28) {
  print(test$Analysis[[i]]$muD)
}

print("Site 125 R1")
load("Analytical_OR_SD_125_P1Q1R1.RData")
print(test$Analysis[[15]]$muD)
for (i in 1:28) {
  print(test$Analysis[[i]]$muD)
}
# for (i in 1:28) {
#   print(test$Analysis[[i]]$muD)
# }



```


## visualization
```{r}
library(stringr)

i = 181
pool.names=c("Cleaf","Csoil","Cstem")
runs <- dir(pattern="*.RData")
runs <- runs[grepl("^Analytical_OR",runs)]
print(runs)

#load("Analytical_OR_SD_263_P1Q1R0.001.RData")
# runs <- runs[grepl("84",runs)]
for (runfile in runs) {
  load(runfile)
  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
      run.muN = sapply(X = test$Analysis, function(x) {
        x$m_starN[i]
      })
      run.muD = sapply(X = test$Analysis, function(x) {
        x$m_starD[i]
      })
      run.probD = sapply(X = test$Analysis, function(x) {
        x$muD
      }) 
  
      
      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
     ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
     type='n',
     ylab=pool.names[i],xlab="time", main=runfile)

    ecoforecastR::ciEnvelope(1:ncol(CI.Na),
                             CI.Na[1,], CI.Na[5,],
                             col=ecoforecastR::col.alpha('#b2df8a',0.7))
    
    # Median CI.Na line
    lines(CI.Na[3,], col="#b2df8a", lwd=2)
    
    # μN line
    lines(run.muN, col="#40826D", lwd=2)
    
    # μD line
    lines(run.muD, col="#9D2933", lwd=2)
    
    # Observations
    points(test$Y[,i], pch = 1)
    
    # === ADD LEGEND HERE ===
    legend("topright",
           legend = c("CI.mean", "muN", "muD", "observed"),
           col    = c("#b2df8a", "#40826D", "#9D2933", "black"),
           lwd    = c(2, 2, 2, NA),
           pch    = c(NA, NA, NA, 1),
           bty    = "n")
    # ========================
    plot(run.probD, main=paste0(runfile, " disturbance probs"))
    lines(run.probD, col="#9D2933", lwd=2)
  }
  
}

```

 TODO:
 * why is Analysis not contracting around the data?
    * reducing obs error by /10 didn't solve
 * why are the disturbance forecast pools crazy?
 * assess disturbance posterior

## Computer experiments
* Run under defaults
* Run under a grid of inc process error and decrease obs error and increase disturbance probability

## postprocessing

```{r}
## Detect set of experiments that have been run
runs <- dir(pattern="*.RData")
runs <- runs[grepl("^Analytical_OR",runs)]
print(runs)
exp = sapply(strsplit(runs,"_",fixed=TRUE),function(x){
  sub(".RData","",x[5],fixed = TRUE)
})
exps <- unique(exp)
print(exps) # the combos of P / Q / R multipliers experimented with
## load experiments that haven't yet been summarized
elist <- list()   ## run once
t04 = 15            ## time of disturbance assessement (2004)
tlo = 28            ## time of last observation
for(i in seq_along(exps)){
  if(is.null(elist[[exps[i]]])){
    esel <- which(exp==exps[i])
    distrate = rep(NA,nrow(dextract)) # 1 x (number of sites)
    divergence = predist = distsize = matrix(NA,nrow(dextract),6) ## pre and post disturbance biomass pools (w/uncertainty) nsite x 6
    for(j in seq_along(esel)){
      ## load file
      load(runs[esel[j]])
      k = as.numeric(unlist(strsplit(runs[esel[j]],"_",fixed=TRUE))[4])
      ## extract detection rate
      distrate[k]=test$Analysis[[t04]]$muD # posterior probability of disturbance

      ## capture of disturbance change
      predist[k,1]    = test$Analysis[[t04-1]]$mu[3]
      predist[k,2:6]  = test$Analysis[[t04-1]]$CI[,3]
      distsize[k,1]   = test$Analysis[[t04]]$mu[3]
      distsize[k,2:6] = test$Analysis[[t04]]$CI[,3]
      
      ## Filter divergence
      divergence[k,1]   = test$Analysis[[tlo]]$mu[3]
      divergence[k,2:6] = test$Analysis[[tlo]]$CI[,3]
    } ## end loop over individual runs
    
    elist[[exps[i]]] = list(distrate=distrate,predist=predist,
                            distsize=distsize,divergence=divergence)
    
  } ## end if not processed
  
} ## end loop over experiments

## grab matching data
obs.pre  = Yw[,t04-1] # 1 x nsite (observation before disturbance)
obs.dist = Yw[,t04] # observation during disturbance
obs.tlo  = Yw[,tlo] # observation during last timestep
obs.class= dextract[,"t2004"]/10 # 1 x nsite
save(elist,obs.pre,obs.dist,obs.tlo,obs.class,file="AnalyticalElist.RData")
```


## analyses

```{r}
pdf("Analytical_OR_exps_diagnostics.pdf")
trim <- function(x,lo=0.0001,hi=0.9999) {
  x[x<lo] <- lo
  x[x>hi] <- hi
  return(x)
}
betatrim <- function(y){
    n.obs <- sum(!is.na(y))
    (y * (n.obs - 1) + 0.5) / n.obs
}
stats <- list()
for(i in seq_along(exps)){
  ## divergence
  div = (elist[[i]]$divergence[,1] - obs.tlo)/
    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5) # 1 x nsite (compute filter divergence)
  div.reg    = lm(div~obs.tlo) # linear model for div in terms of obs.tlo
  div.median = median(div,na.rm = TRUE) # compute median divergence
  div.cov    = sum(div < 1 & div > -1,na.rm=TRUE)/sum(!is.na(div)) # % of sites with filter divergence outside of (-1, 1)
  
  
  hist(div,main=exps[i])
  abline(v=c(-1,1),lty=2)
  plot(obs.tlo,div,col=obs.class,main=exps[i])
  abline(h=c(-1,1),lty=2)
  abline(div.reg,col=2)
  
  ## Disturbance detection
  distrate = elist[[i]]$distrate
  hist(distrate,main=exps[i])
  
  ## Disturbance skill
  ord <- order(distrate)
  plot(distrate[ord],col=obs.class[ord],main=exps[i])
  dist.bar   = mean(distrate,na.rm = TRUE)
  dist.class = tapply(distrate,obs.class,mean,na.rm=TRUE)
  
  ## vs predisturbance biomass
  distrate2 = betatrim(distrate)
  pred.seq = data.frame(obs.pre=seq(0,max(obs.pre,na.rm = TRUE),length=500),
                        distprop = seq(-1,1,length=500))
  dist.reg.pre  = lm(distrate ~ obs.pre)
  dist.beta.pre = betareg::betareg(distrate2 ~ obs.pre,
                                   na.action=na.omit,link="logit")
  plot(obs.pre,distrate,col=obs.class,main=exps[i])
  abline(dist.reg.pre,col=2)
  lines(pred.seq$obs.pre,predict(dist.beta.pre,newdata=pred.seq),col="green")
  
  ## vs proportional disturbance size
  distprop = 1 - elist[[i]]$distsize[,1]/elist[[i]]$predist[,1] ## predicted
  distprop = 1 - obs.dist/obs.pre ## observed
  dist.beta.prop = betareg::betareg(distrate2 ~ distprop,
                                   na.action=na.omit,link="logit")
  plot(distprop,distrate,col=obs.class,main=exps[i])
  lines(pred.seq$distprop,predict(dist.beta.prop,newdata=pred.seq),col="green")
  
  ## vs ABSOLUTE disturbance size
  distabs = obs.pre-obs.dist
  pred.seq$distabs = seq(min(distabs, na.rm = TRUE),max(distabs, na.rm = TRUE),length=500)
  dist.beta.abs = betareg::betareg(distrate2 ~ distabs | distabs,
                                   na.action=na.omit,
                                   link="logit",
                                   link.phi = "log",
                                   start=c(0,5,0.5,0),
                                   type="BR")
  dist.qb.abs = glm(distrate ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
  dist.bin.abs = glm(round(distrate) ~ distabs,
                    na.action=na.omit,
                    family=binomial("logit")
  )
  plot(distabs,distrate,col=obs.class,main=exps[i])
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile"),col="green")
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.75),col="green",lty=2)
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.25),col="green",lty=2)
  lines(pred.seq$distabs,predict(dist.qb.abs,newdata=pred.seq,type="response"),col="purple")
  lines(pred.seq$distabs,predict(dist.bin.abs,newdata=pred.seq,type="response"),col="blue")

  ## vs divergence
  absdiv = abs(div)
  dist.beta.absdiv = betareg::betareg(distrate2 ~ absdiv,
                                   na.action=na.omit,link="logit")
  dist.bin.absdiv = glm(round(distrate) ~ absdiv,
                                   na.action=na.omit,
                        family = binomial("logit"))
  pred.seq$absdiv = seq(0,max(absdiv,na.rm=TRUE),length=500)
  plot(absdiv,distrate,col=obs.class,main=exps[i])
  abline(v=c(-1,1),lty=2,col=3)
  abline(lm(distrate ~ absdiv),col=2)
  lines(pred.seq$absdiv,predict(dist.beta.absdiv,newdata=pred.seq),col="green")
  lines(pred.seq$absdiv,predict(dist.bin.absdiv,newdata=pred.seq,type="response"),col="blue")

  dist.qb.combo = glm(distrate2 ~ obs.pre + distprop +
                        distabs + absdiv + obs.class,
                      na.action=na.omit,
                      family=quasibinomial("logit"))
  dist.bin.combo = glm(round(distrate) ~ obs.pre + distprop +
                        distabs + absdiv + as.factor(obs.class),
                      na.action=na.omit,
                      family=binomial("logit"))
  
# dist.beta.combo = betareg::betareg(distrate2 ~ obs.pre + obs.class + distabs + absdiv,
#                                   na.action=na.omit,link="logit")
  
  stats[[i]] <- list(
    div.reg = div.reg,
    div.median = div.median,
    div.cov = div.cov,
    dist.bar = dist.bar,
    dist.class = dist.class,
    dist.pre =  dist.beta.pre,
    dist.prop = dist.beta.prop,
    dist.beta.abs =  dist.beta.abs,
    dist.bin.abs  = dist.bin.abs,
    dist.absdiv   = dist.beta.absdiv,
    dist.combo = dist.bin.combo
  )

}
dev.off()
save(stats,file="AnalyticalStats.RData")
```

## quick check: find and plot sites with high distabs, but distrate = 0
```{r}
print(exps)

# stats[[1]] = P1Q1R0.001
distabs <- obs.pre-obs.dist
distrate <- elist[[1]]$distrate

# get sites with distabs > 5 but distrate = 0
allsites <- 1:356
distratezero <- distrate == 0
distratezero[is.na(distratezero)] = FALSE
print(length(distabs))
print(length(distratezero))
print(distabs & distratezero)
testsites <- allsites[distabs > 5 & distratezero]

# plot DA for those sites
for (j in testsites) {
  pool.names=c("Cleaf","Csoil","Cstem")
# runs <- dir(pattern="*.RData")
# runs <- runs[grepl("^Analytical_OR",runs)]
# print(runs)
  runfile <- paste0("Analytical_OR_SD_", j, "_P1Q1R0.001.RData")
  load(runfile)
# runs <- runs[grepl("84",runs)]

  for(i in 3){
      CI.Na = sapply(X = test$Analysis,function(x){
        y = x$CI
        y[,i]
      })
  
      
      # CI.Na.2 = sapply(X = test$Analysis, function(x) {
      #   y = x$mcmc
      #   sel.n = which(colnames(y) == paste0("n[",i,"]"))
      #   quantile(y[,sel.n],c(0.025,0.5,0.975))
      # })
    plot(test$Y[,i],
         ylim = range(rbind(CI.Na,test$Y[,i]),na.rm=TRUE),
         type='n',
         ylab=pool.names[i],xlab="time", main=runfile)
    # ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
    ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
    # ecoforecastR::ciEnvelope(1:ncol(CI.Na.2),CI.Na.2[1,],CI.Na.2[3,],col=ecoforecastR::col.alpha("red",0.7))
    lines(CI.Na[3,],col="#b2df8a",lwd=2)
    points(test$Y[,i])
    # lines(CI.D[3,])
}
}
```

factors affecting detection of disturbance
```{r}
Ps = as.numeric(substr(names(elist),2,2))
Qs = as.numeric(substr(names(elist),4,4))
Rs = as.numeric(substr(names(elist),6,6))

## P

## Q

## R

## Divergence
div.cov = sapply(stats,"[[","div.cov")
plot(Rmult[Rs],div.cov,col=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage")
legend("topright",legend=paste0(Qmult[1:2],"x"))

div.median = sapply(stats,"[[","div.median")
plot(Rmult[Rs],div.median,col=Qs,log="x",xlab="Variance Multiplier",ylab="Median Divergance")
legend("topleft",legend=paste0(Qmult[1:2],"x"))

div.slope = sapply(stats,function(x){coef(x$div.reg)[2]})
plot(Rmult[Rs],div.slope,col=Qs,log="x",xlab="Variance Multiplier",ylab="Divergance Slope")
## Disturbance @ default
summary(stats[[4]]$dist.combo)
## nothing matters

## Disturbance @ P1Q2R2 (intermediate)
summary(stats[[4]]$dist.combo)
## detection varies with divergence & disturbance type

## Disturbance magnitude: P1Q1R4
i = 3
summary(stats[[i]]$dist.combo)  ## absolute magnitude of divergence is best predictor
summary(stats[[i]]$dist.bin.abs)
beta = coef(stats[[i]]$dist.bin.abs)
## threshold mx+b = 0
thresh = -beta[1]/beta[2] 

## Disturbance type: P1Q1R4 (reduced divergence)
stats[[i]]$dist.class
tapply(distabs,obs.class,summary) ## obs dist size by class
thresh.prop = tapply(distabs,obs.class,function(x){sum(x>thresh,na.rm = TRUE)/sum(!is.na(x))}) ## prop > detection threshold
thresh.prop
stats[[i]]$dist.class
dclass = stats[[i]]$dist.class
#dclass[1] = 0
dcol = as.numeric(names(stats[[i]]$dist.class)) + 1
plot(thresh.prop,dclass,col=dcol,pch=19)
abline(0,1,lty=2,col=2)
legend("topleft",legend = dcol-1,col=dcol,pch=19)


## additional experiments to run (P1Q1R4)
```

false positives

false negatives

classification success

## preprocess timeseries example for visualization
```{r}
## visualization to help select a site to look at
plotly::plot_ly(data.frame(obs.pre=obs.pre,obs.post=obs.dist),x=~obs.pre,y=~obs.post)
which.min((obs.pre-13.3)^2+(obs.dist-5.1)^2) ## select a point

## settings
site = 188
exp.name = "P1Q1R4" #"P1Q1R1"  ## manually run for each experiment **

## data
NT = 33
Yi = matrix(NA,NT,3)  ## observed data
Yi[1:ncol(Yw),3] = Yw[site,]  ## only Cw observed
  
## model
load(paste0("OR_SD_",site,"_",exp.name,".RData"),envir = viz)
pool.names=c("Cleaf","Csoil","Cstem")
test = viz$test

for(i in 1:3){
  
    CI.Na = sapply(X = test$Analysis,function(x){
      y = x$CI
      sel.n = which(colnames(y) == paste0("n[",i,"]"))
      y[,sel.n]
    })
    
    CI.D = sapply(X = test$Analysis,function(x){
      y = x$CI
      sel.n = which(colnames(y) == paste0("D"))
      y[,sel.n]
    })

    CI.Nf = sapply(X = test$Forecast, function(x){
      mu = x$mufN[i]
      sd = 1/sqrt(x$pfN[i,i])
      return(mu + c(-1.96,0,1.96)*sd)
    })
    
  plot(Yi[,i],
       ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
       type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[3,],col="#b2df8a",lwd=2)
  points(Yi[,i])
  lines(CI.D[3,])
  
  
}
save(Yi,CI.Na,CI.D,file=paste0("OR_TS_",site,"_",exp.name,".RData"))
```

## summary info for filter divergence figures
```{r}
exp.name = "P1Q1R4" #"P1Q1R1"

for(i in seq_along(exps)){
  
  ## divergence
  div = (elist[[i]]$divergence[,1] - obs.tlo)/
    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
  div.reg    = lm(div~obs.tlo)
  div.median = median(div,na.rm = TRUE)
  div.cov    = sum(div < 1 & div > -1,na.rm=TRUE)/sum(!is.na(div))
  
}
```


## Next steps:
* Run with multiple disturbance options and PFT switching
