---
title: "OR_joint_update_analysis"
output: html_document
author: "Jacob Epstein"
---

Postprocessing + analyses for the joint-update runs saved by
`run_joint_update_sites()` in `OR_joint_update.Rmd`. Each run is an `hist` list
saved to `experiments/jointUpdate/site_<idx>_distprod_<TF>_<source>_<model>.rds`,
containing per-timestep vectors `probD`, `overall.analysis`, and the 5-quantile
wood-pool CI matrix `ci5`.

## set up

```{r}
B2C <- 0.5
load("extract.h03v04.RData")                 # bextract, dextract, points
dtime <- 1985:2012
colnames(dextract) <- paste0("t", dtime)
dextract[is.na(dextract)] <- 0
Yw <- bextract * B2C * 1000 * 1e-4            # nsite x nyears observed

indir <- "experiments/jointUpdateYear15"
t04 <- 15          # reference disturbance-assessment timestep (fixed reference year)
tlo <- 28          # last-observation timestep
qprobs <- c(0.025, 0.25, 0.5, 0.75, 0.975)
```

## postprocessing

```{r}
## Detect the joint-update run files and parse site index + experiment label
## (experiment = distprod + Rparams source + model, e.g. "distprod_TRUE_GEDI_het").
runs  <- dir(indir, pattern = "^site_.*\\.rds$")
parts <- strsplit(sub("\\.rds$", "", runs), "_", fixed = TRUE)
sites <- sapply(parts, function(x) as.numeric(x[2]))
exp   <- sapply(parts, function(x) paste(x[3:length(x)], collapse = "_"))
exps  <- unique(exp)
print(exps)

elist <- list()
for (i in seq_along(exps)) {
  esel  <- which(exp == exps[i])
  nsite <- nrow(dextract)
  distrate   <- rep(NA, nsite)                              # posterior prob. of disturbance
  nisbar     <- rep(NA, nsite)                              # per-site average NIS over timesteps
  cfbar      <- q33bar <- Rbar <- rep(NA, nsite)            # per-site mean S components
  innov2bar  <- rep(NA, nsite)                              # per-site mean squared innovation
  predist    <- distsize <- divergence <- matrix(NA, nsite, 6)  # mean + 5 quantiles
  for (j in seq_along(esel)) {
    res <- readRDS(file.path(indir, runs[esel[j]]))
    k   <- sites[esel[j]]

    ## detection rate at the assessment year
    distrate[k]      <- res$probD[t04]

    ## average normalized innovation squared over observed timesteps
    nisbar[k]        <- mean(res$nis, na.rm = TRUE)

    ## per-site mean innovation-variance components (S = CF + q33 + R)
    cfbar[k]         <- mean(res$nis.cf,  na.rm = TRUE)
    q33bar[k]        <- mean(res$nis.q33, na.rm = TRUE)
    Rbar[k]          <- mean(res$nis.R,   na.rm = TRUE)

    ## mean squared innovation (numerator of NIS) over observed timesteps
    innov2bar[k]     <- mean(res$nis.innov^2, na.rm = TRUE)

    ## capture of the disturbance change (pre / during)
    predist[k, 1]    <- res$overall.analysis[t04 - 1]; predist[k, 2:6]    <- res$ci5[t04 - 1, ]
    distsize[k, 1]   <- res$overall.analysis[t04];     distsize[k, 2:6]   <- res$ci5[t04, ]

    ## filter divergence at the last observation
    divergence[k, 1] <- res$overall.analysis[tlo];     divergence[k, 2:6] <- res$ci5[tlo, ]
  }
  elist[[exps[i]]] <- list(distrate = distrate, nisbar = nisbar,
                           cfbar = cfbar, q33bar = q33bar, Rbar = Rbar,
                           innov2bar = innov2bar,
                           predist = predist,
                           distsize = distsize, divergence = divergence)
}

## grab matching observations
obs.pre   <- Yw[, t04 - 1]            # observation before disturbance
obs.dist  <- Yw[, t04]               # observation during disturbance
obs.tlo   <- Yw[, tlo]               # observation at last timestep
obs.class <- dextract[, "t2004"] / 10
save(elist, obs.pre, obs.dist, obs.tlo, obs.class,
     file = file.path(indir, "elist.RData"))
```

## analyses

```{r}
pdf(file.path(indir, "exps_diagnostics.pdf"))
trim <- function(x, lo = 0.0001, hi = 0.9999) {
  x[x < lo] <- lo
  x[x > hi] <- hi
  return(x)
}
betatrim <- function(y) {
  n.obs <- sum(!is.na(y))
  (y * (n.obs - 1) + 0.5) / n.obs
}
stats <- list()

classnames <- c("other", "cut", "road", "fire", "biotic")  # disturbance types
class_vals <- sort(unique(obs.class))
class_vals <- class_vals[class_vals != 0]                   # drop 0 (no disturbance)
classes_in_data <- classnames[class_vals]

## colour by disturbance class; offset by 1 so class 0 (undisturbed) isn't col 0,
## which base R draws in the (invisible) background colour.
class.cols <- c("grey70", "black", "red", "green3", "blue", "cyan", "magenta")  # index = class+1
pt.col <- class.cols[obs.class + 1]

## 95% interval for the mean of n.obs per-step NIS values (chi-sq_N / N)
n.obs  <- ncol(Yw)
nis.ci <- qchisq(c(0.025, 0.975), df = n.obs) / n.obs

for (i in seq_along(exps)) {
  ## normalized innovation squared (per-site average; replaces filter divergence)
  div = elist[[i]]$nisbar
  div.reg    = lm(div ~ obs.tlo)
  div.median = median(div, na.rm = TRUE)
  div.cov    = sum(div > nis.ci[1] & div < nis.ci[2], na.rm = TRUE) / sum(!is.na(div))

  plot(obs.tlo, div, col = pt.col, main = exps[i],
       ylab = "mean NIS", ylim = range(c(0, div, nis.ci), na.rm = TRUE))
  legend("topright", legend = classes_in_data, col = class.cols[class_vals + 1],
         pch = 21, bty = "n", title = "disturbance type")
  abline(h = nis.ci, lty = 2)            # 95% chi-sq_N / N interval
  abline(div.reg, col = 2)

  ## innovation-variance decomposition: stacked bars of mean S = CF + q33 + R per site,
  ## ordered by total S, so it's clear which term dominates (q33 is the Qmult-deflatable one)
  cf  <- elist[[i]]$cfbar; q33 <- elist[[i]]$q33bar; Rv <- elist[[i]]$Rbar
  ok  <- which(is.finite(cf) & is.finite(q33) & is.finite(Rv))
  if (length(ok) > 0) {
    ord <- ok[order(cf[ok] + q33[ok] + Rv[ok])]
    ord <- ord[seq(1, length(ord), by = 10)]   # every 10th site (1/10) for legible bars
    barplot(rbind(cf[ord], q33[ord], Rv[ord]),
            names.arg = ord, las = 2, cex.names = 0.6,
            col = c("grey60", "orange", "steelblue"),
            main = paste(exps[i], "- mean innovation variance S (1/10 of sites)"),
            xlab = "site (ordered by total S)", ylab = "variance (kgC m-2)^2")
    legend("topleft", bty = "n",
           legend = c("CF (forecast spread)", "q33 (process error)", "R (obs error)"),
           fill = c("grey60", "orange", "steelblue"))
  }

  ## Disturbance detection
  distrate <- elist[[i]]$distrate

  ## Disturbance skill
  ord <- order(distrate)
  dist.bar   = mean(distrate, na.rm = TRUE)
  dist.class = tapply(distrate, obs.class, mean, na.rm = TRUE)

  ## vs predisturbance biomass
  distrate2 = betatrim(distrate)
  pred.seq = data.frame(obs.pre = seq(0, max(obs.pre, na.rm = TRUE), length = 500),
                        distprop = seq(-1, 1, length = 500))
  dist.reg.pre  = lm(distrate ~ obs.pre)
  dist.beta.pre = betareg::betareg(distrate2 ~ obs.pre,
                                   na.action = na.omit, link = "logit")

  ## vs proportional disturbance size
  distprop = 1 - elist[[i]]$distsize[, 1] / elist[[i]]$predist[, 1]  ## predicted
  distprop = 1 - obs.dist / obs.pre                                  ## observed
  dist.beta.prop = betareg::betareg(distrate2 ~ distprop,
                                    na.action = na.omit, link = "logit")

  ## vs ABSOLUTE disturbance size
  distabs = obs.pre - obs.dist
  pred.seq$distabs = seq(min(distabs, na.rm = TRUE), max(distabs, na.rm = TRUE), length = 500)
  dist.beta.abs = betareg::betareg(distrate2 ~ distabs | distabs,
                                   na.action = na.omit,
                                   link = "logit",
                                   link.phi = "log",
                                   start = c(0, 5, 0.5, 0),
                                   type = "BR")
  dist.qb.abs = glm(distrate2 ~ distabs,
                    na.action = na.omit,
                    family = quasibinomial("logit"))
  dist.bin.abs = glm(round(distrate) ~ distabs,
                     na.action = na.omit,
                     family = binomial("logit"))
  plot(distabs, distrate, col = pt.col, main = exps[i])
  legend("topright", legend = classes_in_data, col = class.cols[class_vals + 1],
         pch = 21, bty = "n", title = "disturbance type")
  lines(pred.seq$distabs, predict(dist.beta.abs, newdata = pred.seq, type = "quantile"), col = "green")
  lines(pred.seq$distabs, predict(dist.beta.abs, newdata = pred.seq, type = "quantile", at = 0.75), col = "green", lty = 2)
  lines(pred.seq$distabs, predict(dist.beta.abs, newdata = pred.seq, type = "quantile", at = 0.25), col = "green", lty = 2)
  lines(pred.seq$distabs, predict(dist.qb.abs, newdata = pred.seq, type = "response"), col = "purple")
  lines(pred.seq$distabs, predict(dist.bin.abs, newdata = pred.seq, type = "response"), col = "blue")
  #
  # ## vs divergence
  absdiv = abs(div)
  dist.beta.absdiv = betareg::betareg(distrate2 ~ absdiv,
                                       na.action = na.omit, link = "logit")
  dist.bin.absdiv = glm(round(distrate) ~ absdiv,
                         na.action = na.omit,
                         family = binomial("logit"))
  pred.seq$absdiv = seq(0, max(absdiv, na.rm = TRUE), length = 500)
  #
  dist.qb.combo = glm(distrate2 ~ obs.pre + distprop +
                         distabs + absdiv + obs.class,
                       na.action = na.omit,
                       family = quasibinomial("logit"))
  # dist.bin.combo = glm(round(distrate) ~ obs.pre + distprop +
  #                         distabs + absdiv + as.factor(obs.class),
  #                       na.action = na.omit,
  #                       family = binomial("logit"))

  stats[[i]] <- list(
    div.reg = div.reg,
    div.median = div.median,
    div.cov = div.cov,
    dist.bar = dist.bar,
    dist.class = dist.class,
    dist.pre =  dist.beta.pre,
    dist.prop = dist.beta.prop,
    dist.beta.abs =  dist.beta.abs,
    dist.bin.abs  = dist.bin.abs,
    dist.absdiv   = dist.beta.absdiv
    #dist.combo = dist.bin.combo
  )
}
dev.off()
save(stats, file = file.path(indir, "stats.RData"))
```

## summary across experiments

```{r}
## One summary PDF, three pages (one per plot type). Each page lays out all
## experiments side by side with CONSISTENT axes so they can be compared
## directly; every panel is titled with its experiment name.
pdf(file.path(indir, "exps_summary.pdf"), width = 14, height = 7)

distabs <- obs.pre - obs.dist            # absolute disturbance size (obs-derived, common to all exps)

## panel grid: up to 8 experiments -> 2 x 4
nexp <- length(exps)
grid <- c(2, 4)

## shared axis ranges across all experiments
all.nis  <- unlist(lapply(elist, function(e) e$nisbar))
nis.ylim <- range(c(0, all.nis, nis.ci), na.rm = TRUE)
tlo.xlim <- range(obs.tlo, na.rm = TRUE)
all.S    <- unlist(lapply(elist, function(e) e$cfbar + e$q33bar + e$Rbar))
S.ylim   <- c(0, max(all.S, na.rm = TRUE))
distabs.xlim <- range(distabs, na.rm = TRUE)

## ---- Page 1: mean NIS vs final-year biomass ----
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  div <- elist[[i]]$nisbar
  plot(obs.tlo, div, col = pt.col, pch = 19, cex = 0.5,
       xlim = tlo.xlim, ylim = nis.ylim,
       xlab = "final-year biomass", ylab = "mean NIS", main = exps[i])
  abline(h = nis.ci, lty = 2)            # 95% chi-sq_N / N interval
}
mtext("mean NIS vs final-year biomass (dashed = 95% chi-sq/N)", outer = TRUE, cex = 1.2)

## ---- Page 2: innovation-variance decomposition (1/10 of sites) ----
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  cf <- elist[[i]]$cfbar; q33 <- elist[[i]]$q33bar; Rv <- elist[[i]]$Rbar
  ok <- which(is.finite(cf) & is.finite(q33) & is.finite(Rv))
  if (length(ok) == 0) { plot.new(); title(main = exps[i]); next }
  ord <- ok[order(cf[ok] + q33[ok] + Rv[ok])]
  ord <- tail(ord, 20)                          # top 20 sites by total S for legible bars
  barplot(rbind(cf[ord], q33[ord], Rv[ord]),
          names.arg = ord, las = 2, cex.names = 0.5, ylim = S.ylim,
          col = c("grey60", "orange", "steelblue"),
          xlab = "site (top 20 by total S)", ylab = "variance (kgC m-2)^2",
          main = exps[i])
  if (i == 1)
    legend("topleft", bty = "n", cex = 0.8,
           legend = c("CF (forecast spread)", "q33 (process error)", "R (obs error)"),
           fill = c("grey60", "orange", "steelblue"))
}
mtext("innovation variance S = CF + q33 + R (top 20 sites by total S)", outer = TRUE, cex = 1.2)

## ---- Page 3: innovation-variance decomposition (bottom 50 sites) ----
## shared ylim based on the bottom-50 totals (the global S.ylim is driven by the
## top sites and would render these tiny), so bars stay visible and comparable.
bot.S <- unlist(lapply(elist, function(e) {
  ok  <- which(is.finite(e$cfbar) & is.finite(e$q33bar) & is.finite(e$Rbar))
  head(sort((e$cfbar + e$q33bar + e$Rbar)[ok]), 50)
}))
botS.ylim <- c(0, max(bot.S, na.rm = TRUE))
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  cf <- elist[[i]]$cfbar; q33 <- elist[[i]]$q33bar; Rv <- elist[[i]]$Rbar
  ok <- which(is.finite(cf) & is.finite(q33) & is.finite(Rv))
  if (length(ok) == 0) { plot.new(); title(main = exps[i]); next }
  ord <- ok[order(cf[ok] + q33[ok] + Rv[ok])]
  ord <- head(ord, 50)                          # bottom 50 sites by total S
  barplot(rbind(cf[ord], q33[ord], Rv[ord]),
          names.arg = ord, las = 2, cex.names = 0.4, ylim = botS.ylim,
          col = c("grey60", "orange", "steelblue"),
          xlab = "site (bottom 50 by total S)", ylab = "variance (kgC m-2)^2",
          main = exps[i])
  if (i == 1)
    legend("topleft", bty = "n", cex = 0.8,
           legend = c("CF (forecast spread)", "q33 (process error)", "R (obs error)"),
           fill = c("grey60", "orange", "steelblue"))
}
mtext("innovation variance S = CF + q33 + R (bottom 50 sites by total S)", outer = TRUE, cex = 1.2)

## ---- Page 4: disturbance detection vs absolute disturbance size ----
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  distrate <- elist[[i]]$distrate
  plot(distabs, distrate, col = pt.col, pch = 19, cex = 0.5,
       xlim = distabs.xlim, ylim = c(0, 1),
       xlab = "absolute disturbance size", ylab = "P(disturbance)", main = exps[i])
  fit <- try(glm(round(distrate) ~ distabs, family = binomial("logit"),
                 na.action = na.omit), silent = TRUE)
  if (!inherits(fit, "try-error")) {
    xx <- seq(distabs.xlim[1], distabs.xlim[2], length = 200)
    lines(xx, predict(fit, newdata = data.frame(distabs = xx), type = "response"), col = "blue")
  }
  if (i == 1)
    legend("topright", legend = classes_in_data, col = class.cols[class_vals + 1],
           pch = 19, bty = "n", cex = 0.8, title = "disturbance type")
}
mtext("disturbance detection vs absolute disturbance size", outer = TRUE, cex = 1.2)

## ---- Page 5: mean squared innovation (top 20 sites), vs total S ----
## NIS = mean(innov^2)/S. Bars = mean squared innovation (numerator); the points
## overlay total S (= CF+q33+R, the denominator) for the same sites, so where bars
## rise above the points NIS > 1 (the bias-not-variance signature, esp. LandTrendr).
i2.ymax <- 0
for (i in seq_along(exps)) {
  e  <- elist[[i]]
  Stot <- e$cfbar + e$q33bar + e$Rbar
  ok <- which(is.finite(e$innov2bar) & is.finite(Stot))
  if (length(ok) == 0) next
  ord <- tail(ok[order(e$innov2bar[ok])], 20)
  i2.ymax <- max(i2.ymax, e$innov2bar[ord], Stot[ord], na.rm = TRUE)
}
i2.ylim <- c(0, i2.ymax)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e  <- elist[[i]]
  Stot <- e$cfbar + e$q33bar + e$Rbar
  ok <- which(is.finite(e$innov2bar) & is.finite(Stot))
  if (length(ok) == 0) { plot.new(); title(main = exps[i]); next }
  ord <- tail(ok[order(e$innov2bar[ok])], 20)        # top 20 sites by mean innov^2
  bp  <- barplot(e$innov2bar[ord], names.arg = ord, las = 2, cex.names = 0.5,
                 ylim = i2.ylim, col = "tomato",
                 xlab = "site (top 20 by mean innov^2)",
                 ylab = "mean innovation^2 (kgC m-2)^2", main = exps[i])
  points(bp, Stot[ord], pch = 19, cex = 0.7, col = "black")   # total S for comparison
  if (i == 1)
    legend("topleft", bty = "n", cex = 0.8,
           legend = c("mean innov^2 (NIS numerator)", "total S (NIS denominator)"),
           pch = c(15, 19), col = c("tomato", "black"))
}
mtext("mean squared innovation vs total S (top 20 sites by innov^2)", outer = TRUE, cex = 1.2)

par(mfrow = c(1, 1))
dev.off()
```

## disturbance product: where does probD change, and why does it go down?

The summary plots read `probD` at a FIXED assessment year `t04`, but for almost all
sites that year is UNflagged by the product, so they miss the (large) boost that occurs
at each site's actual flagged years. This cell compares with/without the product (a) at
`t04` and (b) averaged over each site's own flagged years, reading the per-year `probD`
straight from the run files.

```{r}
## TRUE runs come from `indir`; the distprod=FALSE baseline (R = GEDI/het) is read
## from a separate directory that holds the no-product runs.
falsedir <- "experiments/jointUpdate"

pdf(file.path(indir, "distprod_TRUE_vs_FALSE.pdf"), width = 10, height = 5)

## source_model combos that have BOTH a TRUE (indir) and a FALSE (falsedir) run
combos <- unique(sub("^distprod_(TRUE|FALSE)_", "", exps))
for (cb in combos) {
  tf <- dir(indir,    pattern = sprintf("_distprod_TRUE_%s\\.rds$",  cb))
  ff <- dir(falsedir, pattern = sprintf("_distprod_FALSE_%s\\.rds$", cb))
  sT <- as.numeric(sapply(strsplit(tf, "_"), `[`, 2))
  sF <- as.numeric(sapply(strsplit(ff, "_"), `[`, 2))
  both <- intersect(sT, sF)
  if (length(both) == 0) next

  pT04 <- pF04 <- pTfl <- pFfl <- rep(NA, length(both))
  flag04 <- logical(length(both))
  for (m in seq_along(both)) {
    s  <- both[m]
    rT <- readRDS(file.path(indir,    sprintf("site_%d_distprod_TRUE_%s.rds",  s, cb)))
    rF <- readRDS(file.path(falsedir, sprintf("site_%d_distprod_FALSE_%s.rds", s, cb)))
    fyrs <- which(dextract[s, ] > 0)               # this site's flagged years (dextract cols)
    fyrs <- fyrs[fyrs <= length(rT$probD)]
    pT04[m]   <- rT$probD[t04]; pF04[m] <- rF$probD[t04]
    flag04[m] <- dextract[s, t04] > 0
    if (length(fyrs) > 0) {
      pTfl[m] <- mean(rT$probD[fyrs], na.rm = TRUE)
      pFfl[m] <- mean(rF$probD[fyrs], na.rm = TRUE)
    }
  }

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  ## (1) at the fixed assessment year t04: points below the diagonal went DOWN.
  down <- sum(pT04 < pF04 - 1e-6, na.rm = TRUE)
  plot(pF04, pT04, col = ifelse(flag04, "red", "grey50"), pch = 19, cex = 0.6,
       xlim = c(0, 1), ylim = c(0, 1),
       xlab = "probD without product", ylab = "probD with product",
       main = sprintf("%s @ t04=%d  (%d sites down)", cb, t04, down))
  abline(0, 1, lty = 2)
  legend("topleft", legend = c("flagged @ t04", "not flagged @ t04"),
         col = c("red", "grey50"), pch = 19, bty = "n", cex = 0.8)
  ## (2) averaged over each site's OWN flagged years: the boost the t04 view misses.
  plot(pFfl, pTfl, col = "darkgreen", pch = 19, cex = 0.6,
       xlim = c(0, 1), ylim = c(0, 1),
       xlab = "mean probD without product", ylab = "mean probD with product",
       main = sprintf("%s, over flagged years", cb))
  abline(0, 1, lty = 2)
}
dev.off()
```

## when were sites flagged as disturbed (by type and year)?

One barplot per disturbance type: the number of sites flagged in each year. The
assessment timestep `t04` is emphasized (red bar + red axis label) since `probD` is read
there — this makes clear how few flags of each type actually fall on that year.

```{r}
dist.codes <- c("10" = "other", "20" = "cut", "30" = "road", "40" = "fire", "50" = "biotic")
codes.present <- names(dist.codes)[sapply(names(dist.codes),
                   function(cc) any(dextract == as.numeric(cc)))]   # types that occur
NTy <- ncol(dextract)

par(mfrow = c(length(codes.present), 1), mar = c(3, 4, 2, 1), oma = c(2, 0, 2, 0))
for (cc in codes.present) {
  cnt  <- colSums(dextract == as.numeric(cc))
  bcol <- rep("steelblue", NTy); bcol[t04] <- "red"          # emphasize the assessment year
  bp   <- barplot(cnt, col = bcol, border = NA, names.arg = dtime, las = 2,
                  cex.names = 0.6, ylab = "# sites flagged",
                  main = sprintf("%s (code %s) - %d flags total", dist.codes[cc], cc, sum(cnt)))
  ## red label marking t04 on the x axis
  axis(1, at = bp[t04], labels = sprintf("t04=%d (%d)", t04, dtime[t04]),
       col.axis = "red", col.ticks = "red", tick = TRUE, line = 0.2, cex.axis = 0.7)
}
mtext("disturbance flags by type and year (red bar = assessment timestep t04)",
      outer = TRUE, cex = 1.1)
```
