---
title: "OR_joint_update_analysis"
output: html_document
author: "Jacob Epstein"
---

Postprocessing + analyses for the joint-update runs saved by the experiment grid in
`OR_joint_update2.Rmd`. Each run is the `hist` list returned by `ORJointUpdate`, saved
to `<indir>/site_<idx>_distprod_<TF>_<source>_<model>_tau<prior>.rds`, with per-timestep

* `year`, `obs`: calendar year of each DA timestep and the assimilated observation;
* `mean`, `var`, `ci95`: wood-pool analysis mean, variance and 95% interval (2.5% / 97.5%
  quantiles of the analysis mixture), marginalised over the tau posterior and the
  mixture components;
* `pDist`: posterior probability of disturbance;
* `pit`: probability integral transform of each observation under its one-step-ahead
  predictive (before that observation is assimilated, integrating over process error).

The formulas are written up in `claude-working-docs/or_joint_update2_analysis_plots.pdf`.

## set up

```{r}
B2C <- 0.5
load("extract.h03v04.RData")                 # bextract, dextract, points
## The products cover different years, so every lookup goes through calendar year:
##   dextract (disturbance product): 1985:2012
##   bextract (biomass / Yw):        1990:2017
##   DA timesteps:                   stored per run in hist$year (1990:2017)
dtime <- 1985:2012
ctime <- 1990:2017
colnames(dextract) <- paste0("t", dtime)
dextract[is.na(dextract)] <- 0
Yw <- bextract * B2C * 1000 * 1e-4            # nsite x nyears observed
colnames(Yw) <- paste0("t", ctime)

indir   <- "experiments/fullRun"
yr.dist <- 2004               # disturbance-assessment year

## turn an experiment label (e.g. "distprod_TRUE_GEDI_het_taugamma") into a readable title
## (e.g. "Disturbance Product, Heteroskedastic R (GEDI magnitude)").
pretty.exp <- function(e) {
  p    <- strsplit(e, "_", fixed = TRUE)[[1]]   # distprod, TRUE/FALSE, source, het/con, tau<prior>[, bias]
  prod <- if (length(p) >= 2 && p[2] == "TRUE") "Disturbance Product" else "No Disturbance Product"
  src  <- if (length(p) >= 3) p[3] else "?"
  rmod <- if (length(p) >= 4 && p[4] == "het") "Heteroskedastic R" else "Constant R"
  bias <- if ("bias" %in% p) ", bias-corrected" else ""
  sprintf("%s, %s (%s magnitude)%s", prod, rmod, src, bias)
}
```

## postprocessing

```{r}
## Detect the joint-update run files and parse site index + experiment label
## (experiment = distprod + Rparams source + model + tau prior).
runs  <- dir(indir, pattern = "^site_.*\\.rds$")
parts <- strsplit(sub("\\.rds$", "", runs), "_", fixed = TRUE)
sites <- sapply(parts, function(x) as.numeric(x[2]))
exp   <- sapply(parts, function(x) paste(x[3:length(x)], collapse = "_"))
exps  <- sort(unique(exp))
print(exps)

nsite <- nrow(dextract)
elist <- list()
for (i in seq_along(exps)) {
  esel <- which(exp == exps[i])
  pdist04 <- div <- pitT <- obsT <- rep(NA, nsite)
  pit.all <- c()
  for (j in esel) {
    k <- sites[j]
    if (is.na(k) || k < 1 || k > nsite) next     # skip runs from a different extract
    res <- readRDS(file.path(indir, runs[j]))
    if (is.null(res$year)) stop(runs[j], ": no $year axis -- stale run from before the 1990 alignment fix; rerun it")
    if (is.null(res$ci95)) stop(runs[j], ": no $ci95 -- run from before the filter-divergence change; rerun it")

    ## detection: posterior P(disturbed) in 2004
    pdist04[k] <- res$pDist[res$year == yr.dist]

    ## filter divergence at the last observed timestep T:
    ## (y_T - analysis mean) / (half the width of the 95% analysis interval)
    tT      <- max(which(!is.na(res$obs)))
    obsT[k] <- res$obs[tT]
    div[k]  <- (res$obs[tT] - res$mean[tT]) / ((res$ci95[tT, "hi"] - res$ci95[tT, "lo"]) / 2)

    ## PIT of the last observation under its one-step-ahead predictive, and all PITs
    pitT[k] <- res$pit[tT]
    pit.all <- c(pit.all, res$pit[!is.na(res$pit)])
  }
  elist[[exps[i]]] <- list(pdist04 = pdist04, div = div, pitT = pitT, obsT = obsT,
                           pit.all = pit.all)
}

## matching observations
obs.pre   <- Yw[, paste0("t", yr.dist - 1)]   # 2003 AGB
obs.dist  <- Yw[, paste0("t", yr.dist)]       # 2004 AGB
distabs   <- obs.pre - obs.dist               # absolute AGB change 2003 -> 2004 (loss > 0)
obs.class <- dextract[, paste0("t", yr.dist)] / 10   # 2004 disturbance-product label
save(elist, obs.pre, obs.dist, obs.class, distabs,
     file = file.path(indir, "elist.RData"))
```

## analyses

Summary PDF: one page per plot, every experiment drawn as its own panel with shared
axes. Points are coloured by the 2004 disturbance-product label.

```{r}
classnames  <- c("other", "cut", "road", "fire", "biotic")          # codes 10..50
class.cols  <- c("grey60", "black", "red", "green3", "blue", "cyan")  # index = class + 1
pt.col      <- class.cols[obs.class + 1]
class.vals  <- sort(unique(obs.class[obs.class != 0]))
legend.labels <- c("undisturbed", classnames[class.vals])
legend.cols   <- c(class.cols[1], class.cols[class.vals + 1])

nexp <- length(exps)
grid <- c(2, ceiling(nexp / 2))
has  <- lapply(elist, function(e) which(is.finite(e$div)))   # sites run per experiment
all.sites <- sort(unique(unlist(has)))
pct  <- function(x) sprintf("%.0f%%", 100 * mean(x, na.rm = TRUE))
pct1 <- function(x) sprintf("%.1f%%", 100 * mean(x, na.rm = TRUE))

## Trend of P(disturbed) in 2004 vs AGB loss: a penalised-spline logistic GAM
## (quasibinomial, logit link, k = 5 basis, smoothness chosen by REML), fitted to the
## sites the product flags in 2004. (Unflagged sites are left out: with the product they
## sit near P = 0 at any loss, so a curve through both groups would fit neither.)
## Not fitted with fewer than min.fit.sites sites. P is clamped to [0.001, 0.999] first: runs that
## saturate at exactly 0 or 1 would otherwise push the logit fit to +-Inf there (and its
## confidence band to the whole [0, 1]), and some P exceed 1 by rounding error.
flag04 <- obs.class > 0
min.fit.sites <- 10
fit_trend <- function(p, grp) {
  p <- pmin(pmax(p, 0.001), 0.999)
  d <- data.frame(p = p, x = distabs)[grp & is.finite(p) & is.finite(distabs), ]
  if (nrow(d) < min.fit.sites || length(unique(d$x)) < 5) return(NULL)
  tryCatch(mgcv::gam(p ~ s(x, k = 5), family = quasibinomial, data = d, method = "REML"),
           error = function(e) { warning("trend fit failed: ", conditionMessage(e)); NULL })
}
## mean curve and its 95% confidence band over the group's own range of loss
trend_curve <- function(fit, n = 200) {
  xx <- seq(min(fit$model$x), max(fit$model$x), length.out = n)
  pr <- predict(fit, data.frame(x = xx), se.fit = TRUE)
  list(x = xx, mean = plogis(pr$fit),
       lo = plogis(pr$fit - 1.96 * pr$se.fit), hi = plogis(pr$fit + 1.96 * pr$se.fit))
}
## loss at which the mean curve first crosses P = 0.5; -Inf / Inf if it stays above /
## below 0.5 over the fitted range, NA without a fit
x50_value <- function(fit) {
  if (is.null(fit)) return(NA_real_)
  cv <- trend_curve(fit, 1000); up <- cv$mean >= 0.5
  if (all(up))  return(-Inf)
  if (!any(up)) return(Inf)
  i <- which(diff(up) != 0)[1]
  approx(cv$mean[i:(i + 1)], cv$x[i:(i + 1)], 0.5)$y
}
x50_label <- function(fit) {
  v <- x50_value(fit)
  if (is.na(v)) return("--")
  if (v == -Inf) return(sprintf("<%.1f", min(fit$model$x)))
  if (v ==  Inf) return(sprintf(">%.1f", max(fit$model$x)))
  sprintf("%.2f", v)
}
stats <- list()
for (i in seq_along(exps)) stats[[exps[i]]] <- list(flag = fit_trend(elist[[i]]$pdist04, flag04))

pdf(file.path(indir, "exps_summary.pdf"), width = 14, height = 7)
obsT.xlim <- range(unlist(lapply(elist, function(e) e$obsT)), na.rm = TRUE)

## ---- Page 1: filter divergence per site ----
## (y_T - analysis mean_T) / half-width of the 95% analysis interval; inside [-1, 1]
## means the final observation falls inside the interval.
div.ylim <- range(c(-1.5, 1.5, unlist(lapply(elist, function(e) e$div))), na.rm = TRUE)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e <- elist[[i]]
  plot(e$obsT, e$div, col = pt.col, pch = 19, xlim = obsT.xlim, ylim = div.ylim,
       xlab = "final observed AGB (kgC m-2)", ylab = "filter divergence", cex.main = 0.8,
       main = sprintf("%s\n%s of sites in [-1, 1]", pretty.exp(exps[i]), pct(abs(e$div) <= 1)))
  rect(par("usr")[1], -1, par("usr")[2], 1, col = adjustcolor("grey", 0.25), border = NA)
  abline(h = 0, lty = 2)
  if (i == 1)
    legend("topright", legend = legend.labels, col = legend.cols,
           pch = 19, bty = "n", cex = 0.8, title = "2004 product label")
}
mtext("filter divergence at the final observation (grey = [-1, 1])", outer = TRUE, cex = 1.2)

## ---- Page 2: PIT calibration of the final observation, pooled over all runs ----
## u = P(y' <= y_T | y_{1:T-1}) for every (experiment, site) run. If the one-step-ahead
## predictive is calibrated, u ~ Uniform(0, 1), so (left) the histogram of u over all runs
## is flat, and (right) the central p-interval of the predictive contains y_T in a
## fraction p of runs, for every p. Grey bands: 95% range expected from binomial
## sampling noise alone if the predictive were calibrated.
pitT.exp  <- lapply(elist, function(e) e$pitT[is.finite(e$pitT)])
pitT.pool <- unlist(pitT.exp, use.names = FALSE)
n.pool    <- length(pitT.pool)
exp.cols  <- hcl.colors(nexp, "Dark 3")
layout(matrix(c(1, 2, 3, 3), 2, byrow = TRUE), heights = c(5, 1))   # legend strip below
par(mar = c(4.5, 4.5, 3.5, 1), oma = c(0, 0, 2, 0))

## (a) histogram of u, all runs pooled
nb   <- 20; br <- seq(0, 1, length.out = nb + 1)
dens <- hist(pitT.pool, breaks = br, plot = FALSE)$counts * nb / n.pool
band <- qbinom(c(0.025, 0.975), n.pool, 1 / nb) * nb / n.pool
in95 <- qbinom(c(0.025, 0.975), n.pool, 0.95) / n.pool
plot(NA, xlim = c(0, 1), ylim = c(0, max(dens, band[2]) * 1.1), xaxs = "i",
     xlab = "PIT u of the final observation", ylab = "density", cex.main = 0.9,
     main = sprintf("all %d runs pooled\n%s in the central 95%% (expected %.1f-%.1f%%)",
                    n.pool, pct1(abs(2 * pitT.pool - 1) <= 0.95), 100 * in95[1], 100 * in95[2]))
rect(0, band[1], 1, band[2], col = adjustcolor("grey", 0.35), border = NA)
rect(br[-(nb + 1)], 0, br[-1], dens, col = "grey55", border = "white")
abline(h = 1, lty = 2)

## (b) central-interval coverage error, per experiment and pooled
lev   <- c(seq(0.05, 0.95, by = 0.05), 0.99)
cover <- function(u) sapply(lev, function(p) mean(abs(2 * u - 1) <= p))
n.exp <- min(lengths(pitT.exp))
bandE <- sapply(lev, function(p) qbinom(c(0.025, 0.975), n.exp, p) / n.exp - p)
bandP <- sapply(lev, function(p) qbinom(c(0.025, 0.975), n.pool, p) / n.pool - p)
err   <- sapply(pitT.exp, function(u) cover(u) - lev)
errP  <- cover(pitT.pool) - lev
plot(NA, xlim = c(0, 1), ylim = range(c(bandE, err, errP)), cex.main = 0.9,
     xlab = "nominal coverage p of the central predictive interval",
     ylab = "empirical - nominal coverage",
     main = "coverage error (> 0: predictive too wide, < 0: too narrow)")
polygon(c(lev, rev(lev)), c(bandE[1, ], rev(bandE[2, ])), col = adjustcolor("grey", 0.25), border = NA)
polygon(c(lev, rev(lev)), c(bandP[1, ], rev(bandP[2, ])), col = adjustcolor("grey", 0.45), border = NA)
abline(h = 0, lty = 2)
matlines(lev, err, col = exp.cols, lty = 1, lwd = 1.3)
lines(lev, errP, lwd = 3)
par(mar = c(0, 0, 0, 0)); plot.new()
legend("center", bty = "n", cex = 0.75, ncol = 3, lwd = c(3, rep(1.3, nexp)),
       col = c("black", exp.cols),
       legend = c(sprintf("all runs pooled (n = %d)", n.pool),
                  sprintf("%s (n = %d)", sapply(exps, pretty.exp), lengths(pitT.exp))))
mtext(sprintf("one-step-ahead PIT calibration of the final observation, %d experiments x %d sites",
              nexp, length(all.sites)), outer = TRUE, cex = 1.2)
layout(1)

## ---- Page 3: PIT histogram over all sites and years ----
## flat at density 1 if calibrated; U-shape = predictive too narrow, hump = too wide,
## slope = biased.
pit.br <- seq(0, 1, by = 0.05)
pit.ymax <- max(sapply(elist, function(e) max(hist(e$pit.all, breaks = pit.br, plot = FALSE)$density)))
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e <- elist[[i]]
  hist(e$pit.all, breaks = pit.br, freq = FALSE, col = "grey80", border = "white",
       ylim = c(0, pit.ymax), xlab = "PIT u", cex.main = 0.8,
       main = sprintf("%s\n%d site-years", pretty.exp(exps[i]), length(e$pit.all)))
  abline(h = 1, lty = 2)
}
mtext("one-step-ahead PIT of every observation, all sites and years (dashed = uniform)",
      outer = TRUE, cex = 1.2)

## ---- Page 4: disturbance detection vs absolute disturbance size ----
distabs.xlim <- range(distabs[all.sites], na.rm = TRUE)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  plot(distabs, elist[[i]]$pdist04, col = pt.col, pch = 19,
       xlim = distabs.xlim, ylim = c(0, 1), cex.main = 0.8,
       xlab = "AGB loss 2003 -> 2004 (kgC m-2)", ylab = "P(disturbed) in 2004",
       main = pretty.exp(exps[i]))
  abline(v = 0, col = "grey80")
  fit <- stats[[i]]$flag
  if (!is.null(fit)) {
    cv <- trend_curve(fit)
    polygon(c(cv$x, rev(cv$x)), c(cv$lo, rev(cv$hi)), col = adjustcolor("black", 0.15), border = NA)
    lines(cv$x, cv$mean, lwd = 2.5)
    ## dashed drop line from the x-axis up to where the trend crosses P = 0.5
    x50 <- x50_value(fit)
    if (is.finite(x50)) {
      segments(x50, par("usr")[3], x50, 0.5, lty = 2, lwd = 1.5)
      lab <- sprintf("%.2f", x50); w <- strwidth(lab, cex = 0.8); h <- strheight(lab, cex = 0.8)
      x0  <- x50 + 0.4 * strwidth("m", cex = 0.8)                 # boxed label right of the line
      rect(x0, 0.5 - 0.9 * h, x0 + w * 1.3, 0.5 + 0.9 * h, col = "white", border = "grey40")
      text(x0 + w * 0.65, 0.5, lab, cex = 0.8)
    }
  }
  if (i == 1) {
    legend("topleft", legend = legend.labels, col = legend.cols,
           pch = 19, bty = "n", cex = 0.8, title = "2004 product label")
    legend("bottomright", legend = c("trend, sites flagged in 2004", "loss at P = 0.5"),
           lty = c(1, 2), lwd = c(2.5, 1.5), bty = "n", cex = 0.8)
  }
}
mtext("disturbance detection vs absolute disturbance size (line: logistic GAM mean for sites flagged in 2004, band = 95% CI)",
      outer = TRUE, cex = 1.2)

## ---- Page 5: summary table ----
## AGB loss at which the flagged-site trend curve crosses P = 0.5
x50 <- sapply(stats, function(s) x50_label(s$flag))
tab <- data.frame(
  experiment = sapply(exps, pretty.exp),
  div.in     = sapply(elist, function(e) pct(abs(e$div) <= 1)),
  div.med    = sapply(elist, function(e) sprintf("%.2f", median(abs(e$div), na.rm = TRUE))),
  pit.in     = sapply(elist, function(e) pct(abs(2 * e$pitT - 1) <= 0.95)),
  pit.all.in = sapply(elist, function(e) pct(abs(2 * e$pit.all - 1) <= 0.95)),
  x50        = x50)
## pooled over every run of every experiment
pool <- function(f) unlist(lapply(elist, `[[`, f), use.names = FALSE)
tab <- rbind(tab, data.frame(
  experiment = "All experiments pooled",
  div.in     = pct1(abs(pool("div")) <= 1),
  div.med    = sprintf("%.2f", median(abs(pool("div")), na.rm = TRUE)),
  pit.in     = pct1(abs(2 * pool("pitT") - 1) <= 0.95),
  pit.all.in = pct1(abs(2 * pool("pit.all") - 1) <= 0.95),
  x50        = "--"))
heads <- c("Experiment", "|div| <= 1", "median |div|", "final PIT\nin 95%",
           "all PITs\nin 95%", "loss at P = 0.5\n(flagged sites)")
par(mfrow = c(1, 1), mar = c(1, 1, 4, 1))
plot.new()
title("Summary by experiment")
xs <- c(0.01, 0.46, 0.56, 0.66, 0.76, 0.86)
yy <- 0.86 - seq_len(nrow(tab)) * min(0.08, 0.8 / nrow(tab))
text(xs, 0.92, heads, adj = c(0, 0.5), font = 2, cex = 0.8)
for (c in seq_along(xs)) text(xs[c], yy, tab[[c]], adj = 0, cex = 0.8)
segments(0, c(0.97, 0.87, mean(yy[nexp + 0:1])), 1, c(0.97, 0.87, mean(yy[nexp + 0:1])), col = "grey55")
dev.off()
save(stats, tab, file = file.path(indir, "stats.RData"))
```

## disturbance product: where does pDist change?

Compares the runs with and without the disturbance product (same R model and tau prior)
(a) at the assessment year `yr.dist` (2004) and (b) averaged over each site's own flagged years.

```{r}
pdf(file.path(indir, "distprod_TRUE_vs_FALSE.pdf"), width = 10, height = 5)
combos <- unique(sub("^distprod_(TRUE|FALSE)_", "", exps))
for (cb in combos) {
  eT <- paste0("distprod_TRUE_", cb); eF <- paste0("distprod_FALSE_", cb)
  if (!all(c(eT, eF) %in% exps)) next
  both <- intersect(sites[exp == eT], sites[exp == eF])
  if (length(both) == 0) next

  pT04 <- pF04 <- pTfl <- pFfl <- rep(NA, length(both))
  flag04 <- dextract[both, paste0("t", yr.dist)] > 0
  for (m in seq_along(both)) {
    s  <- both[m]
    rT <- readRDS(file.path(indir, sprintf("site_%d_%s.rds", s, eT)))
    rF <- readRDS(file.path(indir, sprintf("site_%d_%s.rds", s, eF)))
    fyrs <- dtime[dextract[s, ] > 0]                 # this site's flagged calendar years
    fyrs <- which(rT$year %in% fyrs)                 # ... as DA timesteps
    pT04[m] <- rT$pDist[rT$year == yr.dist]; pF04[m] <- rF$pDist[rF$year == yr.dist]
    if (length(fyrs) > 0) {
      pTfl[m] <- mean(rT$pDist[fyrs], na.rm = TRUE)
      pFfl[m] <- mean(rF$pDist[fyrs], na.rm = TRUE)
    }
  }

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  plot(pF04, pT04, col = ifelse(flag04, "red", "grey50"), pch = 19,
       xlim = c(0, 1), ylim = c(0, 1),
       xlab = "pDist without product", ylab = "pDist with product",
       main = sprintf("%s @ %d", cb, yr.dist))
  abline(0, 1, lty = 2)
  legend("topleft", legend = c("flagged in 2004", "not flagged in 2004"),
         col = c("red", "grey50"), pch = 19, bty = "n", cex = 0.8)
  plot(pFfl, pTfl, col = "darkgreen", pch = 19, xlim = c(0, 1), ylim = c(0, 1),
       xlab = "mean pDist without product", ylab = "mean pDist with product",
       main = sprintf("%s, over flagged years", cb))
  abline(0, 1, lty = 2)
}
dev.off()
```
