---
title: "OR_joint_update_analysis"
output: html_document
author: "Jacob Epstein"
---

Postprocessing + analyses for the joint-update runs saved by the experiment grid in
`OR_joint_update2.Rmd`. Each run is the `hist` list returned by `ORJointUpdate`, saved
to `<indir>/site_<idx>_distprod_<TF>_<source>_<model>_tau<prior>.rds`, with per-timestep
vectors

* `mean`, `var`, `pDist`: wood-pool analysis mean / variance and P(disturbed), marginalised
  over the tau posterior and the mixture components;
* `innov`, `nis`: innovation `y - E[y | past]` and normalized innovation squared
  `innov^2 / S`, from the one-step forecast predictive (marginalised over the previous
  tau posterior, i.e. integrating over the process error);
* `S.fc`, `S.q`, `S.R`: the parts of the innovation variance `S = S.fc + S.q + S.R`
  (forecast spread, process error, observation error).

The formulas are written up in `claude-working-docs/or_joint_update2_analysis_plots.pdf`.

## set up

```{r}
B2C <- 0.5
load("extract.h03v04.RData")                 # bextract, dextract, points
dtime <- 1985:2012
colnames(dextract) <- paste0("t", dtime)
dextract[is.na(dextract)] <- 0
Yw <- bextract * B2C * 1000 * 1e-4            # nsite x nyears observed

indir <- "experiments/testRun"
t04 <- which(dtime == 2004)   # disturbance-assessment timestep (2004)
tlo <- ncol(Yw)               # last-observation timestep

## turn an experiment label (e.g. "distprod_TRUE_GEDI_het_taugamma") into a readable title
## (e.g. "Disturbance Product, Heteroskedastic R (GEDI magnitude)").
pretty.exp <- function(e) {
  p    <- strsplit(e, "_", fixed = TRUE)[[1]]   # distprod, TRUE/FALSE, source, het/con, tau<prior>
  prod <- if (length(p) >= 2 && p[2] == "TRUE") "Disturbance Product" else "No Disturbance Product"
  src  <- if (length(p) >= 3) p[3] else "?"
  rmod <- if (length(p) >= 4 && p[4] == "het") "Heteroskedastic R" else "Constant R"
  sprintf("%s, %s (%s magnitude)", prod, rmod, src)
}
```

## postprocessing

```{r}
## Detect the joint-update run files and parse site index + experiment label
## (experiment = distprod + Rparams source + model + tau prior).
runs  <- dir(indir, pattern = "^site_.*\\.rds$")
parts <- strsplit(sub("\\.rds$", "", runs), "_", fixed = TRUE)
sites <- sapply(parts, function(x) as.numeric(x[2]))
exp   <- sapply(parts, function(x) paste(x[3:length(x)], collapse = "_"))
exps  <- sort(unique(exp))
print(exps)

nsite <- nrow(dextract)
elist <- list()
for (i in seq_along(exps)) {
  esel <- which(exp == exps[i])
  pdist04 <- nisbar <- nis.n <- Sfc <- Sq <- SR <- innov2bar <- rep(NA, nsite)
  for (j in esel) {
    k <- sites[j]
    if (is.na(k) || k < 1 || k > nsite) next     # skip runs from a different extract
    res <- readRDS(file.path(indir, runs[j]))

    ## detection: posterior P(disturbed) in 2004
    pdist04[k]   <- res$pDist[t04]

    ## mean normalized innovation squared over the observed timesteps
    nisbar[k]    <- mean(res$nis, na.rm = TRUE)
    nis.n[k]     <- sum(is.finite(res$nis))

    ## per-site mean innovation-variance components (S = S.fc + S.q + S.R)
    Sfc[k]       <- mean(res$S.fc, na.rm = TRUE)
    Sq[k]        <- mean(res$S.q,  na.rm = TRUE)
    SR[k]        <- mean(res$S.R,  na.rm = TRUE)

    ## mean squared innovation (numerator of NIS)
    innov2bar[k] <- mean(res$innov^2, na.rm = TRUE)
  }
  elist[[exps[i]]] <- list(pdist04 = pdist04, nisbar = nisbar, nis.n = nis.n,
                           Sfc = Sfc, Sq = Sq, SR = SR, innov2bar = innov2bar)
}

## matching observations
obs.pre   <- Yw[, t04 - 1]                 # 2003 AGB
obs.dist  <- Yw[, t04]                     # 2004 AGB
obs.tlo   <- Yw[, tlo]                     # AGB at the last observation
distabs   <- obs.pre - obs.dist            # absolute AGB change 2003 -> 2004 (loss > 0)
obs.class <- dextract[, "t2004"] / 10      # 2004 disturbance-product label
save(elist, obs.pre, obs.dist, obs.tlo, obs.class, distabs,
     file = file.path(indir, "elist.RData"))
```

## analyses

Summary PDF: one page per plot, every experiment drawn as its own panel with shared
axes. Points are coloured by the 2004 disturbance-product label.

```{r}
classnames  <- c("other", "cut", "road", "fire", "biotic")          # codes 10..50
class.cols  <- c("grey60", "black", "red", "green3", "blue", "cyan")  # index = class + 1
pt.col      <- class.cols[obs.class + 1]
class.vals  <- sort(unique(obs.class[obs.class != 0]))
legend.labels <- c("undisturbed", classnames[class.vals])
legend.cols   <- c(class.cols[1], class.cols[class.vals + 1])
S.cols   <- c("grey60", "orange", "steelblue")
S.labels <- c("S.fc (forecast spread)", "S.q (process error)", "S.R (obs error)")

nexp <- length(exps)
grid <- c(2, ceiling(nexp / 2))
has  <- lapply(elist, function(e) which(is.finite(e$nisbar)))   # sites run per experiment

## 95% interval for the mean of N per-step NIS values: chi-sq_N / N with N = number of
## observed timesteps (exact only for a Gaussian predictive, see the writeup)
n.obs  <- max(unlist(lapply(elist, function(e) e$nis.n)), na.rm = TRUE)
nis.ci <- qchisq(c(0.025, 0.975), df = n.obs) / n.obs

## beta regression of P(detect) on disturbance size, fitted only with enough sites
betatrim <- function(y) { n <- sum(!is.na(y)); (y * (n - 1) + 0.5) / n }
safe.betareg <- function(...) {
  tryCatch(betareg::betareg(...),
           error = function(e) { warning("betareg failed: ", conditionMessage(e)); NULL })
}
min.fit.sites <- 20
stats <- list()
for (i in seq_along(exps)) {
  p <- elist[[i]]$pdist04
  fit <- NULL
  if (sum(is.finite(p)) >= min.fit.sites) {
    p2  <- betatrim(p)
    fit <- safe.betareg(p2 ~ distabs, na.action = na.omit, link = "logit")
  }
  stats[[exps[i]]] <- list(dist.beta.abs = fit)
}

pdf(file.path(indir, "exps_summary.pdf"), width = 14, height = 7)

## ---- Page 1: mean NIS vs final-year biomass ----
all.nis  <- unlist(lapply(elist, function(e) e$nisbar))
nis.ylim <- range(c(all.nis, nis.ci), na.rm = TRUE) * c(0.8, 1.25)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e <- elist[[i]]
  plot(obs.tlo, e$nisbar, col = pt.col, pch = 19, log = "y",
       xlim = range(obs.tlo[has[[i]]]), ylim = nis.ylim, cex.main = 0.8,
       xlab = "final-year AGB (kgC m-2)", ylab = "mean NIS", main = pretty.exp(exps[i]))
  rect(par("usr")[1], nis.ci[1], par("usr")[2], nis.ci[2], col = adjustcolor("grey", 0.25), border = NA)
  abline(h = 1, lty = 2)
  if (i == 1)
    legend("topright", legend = legend.labels, col = legend.cols,
           pch = 19, bty = "n", cex = 0.8, title = "2004 product label")
}
mtext("mean NIS vs final-year biomass (dashed = 1, grey = 95% interval of chi-sq_N / N)",
      outer = TRUE, cex = 1.2)

## ---- Page 2: mean NIS, all experiments on one axis ----
par(mfrow = c(1, 1), mar = c(5, 24, 3, 2), oma = c(0, 0, 0, 0))
plot(NA, xlim = nis.ylim, ylim = c(0.5, nexp + 0.5), log = "x", yaxt = "n",
     xlab = "mean NIS (per site)", ylab = "", main = "Mean NIS by experiment")
axis(2, at = seq_len(nexp), labels = sapply(exps, pretty.exp), las = 1, cex.axis = 0.8)
rect(nis.ci[1], 0, nis.ci[2], nexp + 1, col = adjustcolor("grey", 0.25), border = NA)
abline(v = 1, lty = 2)
for (i in seq_along(exps)) {
  s <- has[[i]]
  points(elist[[i]]$nisbar[s], rep(i, length(s)) + seq(-0.15, 0.15, length = length(s)),
         col = pt.col[s], pch = 19)
}
legend("topright", legend = legend.labels, col = legend.cols, pch = 19, bty = "n",
       cex = 0.8, title = "2004 product label")

## ---- Page 3: disturbance detection vs absolute disturbance size ----
all.sites    <- sort(unique(unlist(has)))
distabs.xlim <- range(distabs[all.sites], na.rm = TRUE)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  plot(distabs, elist[[i]]$pdist04, col = pt.col, pch = 19,
       xlim = distabs.xlim, ylim = c(0, 1), cex.main = 0.8,
       xlab = "AGB loss 2003 -> 2004 (kgC m-2)", ylab = "P(disturbed) in 2004",
       main = pretty.exp(exps[i]))
  abline(v = 0, col = "grey80")
  fit <- stats[[i]]$dist.beta.abs
  if (!is.null(fit)) {
    xx <- seq(distabs.xlim[1], distabs.xlim[2], length = 200)
    nd <- data.frame(distabs = xx)
    polygon(c(xx, rev(xx)),
            c(predict(fit, nd, type = "quantile", at = 0.025),
              rev(predict(fit, nd, type = "quantile", at = 0.975))),
            col = adjustcolor("blue", 0.18), border = NA)
    lines(xx, predict(fit, nd, type = "response"), col = "blue", lwd = 2)
  }
  if (i == 1)
    legend("topleft", legend = legend.labels, col = legend.cols,
           pch = 19, bty = "n", cex = 0.8, title = "2004 product label")
}
mtext(sprintf("disturbance detection vs absolute disturbance size%s",
              if (any(!sapply(stats, function(s) is.null(s$dist.beta.abs))))
                " (blue = beta regression mean + 95% band)" else ""),
      outer = TRUE, cex = 1.2)

## ---- Page 4: innovation-variance decomposition per site ----
## stacked bars of the per-site mean S = S.fc + S.q + S.R (all sites if <= 30, else the
## 30 with the largest S), each bar labelled with the site index.
max.bars <- 30
all.S    <- unlist(lapply(elist, function(e) e$Sfc + e$Sq + e$SR))
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e  <- elist[[i]]
  ok <- which(is.finite(e$Sfc + e$Sq + e$SR))
  if (length(ok) == 0) { plot.new(); title(main = pretty.exp(exps[i])); next }
  ord <- tail(ok[order(e$Sfc[ok] + e$Sq[ok] + e$SR[ok])], max.bars)
  barplot(rbind(e$Sfc[ord], e$Sq[ord], e$SR[ord]), names.arg = ord, las = 2,
          cex.names = 0.7, ylim = c(0, max(all.S, na.rm = TRUE)), col = S.cols,
          xlab = "site (ordered by S)", ylab = "variance (kgC m-2)^2",
          main = pretty.exp(exps[i]), cex.main = 0.8)
  if (i == 1) legend("topleft", bty = "n", cex = 0.8, legend = S.labels, fill = S.cols)
}
mtext("mean innovation variance S = S.fc + S.q + S.R per site", outer = TRUE, cex = 1.2)

## ---- Page 5: mean innovation variance per experiment (absolute and share) ----
M <- sapply(elist, function(e) c(S.fc = mean(e$Sfc, na.rm = TRUE),
                                 S.q  = mean(e$Sq,  na.rm = TRUE),
                                 S.R  = mean(e$SR,  na.rm = TRUE)))   # 3 x nexp
colnames(M) <- sapply(exps, pretty.exp)
ord <- order(colSums(M))
par(mfrow = c(1, 2), mar = c(5, 22, 3, 1), oma = c(0, 0, 0, 0))
barplot(M[, ord, drop = FALSE], horiz = TRUE, las = 1, cex.names = 0.75, col = S.cols,
        xlab = "mean innovation variance (kgC m-2)^2", main = "Mean S by experiment")
legend("bottomright", bty = "n", cex = 0.8, legend = S.labels, fill = S.cols)
par(mar = c(5, 1, 3, 2))
barplot(sweep(M, 2, colSums(M), "/")[, ord, drop = FALSE], horiz = TRUE, names.arg = rep("", nexp),
        col = S.cols, xlab = "share of mean S", main = "Share of each component")

## ---- Page 6: mean squared innovation vs mean S per site ----
## NIS = innov^2 / S, so bars above the points mean NIS > 1 (forecast too confident
## or biased), bars below mean NIS < 1 (predictive too wide).
i2.max <- max(unlist(lapply(elist, function(e) c(e$innov2bar, e$Sfc + e$Sq + e$SR))), na.rm = TRUE)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e    <- elist[[i]]
  Stot <- e$Sfc + e$Sq + e$SR
  ok   <- which(is.finite(e$innov2bar) & is.finite(Stot))
  if (length(ok) == 0) { plot.new(); title(main = pretty.exp(exps[i])); next }
  ord <- tail(ok[order(e$innov2bar[ok])], max.bars)
  bp  <- barplot(e$innov2bar[ord], names.arg = ord, las = 2, cex.names = 0.7,
                 ylim = c(0, i2.max), col = "tomato", xlab = "site",
                 ylab = "(kgC m-2)^2", main = pretty.exp(exps[i]), cex.main = 0.8)
  points(bp, Stot[ord], pch = 19)
  if (i == 1)
    legend("topleft", bty = "n", cex = 0.8, pch = c(15, 19), col = c("tomato", "black"),
           legend = c("mean innov^2 (NIS numerator)", "mean S (NIS denominator)"))
}
mtext("mean squared innovation vs mean innovation variance", outer = TRUE, cex = 1.2)

## ---- Page 7: disturbance size at 50% probability of detection (needs the beta fits) ----
## logit-link mean: P = 0.5 where the linear predictor is 0, i.e. distabs = -intercept / slope
x50 <- sapply(stats, function(s) {
  fit <- s$dist.beta.abs
  if (is.null(fit)) return(NA_real_)
  cf <- coef(fit, model = "mean")
  if (length(cf) < 2 || !is.finite(cf[2]) || cf[2] == 0) return(NA_real_)
  as.numeric(-cf[1] / cf[2])
})
if (any(is.finite(x50))) {
  par(mfrow = c(1, 1), mar = c(1, 1, 4, 1))
  plot.new()
  title("AGB loss (kgC m-2) at 50% probability of detection")
  yy <- 0.9 - seq_len(nexp) * 0.08
  text(0.02, 0.9, "Experiment", adj = 0, font = 2)
  text(0.75, 0.9, "AGB loss at P = 0.5", adj = 0, font = 2)
  text(0.02, yy, sapply(exps, pretty.exp), adj = 0, cex = 0.85)
  text(0.75, yy, ifelse(is.finite(x50), formatC(x50, format = "f", digits = 2), "NA"), adj = 0, cex = 0.85)
}
dev.off()
save(stats, file = file.path(indir, "stats.RData"))
```

## disturbance product: where does pDist change?

Compares the runs with and without the disturbance product (same R model and tau prior)
(a) at the assessment year `t04` and (b) averaged over each site's own flagged years.

```{r}
pdf(file.path(indir, "distprod_TRUE_vs_FALSE.pdf"), width = 10, height = 5)
combos <- unique(sub("^distprod_(TRUE|FALSE)_", "", exps))
for (cb in combos) {
  eT <- paste0("distprod_TRUE_", cb); eF <- paste0("distprod_FALSE_", cb)
  if (!all(c(eT, eF) %in% exps)) next
  both <- intersect(sites[exp == eT], sites[exp == eF])
  if (length(both) == 0) next

  pT04 <- pF04 <- pTfl <- pFfl <- rep(NA, length(both))
  flag04 <- dextract[both, t04] > 0
  for (m in seq_along(both)) {
    s  <- both[m]
    rT <- readRDS(file.path(indir, sprintf("site_%d_%s.rds", s, eT)))
    rF <- readRDS(file.path(indir, sprintf("site_%d_%s.rds", s, eF)))
    fyrs <- which(dextract[s, ] > 0)
    fyrs <- fyrs[fyrs <= length(rT$pDist)]
    pT04[m] <- rT$pDist[t04]; pF04[m] <- rF$pDist[t04]
    if (length(fyrs) > 0) {
      pTfl[m] <- mean(rT$pDist[fyrs], na.rm = TRUE)
      pFfl[m] <- mean(rF$pDist[fyrs], na.rm = TRUE)
    }
  }

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  plot(pF04, pT04, col = ifelse(flag04, "red", "grey50"), pch = 19,
       xlim = c(0, 1), ylim = c(0, 1),
       xlab = "pDist without product", ylab = "pDist with product",
       main = sprintf("%s @ %d", cb, dtime[t04]))
  abline(0, 1, lty = 2)
  legend("topleft", legend = c("flagged in 2004", "not flagged in 2004"),
         col = c("red", "grey50"), pch = 19, bty = "n", cex = 0.8)
  plot(pFfl, pTfl, col = "darkgreen", pch = 19, xlim = c(0, 1), ylim = c(0, 1),
       xlab = "mean pDist without product", ylab = "mean pDist with product",
       main = sprintf("%s, over flagged years", cb))
  abline(0, 1, lty = 2)
}
dev.off()
```

## when were sites flagged as disturbed (by type and year)?

One barplot per disturbance type: the number of sites flagged in each year, with the
assessment year `t04` in red.

```{r}
dist.codes <- c("10" = "other", "20" = "cut", "30" = "road", "40" = "fire", "50" = "biotic")
codes.present <- names(dist.codes)[sapply(names(dist.codes),
                   function(cc) any(dextract == as.numeric(cc)))]
NTy <- ncol(dextract)

par(mfrow = c(length(codes.present), 1), mar = c(3, 4, 2, 1), oma = c(2, 0, 2, 0))
for (cc in codes.present) {
  cnt  <- colSums(dextract == as.numeric(cc))
  bcol <- rep("steelblue", NTy); bcol[t04] <- "red"
  barplot(cnt, col = bcol, border = NA, names.arg = dtime, las = 2,
          cex.names = 0.6, ylab = "# sites flagged",
          main = sprintf("%s (code %s) - %d flags total", dist.codes[cc], cc, sum(cnt)))
}
mtext("disturbance flags by type and year (red bar = 2004)", outer = TRUE, cex = 1.1)
```
