---
title: "OR_joint_update_analysis"
output: html_document
author: "Jacob Epstein"
---

Postprocessing + analyses for the joint-update runs saved by the experiment grid in
`OR_joint_update2.Rmd`. Each run is the `hist` list returned by `ORJointUpdate`, saved
to `<indir>/site_<idx>_distprod_<TF>_<source>_<model>_tau<prior>.rds`, with per-timestep

* `year`, `obs`: calendar year of each DA timestep and the assimilated observation;
* `mean`, `var`, `ci95`: wood-pool analysis mean, variance and 95% interval (2.5% / 97.5%
  quantiles of the analysis mixture), marginalised over the tau posterior and the
  mixture components;
* `pDist`: posterior probability of disturbance;
* `pit`: probability integral transform of each observation under its one-step-ahead
  predictive (before that observation is assimilated, integrating over process error).

The formulas are written up in `claude-working-docs/or_joint_update2_analysis_plots.pdf`.

## set up

```{r}
B2C <- 0.5
load("extract.h03v04.RData")                 # bextract, dextract, points
## The products cover different years, so every lookup goes through calendar year:
##   dextract (disturbance product): 1985:2012
##   bextract (biomass / Yw):        1990:2017
##   DA timesteps:                   stored per run in hist$year (1990:2017)
dtime <- 1985:2012
ctime <- 1990:2017
colnames(dextract) <- paste0("t", dtime)
dextract[is.na(dextract)] <- 0
Yw <- bextract * B2C * 1000 * 1e-4            # nsite x nyears observed
colnames(Yw) <- paste0("t", ctime)

indir   <- "experiments/testRun"
yr.dist <- 2004               # disturbance-assessment year

## turn an experiment label (e.g. "distprod_TRUE_GEDI_het_taugamma") into a readable title
## (e.g. "Disturbance Product, Heteroskedastic R (GEDI magnitude)").
pretty.exp <- function(e) {
  p    <- strsplit(e, "_", fixed = TRUE)[[1]]   # distprod, TRUE/FALSE, source, het/con, tau<prior>
  prod <- if (length(p) >= 2 && p[2] == "TRUE") "Disturbance Product" else "No Disturbance Product"
  src  <- if (length(p) >= 3) p[3] else "?"
  rmod <- if (length(p) >= 4 && p[4] == "het") "Heteroskedastic R" else "Constant R"
  sprintf("%s, %s (%s magnitude)", prod, rmod, src)
}
```

## postprocessing

```{r}
## Detect the joint-update run files and parse site index + experiment label
## (experiment = distprod + Rparams source + model + tau prior).
runs  <- dir(indir, pattern = "^site_.*\\.rds$")
parts <- strsplit(sub("\\.rds$", "", runs), "_", fixed = TRUE)
sites <- sapply(parts, function(x) as.numeric(x[2]))
exp   <- sapply(parts, function(x) paste(x[3:length(x)], collapse = "_"))
exps  <- sort(unique(exp))
print(exps)

nsite <- nrow(dextract)
elist <- list()
for (i in seq_along(exps)) {
  esel <- which(exp == exps[i])
  pdist04 <- div <- pitT <- obsT <- rep(NA, nsite)
  pit.all <- c()
  for (j in esel) {
    k <- sites[j]
    if (is.na(k) || k < 1 || k > nsite) next     # skip runs from a different extract
    res <- readRDS(file.path(indir, runs[j]))
    if (is.null(res$year)) stop(runs[j], ": no $year axis -- stale run from before the 1990 alignment fix; rerun it")
    if (is.null(res$ci95)) stop(runs[j], ": no $ci95 -- run from before the filter-divergence change; rerun it")

    ## detection: posterior P(disturbed) in 2004
    pdist04[k] <- res$pDist[res$year == yr.dist]

    ## filter divergence at the last observed timestep T:
    ## (y_T - analysis mean) / (half the width of the 95% analysis interval)
    tT      <- max(which(!is.na(res$obs)))
    obsT[k] <- res$obs[tT]
    div[k]  <- (res$obs[tT] - res$mean[tT]) / ((res$ci95[tT, "hi"] - res$ci95[tT, "lo"]) / 2)

    ## PIT of the last observation under its one-step-ahead predictive, and all PITs
    pitT[k] <- res$pit[tT]
    pit.all <- c(pit.all, res$pit[!is.na(res$pit)])
  }
  elist[[exps[i]]] <- list(pdist04 = pdist04, div = div, pitT = pitT, obsT = obsT,
                           pit.all = pit.all)
}

## matching observations
obs.pre   <- Yw[, paste0("t", yr.dist - 1)]   # 2003 AGB
obs.dist  <- Yw[, paste0("t", yr.dist)]       # 2004 AGB
distabs   <- obs.pre - obs.dist               # absolute AGB change 2003 -> 2004 (loss > 0)
obs.class <- dextract[, paste0("t", yr.dist)] / 10   # 2004 disturbance-product label
save(elist, obs.pre, obs.dist, obs.class, distabs,
     file = file.path(indir, "elist.RData"))
```

## analyses

Summary PDF: one page per plot, every experiment drawn as its own panel with shared
axes. Points are coloured by the 2004 disturbance-product label.

```{r}
classnames  <- c("other", "cut", "road", "fire", "biotic")          # codes 10..50
class.cols  <- c("grey60", "black", "red", "green3", "blue", "cyan")  # index = class + 1
pt.col      <- class.cols[obs.class + 1]
class.vals  <- sort(unique(obs.class[obs.class != 0]))
legend.labels <- c("undisturbed", classnames[class.vals])
legend.cols   <- c(class.cols[1], class.cols[class.vals + 1])

nexp <- length(exps)
grid <- c(2, ceiling(nexp / 2))
has  <- lapply(elist, function(e) which(is.finite(e$div)))   # sites run per experiment
all.sites <- sort(unique(unlist(has)))
pct  <- function(x) sprintf("%.0f%%", 100 * mean(x, na.rm = TRUE))

## beta regression of P(detect) on disturbance size, fitted only with enough sites
betatrim <- function(y) { n <- sum(!is.na(y)); (y * (n - 1) + 0.5) / n }
safe.betareg <- function(...) {
  tryCatch(betareg::betareg(...),
           error = function(e) { warning("betareg failed: ", conditionMessage(e)); NULL })
}
min.fit.sites <- 20
stats <- list()
for (i in seq_along(exps)) {
  p <- elist[[i]]$pdist04
  fit <- NULL
  if (sum(is.finite(p)) >= min.fit.sites) {
    p2  <- betatrim(p)
    fit <- safe.betareg(p2 ~ distabs, na.action = na.omit, link = "logit")
  }
  stats[[exps[i]]] <- list(dist.beta.abs = fit)
}

pdf(file.path(indir, "exps_summary.pdf"), width = 14, height = 7)
obsT.xlim <- range(unlist(lapply(elist, function(e) e$obsT)), na.rm = TRUE)

## ---- Page 1: filter divergence per site ----
## (y_T - analysis mean_T) / half-width of the 95% analysis interval; inside [-1, 1]
## means the final observation falls inside the interval.
div.ylim <- range(c(-1.5, 1.5, unlist(lapply(elist, function(e) e$div))), na.rm = TRUE)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e <- elist[[i]]
  plot(e$obsT, e$div, col = pt.col, pch = 19, xlim = obsT.xlim, ylim = div.ylim,
       xlab = "final observed AGB (kgC m-2)", ylab = "filter divergence", cex.main = 0.8,
       main = sprintf("%s\n%s of sites in [-1, 1]", pretty.exp(exps[i]), pct(abs(e$div) <= 1)))
  rect(par("usr")[1], -1, par("usr")[2], 1, col = adjustcolor("grey", 0.25), border = NA)
  abline(h = 0, lty = 2)
  if (i == 1)
    legend("topright", legend = legend.labels, col = legend.cols,
           pch = 19, bty = "n", cex = 0.8, title = "2004 product label")
}
mtext("filter divergence at the final observation (grey = [-1, 1])", outer = TRUE, cex = 1.2)

## ---- Page 2: centred PIT of the final observation per site ----
## 2u - 1 with u = P(y' <= y_T | y_{1:T-1}); uniform on [-1, 1] if the predictive is
## calibrated, so 95% of sites should fall in [-0.95, 0.95].
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e <- elist[[i]]; cp <- 2 * e$pitT - 1
  plot(e$obsT, cp, col = pt.col, pch = 19, xlim = obsT.xlim, ylim = c(-1, 1),
       xlab = "final observed AGB (kgC m-2)", ylab = "centred PIT, 2u - 1", cex.main = 0.8,
       main = sprintf("%s\n%s of sites in [-0.95, 0.95]", pretty.exp(exps[i]), pct(abs(cp) <= 0.95)))
  rect(par("usr")[1], -0.95, par("usr")[2], 0.95, col = adjustcolor("grey", 0.25), border = NA)
  abline(h = 0, lty = 2)
  if (i == 1)
    legend("topright", legend = legend.labels, col = legend.cols,
           pch = 19, bty = "n", cex = 0.8, title = "2004 product label")
}
mtext("one-step-ahead PIT of the final observation (grey = central 95%)", outer = TRUE, cex = 1.2)

## ---- Page 3: PIT histogram over all sites and years ----
## flat at density 1 if calibrated; U-shape = predictive too narrow, hump = too wide,
## slope = biased.
pit.br <- seq(0, 1, by = 0.05)
pit.ymax <- max(sapply(elist, function(e) max(hist(e$pit.all, breaks = pit.br, plot = FALSE)$density)))
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  e <- elist[[i]]
  hist(e$pit.all, breaks = pit.br, freq = FALSE, col = "grey80", border = "white",
       ylim = c(0, pit.ymax), xlab = "PIT u", cex.main = 0.8,
       main = sprintf("%s\n%d site-years", pretty.exp(exps[i]), length(e$pit.all)))
  abline(h = 1, lty = 2)
}
mtext("one-step-ahead PIT of every observation, all sites and years (dashed = uniform)",
      outer = TRUE, cex = 1.2)

## ---- Page 4: disturbance detection vs absolute disturbance size ----
distabs.xlim <- range(distabs[all.sites], na.rm = TRUE)
par(mfrow = grid, mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(exps)) {
  plot(distabs, elist[[i]]$pdist04, col = pt.col, pch = 19,
       xlim = distabs.xlim, ylim = c(0, 1), cex.main = 0.8,
       xlab = "AGB loss 2003 -> 2004 (kgC m-2)", ylab = "P(disturbed) in 2004",
       main = pretty.exp(exps[i]))
  abline(v = 0, col = "grey80")
  fit <- stats[[i]]$dist.beta.abs
  if (!is.null(fit)) {
    xx <- seq(distabs.xlim[1], distabs.xlim[2], length = 200)
    nd <- data.frame(distabs = xx)
    polygon(c(xx, rev(xx)),
            c(predict(fit, nd, type = "quantile", at = 0.025),
              rev(predict(fit, nd, type = "quantile", at = 0.975))),
            col = adjustcolor("blue", 0.18), border = NA)
    lines(xx, predict(fit, nd, type = "response"), col = "blue", lwd = 2)
  }
  if (i == 1)
    legend("topleft", legend = legend.labels, col = legend.cols,
           pch = 19, bty = "n", cex = 0.8, title = "2004 product label")
}
mtext(sprintf("disturbance detection vs absolute disturbance size%s",
              if (any(!sapply(stats, function(s) is.null(s$dist.beta.abs))))
                " (blue = beta regression mean + 95% band)" else ""),
      outer = TRUE, cex = 1.2)

## ---- Page 5: summary table ----
## logit-link mean: P = 0.5 where the linear predictor is 0, i.e. distabs = -intercept / slope
x50 <- sapply(stats, function(s) {
  fit <- s$dist.beta.abs
  if (is.null(fit)) return(NA_real_)
  cf <- coef(fit, model = "mean")
  if (length(cf) < 2 || !is.finite(cf[2]) || cf[2] == 0) return(NA_real_)
  as.numeric(-cf[1] / cf[2])
})
tab <- data.frame(
  experiment = sapply(exps, pretty.exp),
  div.in     = sapply(elist, function(e) pct(abs(e$div) <= 1)),
  div.med    = sapply(elist, function(e) sprintf("%.2f", median(abs(e$div), na.rm = TRUE))),
  pit.in     = sapply(elist, function(e) pct(abs(2 * e$pitT - 1) <= 0.95)),
  pit.all.in = sapply(elist, function(e) pct(abs(2 * e$pit.all - 1) <= 0.95)),
  x50        = ifelse(is.finite(x50), sprintf("%.2f", x50), "--"))
heads <- c("Experiment", "|div| <= 1", "median |div|", "final PIT\nin 95%",
           "all PITs\nin 95%", "loss at\nP = 0.5")
par(mfrow = c(1, 1), mar = c(1, 1, 4, 1))
plot.new()
title("Summary by experiment")
xs <- c(0.01, 0.50, 0.61, 0.72, 0.83, 0.93)
yy <- 0.86 - seq_len(nexp) * 0.08
text(xs, 0.92, heads, adj = c(0, 0.5), font = 2, cex = 0.8)
for (c in seq_along(xs)) text(xs[c], yy, tab[[c]], adj = 0, cex = 0.8)
segments(0, c(0.97, 0.87), 1, c(0.97, 0.87), col = "grey55")
dev.off()
save(stats, tab, file = file.path(indir, "stats.RData"))
```

## disturbance product: where does pDist change?

Compares the runs with and without the disturbance product (same R model and tau prior)
(a) at the assessment year `yr.dist` (2004) and (b) averaged over each site's own flagged years.

```{r}
pdf(file.path(indir, "distprod_TRUE_vs_FALSE.pdf"), width = 10, height = 5)
combos <- unique(sub("^distprod_(TRUE|FALSE)_", "", exps))
for (cb in combos) {
  eT <- paste0("distprod_TRUE_", cb); eF <- paste0("distprod_FALSE_", cb)
  if (!all(c(eT, eF) %in% exps)) next
  both <- intersect(sites[exp == eT], sites[exp == eF])
  if (length(both) == 0) next

  pT04 <- pF04 <- pTfl <- pFfl <- rep(NA, length(both))
  flag04 <- dextract[both, paste0("t", yr.dist)] > 0
  for (m in seq_along(both)) {
    s  <- both[m]
    rT <- readRDS(file.path(indir, sprintf("site_%d_%s.rds", s, eT)))
    rF <- readRDS(file.path(indir, sprintf("site_%d_%s.rds", s, eF)))
    fyrs <- dtime[dextract[s, ] > 0]                 # this site's flagged calendar years
    fyrs <- which(rT$year %in% fyrs)                 # ... as DA timesteps
    pT04[m] <- rT$pDist[rT$year == yr.dist]; pF04[m] <- rF$pDist[rF$year == yr.dist]
    if (length(fyrs) > 0) {
      pTfl[m] <- mean(rT$pDist[fyrs], na.rm = TRUE)
      pFfl[m] <- mean(rF$pDist[fyrs], na.rm = TRUE)
    }
  }

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  plot(pF04, pT04, col = ifelse(flag04, "red", "grey50"), pch = 19,
       xlim = c(0, 1), ylim = c(0, 1),
       xlab = "pDist without product", ylab = "pDist with product",
       main = sprintf("%s @ %d", cb, yr.dist))
  abline(0, 1, lty = 2)
  legend("topleft", legend = c("flagged in 2004", "not flagged in 2004"),
         col = c("red", "grey50"), pch = 19, bty = "n", cex = 0.8)
  plot(pFfl, pTfl, col = "darkgreen", pch = 19, xlim = c(0, 1), ylim = c(0, 1),
       xlab = "mean pDist without product", ylab = "mean pDist with product",
       main = sprintf("%s, over flagged years", cb))
  abline(0, 1, lty = 2)
}
dev.off()
```

## when were sites flagged as disturbed (by type and year)?

One barplot per disturbance type: the number of sites flagged in each year, with the
assessment year 2004 in red.

```{r}
dist.codes <- c("10" = "other", "20" = "cut", "30" = "road", "40" = "fire", "50" = "biotic")
codes.present <- names(dist.codes)[sapply(names(dist.codes),
                   function(cc) any(dextract == as.numeric(cc)))]
NTy <- ncol(dextract)

par(mfrow = c(length(codes.present), 1), mar = c(3, 4, 2, 1), oma = c(2, 0, 2, 0))
for (cc in codes.present) {
  cnt  <- colSums(dextract == as.numeric(cc))
  bcol <- rep("steelblue", NTy); bcol[dtime == yr.dist] <- "red"
  barplot(cnt, col = bcol, border = NA, names.arg = dtime, las = 2,
          cex.names = 0.6, ylab = "# sites flagged",
          main = sprintf("%s (code %s) - %d flags total", dist.codes[cc], cc, sum(cnt)))
}
mtext("disturbance flags by type and year (red bar = 2004)", outer = TRUE, cex = 1.1)
```
