---
title: "OR_joint_update"
output: html_document
author: "Jacob Epstein"
date: "5/30/26"
---

```{r}
library(statmod)
library(pracma)
library(BayesianTools)       # VSEM forest model
source("StatsAuxiliary.R")   # normal_quantile(p, mean, variance)
source("disturbance.R")      # disturbance.p() forecast perturbation
## Load disturbance rate, state&transition, disturbed biomass stats
load("STprior.RData")
load("extract.h03v04.RData")
load("dis.postprocessed.RData")
dtime <- 1985:2012
colnames(dextract) <- paste0("t",dtime)
dextract[is.na(dextract)] <- 0
d.prior["UN"] = d.prior["UN"] + d.prior[6]                         ## change at runtime (removes <NA>)
d.prior = d.prior[-6]

## Load Calibration
load("fit.RData")
Qsoil = Qs
```

```{r}
latlon = points@coords # lat / lon of each site in dextract 
metFile = "daymet.RData"
if(file.exists(metFile)){
  load(metFile)
} else {
  dm = list()
  for(i in 1:nrow(latlon)){
    dm[[i]] = daymetr::download_daymet(site="h03v04",latlon[i,2],latlon[i,1],start=1985,end=2017,internal=TRUE,simplify=TRUE)
  }
  save(dm,file=metFile)
}

## extract Shortwave and convert to daily PAR
SW2PAR = 0.45
srad = subset(dm[[1]],measurement == "srad..W.m.2.")   
metTime <- as.POSIXct(paste(srad$year,srad$yday,sep="-"),format="%Y-%j")
rad <- matrix(NA,length(dm),length(metTime)) # [number of sites x number of measurements]
print(dim(rad))
for(i in seq_along(dm)){
  srad = subset(dm[[i]],measurement == "srad..W.m.2.")   
  rad[i,] = srad$value*1e-6*86400*SW2PAR  # -> MJ /m2 /day PAR (convert) # QUESTION -- should the changing day length, from the daymet day length values, be used (instead of a fixed 24-hr day length)?
}
plot(metTime,rad[1,],type='l')
## plot(date,PAR,col=2) ## calibration data
```

```{r}
dist <- list()
dist_name <- c(
  "0"  = "none",
  "10" = "other",
  "20" = "cut",
  "30" = "road",
  "40" = "fire",
  "50" = "biotic"
)

dist.forecastparams <- list()

count <- 1
for (dclass in names(dist.priors)) {
  dist.name <- as.character(dist_name[dclass])
  nclust <- length(dist.priors[[dclass]]$means)
  
  for (i in 1:nclust) {
    dist.forecastparams[[count]] <- list()
    dist.forecastparams[[count]]$mu0 <- c(0.2, dist.priors[[dclass]]$means[i])
    dist.forecastparams[[count]]$V0 <- diag(c(0.04, dist.priors[[dclass]]$vars[i]))
    dist.forecastparams[[count]]$name <- paste0(dist.name, i)
    count <- count + 1
  }
}
```

```{r}
## BIOMASS
B2C = 0.5 ##biomass to carbon ratio
Yw <- bextract*B2C*1000*1e-4 ## convert Mg/ha (Mg per hectacre)  to  kg C [356 (num ensemble members) x 28 (num years observed)]
Rw <- (84*B2C*1000*1e-4)^2 ## Observation Error: Kennedy et al 2018 Environ. Res. Lett. 13 025004; Fig 2 RMSE
## units: Mg/ha -> kg/m2
print(dim(Yw))
## grab the distribution of AGB for the first time point from all pixels
AGBprior <- Yw[,1]
ne = length(AGBprior) ## number of ensemble members

## spin up
PARlong <- as.vector(t(rad)) # concatenate PAR time series for each site into one vector
spin = VSEM(theta,PARlong) # Question: why run VSEM on the concatenated PAR? What is theta?
thin = seq(nrow(spin)/2,nrow(spin),length=ne)
spinIC = spin[thin,2:4] # select ne samples from the spin-up 
#for(i in 2:4){
#  hist(spin[thin,i],type='l')
#}

## disturbance recovery trajectory
nNBG = 200 ## years
NBG <- array(NA,c(ne,nNBG,3))
for(i in 1:ne){
  param = theta
  param[9:11] <- spinIC[i,]*c(0.01,1,0.01) ## set leaf and wood to 1% (simulate disturbance)
  raw = VSEM(param,PARlong[1:(nNBG*365.25)]) # simulate for nNBG years
  NBG[i,,] = raw[round(seq(365,nNBG*365.25,length=nNBG)),2:4]  ## grab pools at the end of each year
}
  
## for each AGBprior, grab the matching leaf and soil
ICprior <- matrix(NA,ne,3)
age = rep(NA,ne)
for(i in 1:ne){
  age[i] <- which.min((AGBprior[i]-NBG[i,,3])^2) # age = year in the disturbance recover trajectory where the simulated site has similar AGB to the AGB prior information
  ICprior[i,] <- c(NBG[i,age[i],1:2],AGBprior[i]) # use AGBprior for the model ICs
}
hist(age)
colnames(ICprior) <- c("Cv","Cs","CR") 
pairs(ICprior)
```

```{r}

## ---- Site-independent setup (computed once) --------------------------------
## Number of disturbance sub-classes and number of assimilation years.
n.dist = length(dist.forecastparams)
NT = floor(ncol(rad)/365)

## Background disturbance-type probability per forecast component, by matching
## the component's name to one of "other"/"cut"/"fire"/"biotic".
distname.toidx <- list("other"=2, "cut"=3, "fire"=4, "biotic"=5)
dist.prior.probs <- sapply(dist.forecastparams, function(x) {
  name <- x$name
  for (d in names(distname.toidx)) {
    if (grepl(d, name)) {
      return(d.prior[distname.toidx[[d]]])
    }
  }
})

## Diagonal process-error covariance (per-pool SDs) and prior process-precision
## gamma params (a, b) = (shape, rate). Both site-independent.
Q = diag(c(Ql,Qsoil,Qw)^2)
a <- 1/0.1172475
b <- 1
gamInit <- c(a, b)

## Default heteroskedastic Rparams for GEDI (slope, intercept).
Rparams.GEDI.het <- c(0.01331, 0.09761)
Rparams.GEDI.fixed <- c(0.0000000, 0.8848249)
Rparams.Landtrendr.het <- c(0.013310, 3.012645)
Rparams.Landtrendr.fixed <- c(0.000, 3.79986)

## ---- Per-site argument builder ---------------------------------------------
## make_or_args(site, useDistProduct) assembles the site-specific inputs to
## ORJointUpdate: the (n.dist+1) x NT prior probability matrix p.dist (row 1 =
## undisturbed, rows 2.. = disturbance sub-classes), the bundled disturb.params,
## the NT x 3 observation matrix Y (wood-pool obs in column 3), and the PAR
## driver for that site. Relies on the site-independent globals above plus
## d.prior, dist.priors, dist.forecastparams, dextract, Yw and rad.
## tauPrior selects the prior on the process precision tau:
##   "gamma"      : Gamma(shape = 0.1, rate = 2e-4), which spreads mass over every
##                  decade of the grid (mean tau = 500)
##   "loguniform" : p(tau) proportional to 1/tau (flat in log tau)
make_or_args <- function(site, useDistProduct = TRUE, tauPrior = c("gamma", "loguniform")) {
  tauPrior <- match.arg(tauPrior)
  ## Initialize every year with the background prior (same column for all t).
  p.dist <- array(NA, dim = c(n.dist + 1, NT))
  for (t in 1:NT) {
    p.dist[1:(n.dist+1), t] <- as.numeric(c(d.prior[1], unlist(dist.prior.probs)))
  }

  ## If using the disturbance product, override the prior in years where a
  ## disturbance was detected at this site: replace that class's prior with the
  ## user's-accuracy estimate (Kennedy et al. 2015 table 6). Then give each
  ## component a share of the disturbance probability proportional to its weight,
  ## and the disturbed values to sum to 1 if they sum to greater than 1, and else
  ## set the undisturbed value to 1 - p(disturbed).
  ## within-class mixture weights (mclust proportions; sum to 1 within each class)
  weights <- unlist(sapply(dist.priors, function(x) x$weights))

  if (useDistProduct) {
    distidx.toname = list("10"="OTHER", "20"="CUT", "40"="FIRE", "50"="PEST")
    users_accuracy = c(1-d.prior[1], .77, .92, .77, .78, .78) ## 0 = none, 10 = other, 20 = cut, 30 = road, 40 = fire, 50 = biotic

    for (d in seq(names(dist.priors))) {
      d.name <- names(dist.priors)[d]
      has.dist <- which(dextract[site,] == d.name)
      if (length(has.dist) > 0) {
        is.dist <- which(names(dist.prior.probs) == distidx.toname[[d.name]]) + 1
        p.dist[is.dist, has.dist] <- users_accuracy[dextract[site, has.dist]/10 + 1]
      }
    }
  }

  ## Build the per-year categorical prior for BOTH the product and no-product paths:
  ## split each class's probability across its clusters via `weights` (so multi-cluster
  ## classes are not over-counted), then set the undisturbed prior to the remaining mass
  ## (= 1 - product accuracy in flagged years, = background otherwise). Applying this on
  ## the no-product path too means an UNflagged year has the same prior with or without
  ## the product -- the product only changes flagged years. Previously this lived inside
  ## the useDistProduct branch, so the no-product path kept an un-split, over-counted
  ## disturbance background that inflated probD relative to the product on unflagged sites.
  for (j in 1:ncol(p.dist)) {
    p.dist[2:(n.dist+1), j] <- p.dist[2:(n.dist+1), j] * weights   # split class prob across clusters
    dist.mass <- sum(p.dist[2:(n.dist+1), j])
    if (dist.mass > 1) {                                           # guard: cap total disturbed mass at 1
      p.dist[2:(n.dist+1), j] <- p.dist[2:(n.dist+1), j] / dist.mass
      dist.mass <- 1
    }
    p.dist[1, j] <- 1 - dist.mass                                  # undisturbed = remaining mass
  }

  ## Bundle the disturbance prior + per-class forecast params.
  disturb.params <- list(
    p = p.dist,
    n = n.dist,
    vals = dist.forecastparams,
    alloc.soil = 0.05
  )

  ## NT x 3 observation matrix with the wood-pool obs in column 3.
  Y <- matrix(NA, NT, 3)
  Y[1:ncol(Yw), 3] <- Yw[site, ]
  Y <- Y[!is.na(Y[,3]),,drop=FALSE] # drop rows with wood-pool=NA
  
  ## tau prior and grid
  prior <- switch(tauPrior,
    gamma      = list(type = "gamma", shape = 0.1, rate = 2e-4),
    loguniform = list(type = "loguniform"))
  grid <- list(tauMin = 0.1, tauMax = 10^4)
  list(Y = Y, disturb.params = disturb.params, PAR = rad[site, ], site = site, prior = prior, grid = grid)
}
```

```{r}
# Helper functions for ORJointUpdate

KalmanUpdate <- function(m, C, H, R, y) {
  S <- H %*% C %*% t(H) + R
  K <- C %*% t(H) %*% solve(S)
  m_hat <- m + K %*% (y - H %*% m)
  C_hat <- (diag(nrow(C)) - K %*% H) %*% C
  L_hat <- dnorm(y, mean = H %*% m, sd = sqrt(S))
  
  return(list(m_hat=m_hat, C_hat=C_hat, L_hat=L_hat))
}

# Initial tau weights: grid point g gets the prior mass of its cell,
#   int_{lo_g}^{hi_g} p(tau) dtau,  lo_g = tau_g - (tau_g - tau_{g-1})/2,  hi_g = tau_g + (tau_{g+1} - tau_g)/2,
# i.e. the cell edges are the midpoints to the neighbouring grid points, so the cell
# width follows the grid spacing. The edge cells stop at tau_1 and tau_G and the
# weights are renormalised (the prior is truncated to the grid range).
cell_prior_mass <- function(tauGrid, prior) {
  G   <- length(tauGrid)
  mid <- (tauGrid[-G] + tauGrid[-1]) / 2
  lo  <- c(tauGrid[1], mid)
  hi  <- c(mid, tauGrid[G])
  cdf <- switch(prior$type,
    gamma      = function(x) pgamma(x, shape = prior$shape, rate = prior$rate),
    loguniform = function(x) log(x))
  m <- cdf(hi) - cdf(lo)
  m / sum(m)
}

# moment matching procedure to project down to mixture of two gaussians
project <- function(w, l, mu, sig, D) {
  w1 <- 0
  for (j in 1:2) {
    w1 <- w1 + w[j,1]
  }
  w2 <- 1 - w1
  alpha <- w
  for (j in 1:2) {
    for (k in 0:D) {
      if (k == 0) {
        alpha[j,k+1] <- w[j,k+1] / w1
      } else {
        alpha[j,k+1] <- w[j,k+1] / w2
      }
    }
  }
  mu1 <- alpha[1,1] * mu[1,1,] + alpha[2,1] * mu[2,1,]
  sig1 <- array(0, dim=c(3,3))
  for (j in 1:2) {
    sig1 <- sig1 + alpha[j,1] * (tcrossprod(mu[j,1,]) + sig[j,1,,])
  }
  sig1 <- sig1 - tcrossprod(mu1)
  mu2 <- array(0, dim=c(3))
  for (j in 1:2) {
    for (k in 1:D) {
      mu2 <- mu2 + alpha[j,k+1] * mu[j,k+1,]
    }
  }
  sig2 <- array(0, dim=c(3,3))
  for (j in 1:2) {
    for (k in 1:D) {
      sig2 <- sig2 + alpha[j,k+1] * (tcrossprod(mu[j,k+1,]) + sig[j,k+1,,])
    }
  }
  sig2 <- sig2 - tcrossprod(mu2)
  return(list(w1=w1,w2=w2,mu1=mu1,mu2=mu2,sig1=sig1,sig2=sig2))
  
}
```

```{r}
##' Approximate joint disturbance update
##'
##' @param gamInit c(a, b) Gamma prior (shape, rate) for tau
##' @param Y NT x 3 observations (wood/AGB pool in column 3; NA => forecast-only step)
##' @param Rparams c(slope, intercept): observation error R = y*slope + intercept
##' @param Q process-error covariance; Q[1,1],Q[2,2] used for foliage/soil, Q[3,3] set to 1/tau_g
##' @param PAR daily PAR driver (length NT*365)
##' @param Xic J x 3 initial-condition ensemble (ICprior)
##' @param THETA VSEM parameter vector/matrix (cols 9:11 overwritten with the IC ensemble)
##' @param dist disturbance list: p (D+1 x NT prior), n (# classes), vals (per-class mu0/V0), alloc.soil
##' @param G number of precision values in the grid
##' @param N ensemble size
##' @param prior prior for tau: list(type = "gamma", shape, rate) or list(type = "loguniform")
##' @param grid list(tauMin, tauMax) for the log-spaced tau grid
##' @return list of length-NT vectors, each marginalised over the tau grid with the
##'   tau posterior pi_t(g) and over the mixture components (j, k):
##'   mean  = analysis mean of the wood pool
##'   var   = analysis variance of the wood pool (law of total variance)
##'   pDist = posterior probability of disturbance
##'   and the one-step-ahead innovation diagnostics, from the forecast predictive of y_t
##'   marginalised over the tau posterior pi_{t-1}(g), the components j and the
##'   disturbance prior beta_k (i.e. before y_t is assimilated):
##'   innov = y_t - E[y_t | y_{1:t-1}]
##'   nis   = innov^2 / S, with S = Var[y_t | y_{1:t-1}] = S.fc + S.q + S.R
##'   S.fc  = forecast spread (ensemble variance + spread of the component means)
##'   S.q   = process error, beta_0 * E_{pi_{t-1}}[1/tau]
##'   S.R   = observation error R_t
ORJointUpdate <- function(Y, Rparams, Q, PAR, Xic, THETA, dist, G = 200, N=100, prior, grid) {
  NT <- dim(Y)[1]
  D  <- dist$n                      # number of disturbance sub-classes
  icMean <- colMeans(Xic)
  icCov <- cov(Xic)
  
  # initialize grid
  tauGrid <- exp(seq(log(grid$tauMin), log(grid$tauMax), length.out = G)) # log-spaced grid
  # tauGrid <- seq(grid$tauMin, grid$tauMax, length.out = G)
  
  # initial tau weights: log prior mass of each grid cell (see cell_prior_mass)
  qs <- log(cell_prior_mass(tauGrid, prior))
  prevW <- array(0.5, dim=c(G, 2))
  prevMu <- array(0, dim=c(G, 2, 3))
  prevSig <- array(0, dim=c(G, 2, 3, 3))
  for (i in 1:G) {
    for (j in 1:2) {
      prevMu[i,j,] <- icMean
      prevSig[i,j,,] <- icCov
    }
  }
  
  hist <- list(
    mean  = rep(NA_real_, NT),   # wood-pool analysis mean
    var   = rep(NA_real_, NT),   # wood-pool analysis variance
    pDist = rep(NA_real_, NT),   # P(disturbed)
    innov = rep(NA_real_, NT),   # innovation y - predictive mean
    nis   = rep(NA_real_, NT),   # normalized innovation squared innov^2 / S
    S.fc  = rep(NA_real_, NT),   # S part: forecast spread
    S.q   = rep(NA_real_, NT),   # S part: process error
    S.R   = rep(NA_real_, NT)    # S part: observation error
  )
  
  for (t in 1:NT) {
    # sample from proposal distribution
    propSamples <- array(0, dim=c(N, 3))
    n <- 1L; tries <- 0L
    while (n <= N) {
      tries <- tries + 1L
      i <- sample.int(G, 1L); j <- sample.int(2L, 1L)
      x <- mvnfast::rmvn(1, prevMu[i,j,], prevSig[i,j,,])
      if (all(x > 0)) { propSamples[n, ] <- x; n <- n + 1L }
      if (tries > 100L * N) stop(sprintf("t=%d: proposal rejection rate too high", t))
    }
    
    muF <- array(0, dim=c(G, 2, D+1, 3))
    sigF <- array(0, dim=c(G, 2, D+1, 3, 3))
    tsel = (t-1)*365 + 1:365
    ## ---- once per timestep ----

    # 1. VSEM is deterministic in THETA -> identical for every k
    sim0 <- matrix(0, N, 3)
    for (n in seq_len(N)) {
      THETA[9:11] <- propSamples[n, ]
      sim0[n, ]   <- VSEM(THETA, PAR = PAR[tsel])[365, 2:4]
    }
    
    # 2. flatten (g,j) -> r, cache the Cholesky factors
    P        <- 2L * G
    compMu   <- matrix(0, P, 3)
    compChol <- vector("list", P)
    for (j in 1:2) for (g in 1:G) {
      r <- (j - 1L) * G + g
      compMu[r, ]   <- prevMu[g, j, ]
      compChol[[r]] <- chol(prevSig[g, j, , ])
    }
    
    # 3. one P x N density matrix, vectorized over samples
    logDens <- matrix(0, P, N)
    for (r in seq_len(P)) {
      logDens[r, ] <- mvnfast::dmvn(propSamples, compMu[r, ], compChol[[r]],
                                    log = TRUE, isChol = TRUE)
    }
    
    # 4. proposal density and all weights, in log space
    mx      <- apply(logDens, 2, max)
    logProp <- mx + log(colSums(exp(sweep(logDens, 2, mx)))) - log(P)
    logW    <- sweep(logDens, 2, logProp)
    W       <- exp(logW - apply(logW, 1, max))
    W       <- W / rowSums(W)          # P x N, rows sum to 1
    cN      <- 1 - rowSums(W^2)        # P
    
    for (k in 0:D) {
      sim <- sim0
      if (k > 0) {
        mu0 <- dist$vals[[k]]$mu0
        V0  <- dist$vals[[k]]$V0
        for (n in seq_len(N)) sim[n, ] <- disturbance.p(sim[n, ], mu0, V0, dist$alloc.soil)
      }
    
      MU <- W %*% sim                              # P x 3, every mean in one matmul
    
      for (r in seq_len(P)) {
        g  <- (r - 1L) %%  G + 1L
        j  <- (r - 1L) %/% G + 1L
        Sc <- sweep(sim, 2, MU[r, ])
        S  <- crossprod(Sc, W[r, ] * Sc) / cN[r]
        if (k == 0) {
          S <- S + Q
          S[3, 3] <- S[3, 3] - Q[3, 3] + 1 / tauGrid[g]
        }
        muF[g, j, k + 1, ]    <- MU[r, ]
        sigF[g, j, k + 1, , ] <- S
      }
    }
    
    # compute analysis
    lA <- array(0, dim=c(G, 2, D+1))
    muA <- array(0, dim=c(G, 2, D+1, 3))
    sigA <- array(0, dim=c(G, 2, D+1, 3, 3))
    H  <- matrix(c(0, 0, 1), 1, 3)
    R <- Y[t, 3] * Rparams[1] + Rparams[2] # heteroskedastic obs error
    for (j in 1:2) {
      for (k in 0:D) {
        for (g in 1:G) {
          out <- KalmanUpdate(muF[g,j,k+1,], sigF[g,j,k+1,,], H, R, Y[t, 3])
          muA[g,j,k+1,] <- out$m_hat
          sigA[g,j,k+1,,] <- out$C_hat
          lA[g,j,k+1] <- out$L_hat
        }
      }
    }
    
    # compute tau posterior on grid
    beta <- array(0, dim=c(D+1))
    for (k in 0:D) {
      beta[k+1] = dist$p[k+1,t]
    }

    # innovation diagnostics: moment-match the forecast predictive of y_t, a mixture over
    # (g, j, k) with weights pi_{t-1}(g) * prevW[g,j] * beta_k, each component
    # N(muF[g,j,k,3], sigF[g,j,k,3,3] + R). For k = 0, sigF[,,1,3,3] includes 1/tau_g.
    piPrev <- exp(qs - max(qs)); piPrev <- piPrev / sum(piPrev)          # pi_{t-1}(g)
    wF     <- outer(outer(piPrev, rep(1, 2)) * prevW, beta)               # G x 2 x (D+1)
    wF     <- wF / sum(wF)
    muY    <- sum(wF * muF[, , , 3])
    S.q    <- sum(wF[, , 1] / tauGrid)                                    # beta_0 E[1/tau]
    S.fc   <- sum(wF * (sigF[, , , 3, 3] + (muF[, , , 3] - muY)^2)) - S.q
    hist$innov[t] <- Y[t, 3] - muY
    hist$S.fc[t]  <- S.fc
    hist$S.q[t]   <- S.q
    hist$S.R[t]   <- R
    hist$nis[t]   <- hist$innov[t]^2 / (S.fc + S.q + R)
    logz <- array(0, dim=c(G))
    for (g in 1:G) {
      zsum <- 0
      for (j in 1:2) {
        for (k in 0:D) {
          zsum <- zsum + prevW[g,j] * beta[k+1] * lA[g,j,k+1]
        }
      }
      logz[g] <- log(zsum)
      qs[g] <- qs[g] + logz[g]
    }
    
    # tau posterior on the grid, pi_t(g)
    m <- max(qs)
    ws <- exp(qs - m) / sum(exp(qs - m))
    
    # analysis weights
    wA <- array(0, dim=c(G, 2, D+1))
    for (g in 1:G) {
      for (j in 1:2) {
        for (k in 0:D) {
          wA[g,j,k+1] <- prevW[g,j] * beta[k+1] * lA[g,j,k+1] / exp(logz[g])
        }
      }
    }
    
    # recompute prev posterior vals (KL projection)
    for (g in 1:G) {
      out <- project(wA[g,,], lA[g,,], muA[g,,,], sigA[g,,,,], D)
      prevW[g,1] <- out$w1
      prevW[g,2] <- out$w2
      prevMu[g,1,] <- out$mu1
      prevMu[g,2,] <- out$mu2
      prevSig[g,1,,] <- out$sig1
      prevSig[g,2,,] <- out$sig2
    }
    
    # store analysis values for this tstep, marginalised over the tau grid (weights
    # pi_t(g) = ws) and the components (j, k) (weights wA[g, , ], which sum to 1 for each g)
    pDist.g <- apply(wA[, , -1, drop = FALSE], 1, sum)                  # P_t(g)
    mean.g  <- apply(wA * muA[, , , 3], 1, sum)                           # E[wood | g]
    msq.g   <- apply(wA * (sigA[, , , 3, 3] + muA[, , , 3]^2), 1, sum)   # E[wood^2 | g]
    hist$mean[t]  <- sum(ws * mean.g)
    hist$var[t]   <- sum(ws * msq.g) - hist$mean[t]^2
    hist$pDist[t] <- sum(ws * pDist.g)
  }
  
  return(hist)
}

```

```{r}
## Generate ORJointUpdate inputs for each requested site, run the joint update,
## and save each site's diagnostics to experiments/jointUpdate. Returns the
## results invisibly (keyed by site) for convenience.
## Rparams.source ("GEDI"/"LandTrendr") and Rparams.model ("het"/"con") describe the
## observation-error parameters in use; together with tauPrior ("gamma"/"loguniform")
## they are recorded in each output filename.
run_joint_update_sites <- function(site.idx, Rparams = Rparams.GEDI.het,
                                   useDistProduct = TRUE,
                                   tauPrior = c("gamma", "loguniform"),
                                   Rparams.source = c("GEDI", "LandTrendr"),
                                   Rparams.model  = c("het", "con"),
                                   outdir = "experiments/jointUpdateTestSep") {
  Rparams.source <- match.arg(Rparams.source)
  Rparams.model  <- match.arg(Rparams.model)
  tauPrior       <- match.arg(tauPrior)
  if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
  results <- list()
  for (site in site.idx) {
    message("running joint update for site ", site)
    # function(Y, Rparams, Q, PAR, Xic, THETA, dist, G = 200, N=100, prior)
    args <- make_or_args(site, useDistProduct, tauPrior)
    out <- ORJointUpdate(Y = args$Y, Rparams = Rparams,
                         Q = Q, PAR = args$PAR, Xic = ICprior,
                         THETA = theta[1:11], dist = args$disturb.params, prior=args$prior,
                         grid = args$grid)
    fname <- file.path(outdir,
                       sprintf("site_%d_distprod_%s_%s_%s_tau%s.rds",
                               site, useDistProduct, Rparams.source, Rparams.model, tauPrior))
    saveRDS(out, fname)
    results[[as.character(site)]] <- out
  }
  invisible(results)
}
```

```{r run-single}
## Run + save for the site(s) of interest.
## tauPrior = "gamma" or "loguniform" selects the prior on the process precision tau.
run_joint_update_sites(site.idx = c(300), Rparams = Rparams.GEDI.het, useDistProduct = TRUE,
                       tauPrior = "loguniform", Rparams.source = "GEDI", Rparams.model = "het")
```

## experiment grid

Runs every combination of observation-error model (GEDI / LandTrendr magnitude x
heteroskedastic / constant R) and disturbance product (with / without), i.e. 8
experiments, over `grid.sites`. Each (experiment, site) run is saved to `grid.outdir`
with the usual `site_<idx>_distprod_<TF>_<source>_<model>_tau<prior>.rds` name, which
`OR_joint_update_analysis.Rmd` parses. Runs are spread over `grid.cores` forked
workers; existing files are skipped unless `grid.overwrite = TRUE`.

```{r run-grid}
grid.outdir    <- "experiments/testRun"
grid.tauPrior  <- "gamma"
grid.overwrite <- FALSE
grid.cores     <- max(1, parallel::detectCores() - 2)

## test sites: one random site per 2004 disturbance-product label
## (0 = none, 10 = other, 20 = cut, 40 = fire, 50 = biotic) -> 5 sites
set.seed(2004)
lab04      <- dextract[, "t2004"]
grid.sites <- sapply(sort(unique(lab04)), function(l) {
  s <- which(lab04 == l)
  s[sample.int(length(s), 1)]
})
## grid.sites <- seq_len(nrow(dextract))   # full run

## observation-error models: label -> (Rparams, source, model)
Rparams.grid <- list(
  GEDI_het       = list(Rparams = Rparams.GEDI.het,         source = "GEDI",       model = "het"),
  GEDI_con       = list(Rparams = Rparams.GEDI.fixed,       source = "GEDI",       model = "con"),
  LandTrendr_het = list(Rparams = Rparams.Landtrendr.het,   source = "LandTrendr", model = "het"),
  LandTrendr_con = list(Rparams = Rparams.Landtrendr.fixed, source = "LandTrendr", model = "con")
)
exp.grid <- expand.grid(R = names(Rparams.grid), useDistProduct = c(TRUE, FALSE),
                        site = grid.sites, stringsAsFactors = FALSE)

run_grid_row <- function(i) {
  e <- exp.grid[i, ]
  r <- Rparams.grid[[e$R]]
  fname <- file.path(grid.outdir, sprintf("site_%d_distprod_%s_%s_%s_tau%s.rds", e$site,
                                          e$useDistProduct, r$source, r$model, grid.tauPrior))
  if (file.exists(fname) && !grid.overwrite) return(fname)
  run_joint_update_sites(e$site, Rparams = r$Rparams, useDistProduct = e$useDistProduct,
                         tauPrior = grid.tauPrior, Rparams.source = r$source,
                         Rparams.model = r$model, outdir = grid.outdir)
  fname
}

RNGkind("L'Ecuyer-CMRG"); set.seed(1)
grid.out <- parallel::mclapply(seq_len(nrow(exp.grid)), run_grid_row, mc.cores = grid.cores)
failed   <- which(sapply(grid.out, inherits, "try-error"))
if (length(failed)) print(cbind(exp.grid[failed, ], error = unlist(grid.out[failed])))
```

## analysis

```{r}
## Load a saved ORJointUpdate run. The site index and run settings are parsed from
## the file name (site_<idx>_distprod_<TF>_<source>_<model>_tau<prior>.rds), and the
## matching wood-pool observations are taken from Yw (same NA handling as make_or_args).
load_hist <- function(file) {
  h     <- readRDS(file)
  tag   <- sub("\\.rds$", "", basename(file))
  site  <- as.integer(sub("^site_(\\d+)_.*$", "\\1", tag))
  obs   <- Yw[site, ]
  h$obs  <- obs[!is.na(obs)]
  h$year <- dtime[seq_along(h$mean)]
  h$site <- site
  h$tag  <- tag
  h
}

## shared look: one palette, light grid, no box
.pal <- list(mean = "#2a78d6", band = "#2a78d6", obs = "#0b0b0b", dist = "#eb6834",
             grid = "#e6e6e3", ink = "#52514e")
.base_par <- function() par(mar = c(4.2, 4.8, 3.2, 1), las = 1, bty = "n",
                            col.axis = .pal$ink, fg = .pal$ink, cex.axis = 0.9)

## Wood-pool analysis: mean (line), 95% interval mean +/- 1.96 sd (band), observations (points)
plot_analysis <- function(h, main = sprintf("Site %d: wood-pool analysis", h$site)) {
  op <- .base_par(); on.exit(par(op))
  x  <- h$year; sd <- sqrt(pmax(h$var, 0))
  lo <- h$mean - 1.96 * sd; hi <- h$mean + 1.96 * sd
  plot(NA, xlim = range(x), ylim = range(c(lo, hi, h$obs), na.rm = TRUE),
       xlab = "year", ylab = expression("wood carbon (kg C m"^-2*")"), main = main)
  abline(h = axTicks(2), col = .pal$grid)
  polygon(c(x, rev(x)), c(lo, rev(hi)), col = adjustcolor(.pal$band, 0.2), border = NA)
  lines(x, h$mean, col = .pal$mean, lwd = 2.5)
  points(x[seq_along(h$obs)], h$obs, pch = 21, bg = .pal$obs, col = "white", cex = 1.1)
  legend("topright", bty = "n", cex = 0.85, text.col = .pal$ink,
         legend = c("analysis mean", "95% interval", "observation"),
         col = c(.pal$mean, adjustcolor(.pal$band, 0.35), .pal$obs),
         lwd = c(2.5, 10, NA), pch = c(NA, NA, 16))
  invisible(h)
}

## Posterior probability of disturbance per year (bars), with the 0.5 line
plot_pdist <- function(h, main = sprintf("Site %d: probability of disturbance", h$site)) {
  op <- .base_par(); on.exit(par(op))
  x <- h$year; p <- h$pDist
  plot(NA, xlim = range(x) + c(-0.6, 0.6), ylim = c(0, 1), xlab = "year",
       ylab = "P(disturbed)", main = main)
  abline(h = seq(0, 1, 0.25), col = .pal$grid)
  abline(h = 0.5, col = .pal$ink, lty = 2)
  rect(x - 0.38, 0, x + 0.38, p, col = .pal$dist, border = NA)
  big <- which(p >= 0.5)
  if (length(big)) text(x[big], p[big], sprintf("%.2f", p[big]), pos = 3, cex = 0.75,
                        col = .pal$ink, xpd = NA)
  invisible(h)
}
```

```{r, fig.width = 9, fig.height = 4.5}
h <- load_hist("experiments/jointUpdateTestSep/site_300_distprod_TRUE_GEDI_het_taugamma.rds")
plot_analysis(h)
plot_pdist(h)
```
