---
title: "OR_joint_update"
output: html_document
author: "Jacob Epstein"
date: "5/30/26"
---

```{r}
library(statmod)
library(pracma)
library(BayesianTools)       # VSEM forest model
source("StatsAuxiliary.R")   # normal_quantile(p, mean, variance)
source("disturbance.R")      # disturbance.p() forecast perturbation
## Load disturbance rate, state&transition, disturbed biomass stats
load("STprior.RData")
load("extract.h03v04.RData")
load("dis.postprocessed.RData")
## The products cover different years (see StateTransition.Rmd), so every lookup
## goes through calendar year rather than column index:
##   dextract (disturbance product): 1985:2012
##   bextract (biomass / Yw):        1990:2017
##   DA timesteps:                   1990:2017 (timestep t <-> year atime[t])
dtime <- 1985:2012
ctime <- 1990:2017
atime <- 1990:2017
colnames(dextract) <- paste0("t",dtime)
dextract[is.na(dextract)] <- 0
d.prior["UN"] = d.prior["UN"] + d.prior[6]                         ## change at runtime (removes <NA>)
d.prior = d.prior[-6]

## Load Calibration
load("fit.RData")
Qsoil = Qs
```

```{r}
latlon = points@coords # lat / lon of each site in dextract 
## Cached met must be for exactly these sites, in this order: rad[i,] is used as
## site i's driver. (The older daymet.RData was for a different 559-point sample.)
metFile = "daymet.h03v04.RData"
met_matches_points <- function(dm) {
  length(dm) == nrow(latlon) &&
    all(abs(sapply(dm, function(d) d$longitude[1]) - latlon[,1]) < 1e-3) &&
    all(abs(sapply(dm, function(d) d$latitude[1])  - latlon[,2]) < 1e-3)
}
if(file.exists(metFile)){
  load(metFile)
  stopifnot(met_matches_points(dm))
} else {
  dm = list()
  for(i in 1:nrow(latlon)){
    dm[[i]] = daymetr::download_daymet(site="h03v04",latlon[i,2],latlon[i,1],start=1985,end=2017,internal=TRUE,simplify=TRUE)
  }
  save(dm,file=metFile)
}

## extract Shortwave and convert to daily PAR
SW2PAR = 0.45
srad = subset(dm[[1]],measurement == "srad..W.m.2.")   
metTime <- as.POSIXct(paste(srad$year,srad$yday,sep="-"),format="%Y-%j")
rad <- matrix(NA,length(dm),length(metTime)) # [number of sites x number of measurements]
print(dim(rad))
for(i in seq_along(dm)){
  srad = subset(dm[[i]],measurement == "srad..W.m.2.")   
  rad[i,] = srad$value*1e-6*86400*SW2PAR  # -> MJ /m2 /day PAR (convert) # QUESTION -- should the changing day length, from the daymet day length values, be used (instead of a fixed 24-hr day length)?
}
plot(metTime,rad[1,],type='l')

## PAR restricted to the DA window (daymet has 365 days every year, so each DA
## timestep is one contiguous 365-day block). The full 1985-2017 `rad` is still
## used for spin-up below.
radDA <- rad[, srad$year %in% atime, drop = FALSE]
stopifnot(ncol(radDA) == 365 * length(atime))
## plot(date,PAR,col=2) ## calibration data
```

```{r}
dist <- list()
dist_name <- c(
  "0"  = "none",
  "10" = "other",
  "20" = "cut",
  "30" = "road",
  "40" = "fire",
  "50" = "biotic"
)

dist.forecastparams <- list()

count <- 1
for (dclass in names(dist.priors)) {
  dist.name <- as.character(dist_name[dclass])
  nclust <- length(dist.priors[[dclass]]$means)
  
  for (i in 1:nclust) {
    dist.forecastparams[[count]] <- list()
    dist.forecastparams[[count]]$mu0 <- c(0.2, dist.priors[[dclass]]$means[i])
    dist.forecastparams[[count]]$V0 <- diag(c(0.04, dist.priors[[dclass]]$vars[i]))
    dist.forecastparams[[count]]$name <- paste0(dist.name, i)
    count <- count + 1
  }
}
```

```{r}
## BIOMASS
B2C = 0.5 ##biomass to carbon ratio
Yw <- bextract*B2C*1000*1e-4 ## convert Mg/ha (Mg per hectacre)  to  kg C [nsite x 28 years, 1990:2017]
colnames(Yw) <- paste0("t", ctime)
Rw <- (84*B2C*1000*1e-4)^2 ## Observation Error: Kennedy et al 2018 Environ. Res. Lett. 13 025004; Fig 2 RMSE
## units: Mg/ha -> kg/m2
print(dim(Yw))
## grab the distribution of AGB at the start of the DA window from all pixels
AGBprior <- Yw[, paste0("t", atime[1])]
ne = length(AGBprior) ## number of ensemble members

## spin up
PARlong <- as.vector(t(rad)) # concatenate PAR time series for each site into one vector
spin = VSEM(theta,PARlong) # Question: why run VSEM on the concatenated PAR? What is theta?
thin = seq(nrow(spin)/2,nrow(spin),length=ne)
spinIC = spin[thin,2:4] # select ne samples from the spin-up 
#for(i in 2:4){
#  hist(spin[thin,i],type='l')
#}

## disturbance recovery trajectory
nNBG = 200 ## years
NBG <- array(NA,c(ne,nNBG,3))
for(i in 1:ne){
  param = theta
  param[9:11] <- spinIC[i,]*c(0.01,1,0.01) ## set leaf and wood to 1% (simulate disturbance)
  raw = VSEM(param,PARlong[1:(nNBG*365.25)]) # simulate for nNBG years
  NBG[i,,] = raw[round(seq(365,nNBG*365.25,length=nNBG)),2:4]  ## grab pools at the end of each year
}
  
## for each AGBprior, grab the matching leaf and soil
ICprior <- matrix(NA,ne,3)
age = rep(NA,ne)
for(i in 1:ne){
  age[i] <- which.min((AGBprior[i]-NBG[i,,3])^2) # age = year in the disturbance recover trajectory where the simulated site has similar AGB to the AGB prior information
  ICprior[i,] <- c(NBG[i,age[i],1:2],AGBprior[i]) # use AGBprior for the model ICs
}
hist(age)
colnames(ICprior) <- c("Cv","Cs","CR") 
pairs(ICprior)
```

```{r}

## ---- Site-independent setup (computed once) --------------------------------
## Number of disturbance sub-classes and number of assimilation years.
n.dist = length(dist.forecastparams)
NT = length(atime)

## Background disturbance-type probability per forecast component, by matching
## the component's name to one of "other"/"cut"/"fire"/"biotic".
distname.toidx <- list("other"=2, "cut"=3, "fire"=4, "biotic"=5)
dist.prior.probs <- sapply(dist.forecastparams, function(x) {
  name <- x$name
  for (d in names(distname.toidx)) {
    if (grepl(d, name)) {
      return(d.prior[distname.toidx[[d]]])
    }
  }
})

## Diagonal process-error covariance (per-pool SDs) and prior process-precision
## gamma params (a, b) = (shape, rate). Both site-independent.
Q = diag(c(Ql,Qsoil,Qw)^2)
a <- 1/0.1172475
b <- 1
gamInit <- c(a, b)

## Default heteroskedastic Rparams for GEDI (slope, intercept).
Rparams.GEDI.het <- c(0.01331, 0.09761)
Rparams.GEDI.fixed <- c(0.0000000, 0.8848249)
Rparams.Landtrendr.het <- c(0.013310, 3.012645)
Rparams.Landtrendr.fixed <- c(0.000, 3.79986)

## ---- Per-site argument builder ---------------------------------------------
## make_or_args(site, useDistProduct) assembles the site-specific inputs to
## ORJointUpdate: the (n.dist+1) x NT prior probability matrix p.dist (row 1 =
## undisturbed, rows 2.. = disturbance sub-classes), the bundled disturb.params,
## the NT x 3 observation matrix Y (wood-pool obs in column 3), and the PAR
## driver for that site. Relies on the site-independent globals above plus
## d.prior, dist.priors, dist.forecastparams, dextract, Yw and rad.
## tauPrior selects the prior on the process precision tau:
##   "gamma"      : Gamma(shape = 0.1, rate = 2e-4), which spreads mass over every
##                  decade of the grid (mean tau = 500)
##   "loguniform" : p(tau) proportional to 1/tau (flat in log tau)
make_or_args <- function(site, useDistProduct = TRUE, tauPrior = c("gamma", "loguniform")) {
  tauPrior <- match.arg(tauPrior)
  ## Initialize every year with the background prior (same column for all t).
  p.dist <- array(NA, dim = c(n.dist + 1, NT))
  for (t in 1:NT) {
    p.dist[1:(n.dist+1), t] <- as.numeric(c(d.prior[1], unlist(dist.prior.probs)))
  }

  ## If using the disturbance product, override the prior in years where a
  ## disturbance was detected at this site: replace that class's prior with the
  ## user's-accuracy estimate (Kennedy et al. 2015 table 6). Then give each
  ## component a share of the disturbance probability proportional to its weight,
  ## and the disturbed values to sum to 1 if they sum to greater than 1, and else
  ## set the undisturbed value to 1 - p(disturbed).
  ## within-class mixture weights (mclust proportions; sum to 1 within each class)
  weights <- unlist(sapply(dist.priors, function(x) x$weights))

  if (useDistProduct) {
    distidx.toname = list("10"="OTHER", "20"="CUT", "40"="FIRE", "50"="PEST")
    users_accuracy = c(1-d.prior[1], .77, .92, .77, .78, .78) ## 0 = none, 10 = other, 20 = cut, 30 = road, 40 = fire, 50 = biotic

    ## product years inside the DA window and their DA timesteps
    dyrs <- intersect(dtime, atime)
    dprod <- dextract[site, paste0("t", dyrs)]
    for (d in seq(names(dist.priors))) {
      d.name <- names(dist.priors)[d]
      has.dist <- which(dprod == d.name)                 # index into dyrs
      if (length(has.dist) > 0) {
        is.dist <- which(names(dist.prior.probs) == distidx.toname[[d.name]]) + 1
        p.dist[is.dist, match(dyrs[has.dist], atime)] <- users_accuracy[dprod[has.dist]/10 + 1]
      }
    }
  }

  ## Build the per-year categorical prior for BOTH the product and no-product paths:
  ## split each class's probability across its clusters via `weights` (so multi-cluster
  ## classes are not over-counted), then set the undisturbed prior to the remaining mass
  ## (= 1 - product accuracy in flagged years, = background otherwise). Applying this on
  ## the no-product path too means an UNflagged year has the same prior with or without
  ## the product -- the product only changes flagged years. Previously this lived inside
  ## the useDistProduct branch, so the no-product path kept an un-split, over-counted
  ## disturbance background that inflated probD relative to the product on unflagged sites.
  for (j in 1:ncol(p.dist)) {
    p.dist[2:(n.dist+1), j] <- p.dist[2:(n.dist+1), j] * weights   # split class prob across clusters
    dist.mass <- sum(p.dist[2:(n.dist+1), j])
    if (dist.mass > 1) {                                           # guard: cap total disturbed mass at 1
      p.dist[2:(n.dist+1), j] <- p.dist[2:(n.dist+1), j] / dist.mass
      dist.mass <- 1
    }
    p.dist[1, j] <- 1 - dist.mass                                  # undisturbed = remaining mass
  }

  ## Bundle the disturbance prior + per-class forecast params.
  disturb.params <- list(
    p = p.dist,
    n = n.dist,
    vals = dist.forecastparams,
    alloc.soil = 0.05
  )

  ## NT x 3 observation matrix with the wood-pool obs in column 3.
  ## Rows stay aligned with DA timesteps; years without an observation stay NA
  ## (forecast-only step in ORJointUpdate) rather than being dropped, which would
  ## shift Y out of step with p.dist and PAR.
  Y <- matrix(NA, NT, 3)
  oyrs <- intersect(ctime, atime)
  Y[match(oyrs, atime), 3] <- Yw[site, paste0("t", oyrs)]

  ## tau prior and grid
  prior <- switch(tauPrior,
    gamma      = list(type = "gamma", shape = 0.1, rate = 2e-4),
    loguniform = list(type = "loguniform"))
  grid <- list(tauMin = 0.1, tauMax = 10^4)
  list(Y = Y, disturb.params = disturb.params, PAR = radDA[site, ], site = site, prior = prior, grid = grid)
}
```

```{r}
# Helper functions for ORJointUpdate

KalmanUpdate <- function(m, C, H, R, y) {
  S <- H %*% C %*% t(H) + R
  K <- C %*% t(H) %*% solve(S)
  m_hat <- m + K %*% (y - H %*% m)
  C_hat <- (diag(nrow(C)) - K %*% H) %*% C
  L_hat <- dnorm(y, mean = H %*% m, sd = sqrt(S))
  
  return(list(m_hat=m_hat, C_hat=C_hat, L_hat=L_hat))
}

# Quantiles of a 1-d Gaussian mixture sum_i w_i N(mu_i, sd_i^2), by root-finding on its
# CDF. Components with negligible weight are dropped first.
mix_quantile <- function(p, w, mu, sd) {
  keep <- w > 1e-12 * sum(w)
  w <- w[keep] / sum(w[keep]); mu <- mu[keep]; sd <- sd[keep]
  lo <- min(mu - 8 * sd); hi <- max(mu + 8 * sd)
  sapply(p, function(pp) uniroot(function(x) sum(w * pnorm(x, mu, sd)) - pp,
                                 c(lo, hi), tol = 1e-8)$root)
}

# Initial tau weights: grid point g gets the prior mass of its cell,
#   int_{lo_g}^{hi_g} p(tau) dtau,  lo_g = tau_g - (tau_g - tau_{g-1})/2,  hi_g = tau_g + (tau_{g+1} - tau_g)/2,
# i.e. the cell edges are the midpoints to the neighbouring grid points, so the cell
# width follows the grid spacing. The edge cells stop at tau_1 and tau_G and the
# weights are renormalised (the prior is truncated to the grid range).
cell_prior_mass <- function(tauGrid, prior) {
  G   <- length(tauGrid)
  mid <- (tauGrid[-G] + tauGrid[-1]) / 2
  lo  <- c(tauGrid[1], mid)
  hi  <- c(mid, tauGrid[G])
  cdf <- switch(prior$type,
    gamma      = function(x) pgamma(x, shape = prior$shape, rate = prior$rate),
    loguniform = function(x) log(x))
  m <- cdf(hi) - cdf(lo)
  m / sum(m)
}

# moment matching procedure to project down to mixture of two gaussians
project <- function(w, l, mu, sig, D) {
  w1 <- 0
  for (j in 1:2) {
    w1 <- w1 + w[j,1]
  }
  w2 <- 1 - w1
  alpha <- w
  for (j in 1:2) {
    for (k in 0:D) {
      if (k == 0) {
        alpha[j,k+1] <- w[j,k+1] / w1
      } else {
        alpha[j,k+1] <- w[j,k+1] / w2
      }
    }
  }
  mu1 <- alpha[1,1] * mu[1,1,] + alpha[2,1] * mu[2,1,]
  sig1 <- array(0, dim=c(3,3))
  for (j in 1:2) {
    sig1 <- sig1 + alpha[j,1] * (tcrossprod(mu[j,1,]) + sig[j,1,,])
  }
  sig1 <- sig1 - tcrossprod(mu1)
  mu2 <- array(0, dim=c(3))
  for (j in 1:2) {
    for (k in 1:D) {
      mu2 <- mu2 + alpha[j,k+1] * mu[j,k+1,]
    }
  }
  sig2 <- array(0, dim=c(3,3))
  for (j in 1:2) {
    for (k in 1:D) {
      sig2 <- sig2 + alpha[j,k+1] * (tcrossprod(mu[j,k+1,]) + sig[j,k+1,,])
    }
  }
  sig2 <- sig2 - tcrossprod(mu2)
  # a group with zero total weight has undefined moments (0/0 above); give it the other
  # group's moments so it stays finite -- its weight is 0, so it only enters the proposal
  if (!(w1 > 0)) { mu1 <- mu2; sig1 <- sig2 }
  if (!(w2 > 0)) { mu2 <- mu1; sig2 <- sig1 }
  return(list(w1=w1,w2=w2,mu1=mu1,mu2=mu2,sig1=sig1,sig2=sig2))
  
}

# project() applied to every tau grid point g at once. w: G x 2 x (D+1) weights,
# mu: G x 2 x (D+1) x 3, sig: G x 2 x (D+1) x 3 x 3. Group 1 is the undisturbed
# components (k = 0), group 2 the disturbed ones (k >= 1). Returns w (G x 2),
# mu (G x 2 x 3) and sig (G x 2 x 3 x 3), indexed [g, group, ...].
project_grid <- function(w, mu, sig) {
  G <- dim(w)[1]; C <- prod(dim(w)[-1])        # components c = j + 2k, j fastest
  muC  <- array(mu,  c(G, C, 3))
  sigC <- array(sig, c(G, C, 3, 3))
  w1 <- w[, 1, 1] + w[, 2, 1]
  w2 <- 1 - w1
  grp <- list(seq_len(C) <= 2, seq_len(C) > 2)  # group 1: k = 0
  wg  <- list(w1, w2)
  out <- list(w = cbind(w1, w2, deparse.level = 0), mu = array(0, c(G, 2, 3)),
              sig = array(0, c(G, 2, 3, 3)))
  for (h in 1:2) {
    A <- matrix(w, G, C) / wg[[h]]              # alpha: within-group weights
    A[, !grp[[h]]] <- 0
    for (a in 1:3) out$mu[, h, a] <- rowSums(A * muC[, , a])
    for (a in 1:3) for (b in 1:3)
      out$sig[, h, a, b] <- rowSums(A * (muC[, , a] * muC[, , b] + sigC[, , a, b])) -
                            out$mu[, h, a] * out$mu[, h, b]
  }
  # a group with zero total weight has undefined moments (0/0 above); give it the other
  # group's moments so it stays finite -- its weight is 0, so it only enters the proposal
  z1 <- !(w1 > 0); z2 <- !(w2 > 0) & !z1
  out$mu[z1, 1, ]    <- out$mu[z1, 2, ];    out$sig[z1, 1, , ] <- out$sig[z1, 2, , ]
  out$mu[z2, 2, ]    <- out$mu[z2, 1, ];    out$sig[z2, 2, , ] <- out$sig[z2, 1, , ]
  out
}
```

```{r}
##' Approximate joint disturbance update
##'
##' @param gamInit c(a, b) Gamma prior (shape, rate) for tau
##' @param Y NT x 3 observations (wood/AGB pool in column 3; NA => forecast-only step)
##' @param Rparams c(slope, intercept): observation error R = y*slope + intercept
##' @param Q process-error covariance; Q[1,1],Q[2,2] used for foliage/soil, Q[3,3] set to 1/tau_g
##' @param PAR daily PAR driver (length NT*365)
##' @param Xic J x 3 initial-condition ensemble (ICprior)
##' @param THETA VSEM parameter vector/matrix (cols 9:11 overwritten with the IC ensemble)
##' @param dist disturbance list: p (D+1 x NT prior), n (# classes), vals (per-class mu0/V0), alloc.soil
##' @param G number of precision values in the grid
##' @param N ensemble size
##' @param prior prior for tau: list(type = "gamma", shape, rate) or list(type = "loguniform")
##' @param grid list(tauMin, tauMax) for the log-spaced tau grid
##' @param bias NULL (no bias correction) or a list configuring an online additive
##'   bias correction b_t of the undisturbed (k = 0) wood-pool forecast mean:
##'     rho    discount factor in (0, 1] on past residuals (default 0.9)
##'     denom  "N": b_{t+1} = S_t / N_t (weighted discounted mean, default);
##'            "t": b_{t+1} = S_t / n_t, n_t = # obs used so far (shrinks b toward 0)
##'     skip   number of initial observed steps not used for b (default 1). The first
##'            forecast starts from the population-wide IC prior (ICprior), so its
##'            residual measures IC mismatch, not VSEM bias.
##'     burnin number of observed steps after `skip` that use w_t = 1 instead of
##'            P(z_t = 0 | y_{1:t}) (guards against bias being absorbed as disturbance)
##'     wMin   b is only updated on steps with P(z_t = 0 | y_{1:t}) >= wMin (after burn-in)
##'   Each observed step, with posterior component weights
##'     tw_{gjk} prop. to pi_{t-1}(g) prevW[g,j] beta_k N(y_t; muF_{gjk,3} + b_t 1{k=0}, sigF_{gjk,33} + R_t),
##'   r_t = sum_{gj} tw_{gj0} (y_t - muF_{gj0,3}) / sum_{gj} tw_{gj0}   (residual vs the RAW forecast),
##'   w_t = sum_{gj} tw_{gj0} = P(z_t = 0 | y_{1:t}),
##'   S_t = rho S_{t-1} + w_t r_t,  N_t = rho N_{t-1} + w_t,  b_{t+1} = S_t / N_t.
##'   b_t is added to muF[, , k = 0, 3] (only the undisturbed components) before the
##'   analysis at step t; b_1 = 0. y_t only enters b from t + 1 onward.
##' @return list of length-NT vectors, each marginalised over the tau grid with the
##'   tau posterior pi_t(g) and over the mixture components (j, k):
##'   mean  = analysis mean of the wood pool
##'   var   = analysis variance of the wood pool (law of total variance)
##'   pDist = posterior probability of disturbance
##'   ci95  = NT x 2 matrix, 2.5% / 97.5% quantiles of the wood-pool analysis mixture
##'   pit   = probability integral transform of y_t under its one-step-ahead predictive
##'           (marginalised over the tau posterior pi_{t-1}(g), the components j and the
##'           disturbance prior beta_k, i.e. before y_t is assimilated); NA without obs
##'   wDrop = forecast weight of components dropped for degenerate importance weights
##'   bias, biasResid, biasW = b_t applied at step t, and r_t, w_t (NA without obs)
##'   predMeanRaw = predictive mean of y_t without the bias shift
ORJointUpdate <- function(Y, Rparams, Q, PAR, Xic, THETA, dist, G = 200, N=100, prior, grid,
                          bias = NULL) {
  if (!is.null(bias)) {
    bias <- modifyList(list(rho = 0.9, denom = "N", skip = 1, burnin = 0, wMin = 0), bias)
    bias$denom <- match.arg(bias$denom, c("N", "t"))
  }
  bS <- 0; bN <- 0; bObs <- 0; bHat <- 0   # bias-correction state (S_t, N_t, # obs, b_{t+1})
  NT <- dim(Y)[1]
  D  <- dist$n                      # number of disturbance sub-classes
  icMean <- colMeans(Xic)
  icCov <- cov(Xic)
  
  # initialize grid
  tauGrid <- exp(seq(log(grid$tauMin), log(grid$tauMax), length.out = G)) # log-spaced grid
  # tauGrid <- seq(grid$tauMin, grid$tauMax, length.out = G)
  
  # initial tau weights: log prior mass of each grid cell (see cell_prior_mass)
  qs <- log(cell_prior_mass(tauGrid, prior))
  prevW <- array(0.5, dim=c(G, 2))
  prevMu <- array(0, dim=c(G, 2, 3))
  prevSig <- array(0, dim=c(G, 2, 3, 3))
  for (i in 1:G) {
    for (j in 1:2) {
      prevMu[i,j,] <- icMean
      prevSig[i,j,,] <- icCov
    }
  }
  
  hist <- list(
    mean  = rep(NA_real_, NT),   # wood-pool analysis mean
    var   = rep(NA_real_, NT),   # wood-pool analysis variance
    ci95  = matrix(NA_real_, NT, 2, dimnames = list(NULL, c("lo", "hi"))),  # 95% interval
    pDist = rep(NA_real_, NT),   # P(disturbed)
    pit   = rep(NA_real_, NT),   # P(y_t' <= y_t | y_{1:t-1}), one-step-ahead PIT
    wDrop = rep(NA_real_, NT),   # forecast weight of degenerate (dropped) components
    ## one-step-ahead predictive of y_t (NA on steps with no observation):
    predMean    = rep(NA_real_, NT),   # E[y_t | y_{1:t-1}]
    predVar     = rep(NA_real_, NT),   # Var[y_t | y_{1:t-1}], including R
    predProcVar = rep(NA_real_, NT),   # its process-error part: E over components of 1{k=0}/tau_g
    predR       = rep(NA_real_, NT),   # its observation-error part R_t
    procVar     = rep(NA_real_, NT),   # posterior mean process variance E[1/tau | y_{1:t}]
    predMeanRaw = rep(NA_real_, NT),   # E[y_t | y_{1:t-1}] without the bias shift
    bias        = rep(0, NT),          # b_t added to the k = 0 wood forecast mean at step t
    biasResid   = rep(NA_real_, NT),   # r_t: P(z=0)-weighted residual vs the raw k = 0 forecast
    biasW       = rep(NA_real_, NT)    # w_t = P(z_t = 0 | y_{1:t})
  )
  
  for (t in 1:NT) {
    # sample from proposal distribution
    propSamples <- array(0, dim=c(N, 3))
    n <- 1L; tries <- 0L
    while (n <= N) {
      tries <- tries + 1L
      i <- sample.int(G, 1L); j <- sample.int(2L, 1L)
      x <- mvnfast::rmvn(1, prevMu[i,j,], prevSig[i,j,,])
      if (all(x > 0)) { propSamples[n, ] <- x; n <- n + 1L }
      if (tries > 100L * N) stop(sprintf("t=%d: proposal rejection rate too high", t))
    }
    
    muF <- array(0, dim=c(G, 2, D+1, 3))
    sigF <- array(0, dim=c(G, 2, D+1, 3, 3))
    tsel = (t-1)*365 + 1:365
    ## ---- once per timestep ----

    # 1. VSEM is deterministic in THETA -> identical for every k
    sim0 <- matrix(0, N, 3)
    for (n in seq_len(N)) {
      THETA[9:11] <- propSamples[n, ]
      sim0[n, ]   <- VSEM(THETA, PAR = PAR[tsel])[365, 2:4]
    }
    
    # 2. flatten (g,j) -> r, cache the Cholesky factors
    P        <- 2L * G
    compMu   <- matrix(0, P, 3)
    compChol <- vector("list", P)
    for (j in 1:2) for (g in 1:G) {
      r <- (j - 1L) * G + g
      compMu[r, ]   <- prevMu[g, j, ]
      compChol[[r]] <- chol(prevSig[g, j, , ])
    }
    
    # 3. one P x N density matrix, vectorized over samples
    logDens <- matrix(0, P, N)
    for (r in seq_len(P)) {
      logDens[r, ] <- mvnfast::dmvn(propSamples, compMu[r, ], compChol[[r]],
                                    log = TRUE, isChol = TRUE)
    }
    
    # 4. proposal density and all weights, in log space
    mx      <- apply(logDens, 2, max)
    logProp <- mx + log(colSums(exp(sweep(logDens, 2, mx)))) - log(P)
    logW    <- sweep(logDens, 2, logProp)
    W       <- exp(logW - apply(logW, 1, max))
    W       <- W / rowSums(W)          # P x N, rows sum to 1
    cN      <- 1 - rowSums(W^2)        # P

    # A component whose importance weights collapse onto ~1 sample (cN ~ 0) has no
    # estimable forecast moments (S / cN -> Inf, then NaN likelihoods). This happens to
    # components left far from every proposal sample, e.g. the undisturbed component at
    # high tau right after a clear-cut, whose weight is then negligible. Such components
    # get the unweighted ensemble moments as a finite placeholder and zero likelihood
    # below (so zero weight); hist$wDrop records the forecast weight removed this way.
    degen <- cN < 1e-8
    W[degen, ] <- 1 / N
    cN[degen]  <- 1 - 1 / N
    dropF <- matrix(degen, G, 2)       # r = (j - 1) * G + g  ->  [g, j]
    
    ia <- rep(1:3, 3); ib <- rep(1:3, each = 3)   # column (a, b) of a flattened 3 x 3
    for (k in 0:D) {
      sim <- sim0
      if (k > 0) {
        sim <- disturbance.p.rows(sim0, dist$vals[[k]]$mu0, dist$vals[[k]]$V0, dist$alloc.soil)
      }

      MU <- W %*% sim                              # P x 3, every mean in one matmul

      # every weighted covariance in one matmul (rows of W sum to 1):
      #   S_r = (sum_n W[r,n] x_n x_n' - MU_r MU_r') / cN_r,
      # with x centred on the ensemble mean first to limit cancellation
      ctr <- colMeans(sim)
      Xc  <- sweep(sim, 2, ctr)
      MUc <- sweep(MU, 2, ctr)
      S   <- (W %*% (Xc[, ia] * Xc[, ib]) - MUc[, ia] * MUc[, ib]) / cN   # P x 9
      if (k == 0) {
        S <- sweep(S, 2, as.vector(Q), "+")
        S[, 9] <- S[, 9] - Q[3, 3] + 1 / rep(tauGrid, 2)    # S[3, 3]: process var 1/tau_g
      }
      # r = (j - 1) * G + g runs g fastest, so P x 3 / P x 9 fill [g, j, , ] in order
      muF[, , k + 1, ]    <- MU
      sigF[, , k + 1, , ] <- S
    }

    # bias correction: shift the undisturbed wood forecast mean by b_t (raw copy kept
    # for the residual r_t, which is measured against the uncorrected forecast)
    muF0raw <- muF[, , 1, 3]                       # G x 2
    if (!is.null(bias)) muF[, , 1, 3] <- muF0raw + bHat
    hist$bias[t] <- bHat
    
    # compute analysis
    lA <- array(0, dim=c(G, 2, D+1))
    muA <- array(0, dim=c(G, 2, D+1, 3))
    sigA <- array(0, dim=c(G, 2, D+1, 3, 3))
    if (is.na(Y[t, 3])) {
      # no observation: analysis = forecast, flat likelihood (weights move only by the prior p)
      muA  <- muF
      sigA <- sigF
      lA[] <- 1
    } else {
      R <- Y[t, 3] * Rparams[1] + Rparams[2] # heteroskedastic obs error
      # KalmanUpdate for every (g, j, k) at once. H = (0, 0, 1) and R is scalar, so with
      # s = C33 + R:  K = C[, 3] / s,  m_hat = m + K (y - m3),  C_hat = C - K C[3, ]
      s    <- sigF[, , , 3, 3] + R                                  # G x 2 x (D+1)
      K    <- sigF[, , , , 3] / as.vector(s)                         # G x 2 x (D+1) x 3
      muA  <- muF + K * as.vector(Y[t, 3] - muF[, , , 3])
      for (a in 1:3) for (b in 1:3) sigA[, , , a, b] <- sigF[, , , a, b] - K[, , , a] * sigF[, , , 3, b]
      lA[] <- dnorm(Y[t, 3], mean = muF[, , , 3], sd = sqrt(s))
    }
    for (k in 0:D) lA[, , k + 1][dropF] <- 0   # degenerate forecast components (see above)
    
    # compute tau posterior on grid
    beta <- array(0, dim=c(D+1))
    for (k in 0:D) {
      beta[k+1] = dist$p[k+1,t]
    }

    # PIT of y_t under its one-step-ahead predictive, a mixture over (g, j, k) with
    # weights pi_{t-1}(g) * prevW[g,j] * beta_k (degenerate components dropped) and
    # components N(muF[g,j,k,3], sigF[g,j,k,3,3] + R); for k = 0, sigF[,,1,3,3] includes
    # the process variance 1/tau_g. Left NA on forecast-only steps with no observation.
    piPrev <- exp(qs - max(qs)); piPrev <- piPrev / sum(piPrev)            # pi_{t-1}(g)
    hist$wDrop[t] <- sum(piPrev * rowSums(prevW * dropF))                  # dropped weight
    if (!is.na(Y[t, 3])) {
      wF <- outer(outer(piPrev, rep(1, 2)) * prevW, beta) * as.vector(!dropF)  # G x 2 x (D+1)
      wF <- wF / sum(wF)
      hist$pit[t] <- sum(wF * pnorm(Y[t, 3], muF[, , , 3], sqrt(sigF[, , , 3, 3] + R)))
      hist$predMean[t]    <- sum(wF * muF[, , , 3])
      hist$predMeanRaw[t] <- hist$predMean[t] - bHat * sum(wF[, , 1])
      hist$predVar[t]     <- sum(wF * (sigF[, , , 3, 3] + R + muF[, , , 3]^2)) - hist$predMean[t]^2
      hist$predProcVar[t] <- sum(wF[, , 1] / tauGrid)    # k = 0 components carry 1/tau_g
      hist$predR[t]       <- R
    }
    wPrior <- outer(prevW, beta)                     # G x 2 x (D+1): prevW[g,j] * beta_k
    zsum   <- apply(wPrior * lA, 1, sum)             # marginal likelihood of y_t given tau_g
    logz   <- log(zsum)
    qs     <- qs + logz
    if (any(!is.finite(logz))) stop(sprintf("t=%d: zero or non-finite marginal likelihood at %d tau grid points", t, sum(!is.finite(logz))))
    
    # tau posterior on the grid, pi_t(g)
    m <- max(qs)
    ws <- exp(qs - m) / sum(exp(qs - m))
    hist$procVar[t] <- sum(ws / tauGrid)
    
    # analysis weights
    wA <- wPrior * lA / zsum

    # bias update from y_t (applied from t + 1 on). Posterior weights of the undisturbed
    # components, tw_{gj0} = pi_t(g) wA[g, j, 0], which equals
    # pi_{t-1}(g) prevW[g,j] beta_0 N(y_t; muF + b, sigF + R) / normaliser.
    if (!is.null(bias) && !is.na(Y[t, 3])) {
      tw0 <- ws * wA[, , 1]                        # G x 2
      wt  <- sum(tw0)                              # P(z_t = 0 | y_{1:t})
      if (wt > 0) {
        rt <- sum(tw0 * (Y[t, 3] - muF0raw)) / wt
        bObs <- bObs + 1
        burn <- bObs <= bias$skip + bias$burnin
        if (burn) wt <- 1                          # burn-in: full weight
        hist$biasResid[t] <- rt
        hist$biasW[t]     <- wt
        if (bObs > bias$skip && (burn || wt >= bias$wMin)) {
          bS <- bias$rho * bS + wt * rt
          bN <- bias$rho * bN + wt
          bHat <- if (bias$denom == "N") bS / bN else bS / (bObs - bias$skip)
        }
      }
    }
    
    # recompute prev posterior vals (KL projection), for every tau grid point at once
    out     <- project_grid(wA, muA, sigA)
    prevW   <- out$w
    prevMu  <- out$mu
    prevSig <- out$sig
    
    # store analysis values for this tstep, marginalised over the tau grid (weights
    # pi_t(g) = ws) and the components (j, k) (weights wA[g, , ], which sum to 1 for each g)
    pDist.g <- apply(wA[, , -1, drop = FALSE], 1, sum)                  # P_t(g)
    mean.g  <- apply(wA * muA[, , , 3], 1, sum)                           # E[wood | g]
    msq.g   <- apply(wA * (sigA[, , , 3, 3] + muA[, , , 3]^2), 1, sum)   # E[wood^2 | g]
    hist$mean[t]  <- sum(ws * mean.g)
    hist$var[t]   <- sum(ws * msq.g) - hist$mean[t]^2
    hist$pDist[t] <- sum(ws * pDist.g)
    hist$ci95[t, ] <- mix_quantile(c(0.025, 0.975), ws * wA, muA[, , , 3], sqrt(sigA[, , , 3, 3]))
  }
  
  return(hist)
}

```

```{r}
## Generate ORJointUpdate inputs for each requested site, run the joint update,
## and save each site's diagnostics to experiments/jointUpdate. Returns the
## results invisibly (keyed by site) for convenience.
## Rparams.source ("GEDI"/"LandTrendr") and Rparams.model ("het"/"con") describe the
## observation-error parameters in use; together with tauPrior ("gamma"/"loguniform")
## they are recorded in each output filename.
run_joint_update_sites <- function(site.idx, Rparams = Rparams.GEDI.het,
                                   useDistProduct = TRUE,
                                   tauPrior = c("gamma", "loguniform"),
                                   Rparams.source = c("GEDI", "LandTrendr"),
                                   Rparams.model  = c("het", "con"),
                                   outdir = "experiments/jointUpdate1990",
                                   bias = NULL) {
  Rparams.source <- match.arg(Rparams.source)
  Rparams.model  <- match.arg(Rparams.model)
  tauPrior       <- match.arg(tauPrior)
  if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
  results <- list()
  for (site in site.idx) {
    message("running joint update for site ")
    # function(Y, Rparams, Q, PAR, Xic, THETA, dist, G = 200, N=100, prior)
    args <- make_or_args(site, useDistProduct, tauPrior)
    out <- ORJointUpdate(Y = args$Y, Rparams = Rparams,
                         Q = Q, PAR = args$PAR, Xic = ICprior,
                         THETA = theta[1:11], dist = args$disturb.params, prior=args$prior,
                         grid = args$grid, bias = bias)
    out$year <- atime                  # calendar year of each DA timestep
    out$obs  <- args$Y[, 3]            # assimilated obs, aligned with out$year (NA = none)
    fname <- file.path(outdir,
                       sprintf("site_%d_distprod_%s_%s_%s_tau%s.rds",
                               site, useDistProduct, Rparams.source, Rparams.model, tauPrior))
    if (!is.null(bias)) fname <- sub("\\.rds$", "_bias.rds", fname)
    out$biasConfig <- bias
    saveRDS(out, fname)
    results[[as.character(site)]] <- out
  }
  invisible(results)
}
```

```{r run-single}
## Run + save for the site(s) of interest.
## tauPrior = "gamma" or "loguniform" selects the prior on the process precision tau.
run_joint_update_sites(site.idx = c(300), Rparams = Rparams.GEDI.het, useDistProduct = TRUE,
                       tauPrior = "loguniform", Rparams.source = "GEDI", Rparams.model = "het")
```

## experiment grid

Runs every combination of observation-error model (GEDI / LandTrendr magnitude x
heteroskedastic / constant R), disturbance product (with / without) and tau prior
(gamma / log-uniform), i.e. 16 experiments, over every site in `dextract`. Each
(experiment, site) run is saved to `grid.outdir` with the usual
`site_<idx>_distprod_<TF>_<source>_<model>_tau<prior>.rds` name, which
`OR_joint_update_analysis.Rmd` parses. Runs are spread over `grid.cores` forked
workers; existing files are skipped unless `grid.overwrite = TRUE`, so an interrupted
run can be restarted and picks up where it left off. Progress (% finished, failures,
elapsed time, ETA) is printed every time a run completes.

```{r run-grid}
grid.outdir    <- "experiments/fullRun"
grid.tauPriors <- "gamma"
grid.overwrite <- FALSE
grid.cores     <- max(1, parallel::detectCores() - 2)
grid.sites     <- seq_len(nrow(dextract))   # all sites

## observation-error models: label -> (Rparams, source, model)
Rparams.grid <- list(
  GEDI_het       = list(Rparams = Rparams.GEDI.het,         source = "GEDI",       model = "het"),
  GEDI_con       = list(Rparams = Rparams.GEDI.fixed,       source = "GEDI",       model = "con"),
  LandTrendr_het = list(Rparams = Rparams.Landtrendr.het,   source = "LandTrendr", model = "het"),
  LandTrendr_con = list(Rparams = Rparams.Landtrendr.fixed, source = "LandTrendr", model = "con")
)
exp.grid <- expand.grid(R = names(Rparams.grid), useDistProduct = c(TRUE, FALSE),
                        tauPrior = grid.tauPriors, site = grid.sites, stringsAsFactors = FALSE)
exp.grid$file <- with(exp.grid, file.path(grid.outdir, sprintf(
  "site_%d_distprod_%s_%s_%s_tau%s.rds", site, useDistProduct,
  sapply(Rparams.grid[R], `[[`, "source"), sapply(Rparams.grid[R], `[[`, "model"), tauPrior)))
if (!dir.exists(grid.outdir)) dir.create(grid.outdir, recursive = TRUE)

run_grid_row <- function(i) {
  e <- exp.grid[i, ]
  r <- Rparams.grid[[e$R]]
  set.seed(i)   # per-run seed: results don't depend on which worker / in what order
  ## catch errors per run so one failure doesn't take down the worker
  tryCatch({
    run_joint_update_sites(e$site, Rparams = r$Rparams, useDistProduct = e$useDistProduct,
                           tauPrior = e$tauPrior, Rparams.source = r$source,
                           Rparams.model = r$model, outdir = grid.outdir)
    e$file
  }, error = function(err) structure(conditionMessage(err), class = "try-error"))
}

fmt_secs <- function(s) {
  if (!is.finite(s)) return("--")
  s <- round(s); sprintf("%dh%02dm%02ds", s %/% 3600, (s %% 3600) %/% 60, s %% 60)
}

## Worker pool: keep grid.cores forked runs going and collect them as they finish, so
## the main process can report progress after every completed run.
n.total  <- nrow(exp.grid)
grid.out <- vector("list", n.total)
todo     <- if (grid.overwrite) seq_len(n.total) else which(!file.exists(exp.grid$file))
grid.out[setdiff(seq_len(n.total), todo)] <- exp.grid$file[setdiff(seq_len(n.total), todo)]
n.skip   <- n.total - length(todo)
n.done   <- n.skip; n.fail <- 0
running  <- list()   # pid -> list(job, row)
t0       <- Sys.time()

report <- function() {
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  n.new   <- n.done - n.skip
  eta     <- if (n.new > 0) elapsed / n.new * (n.total - n.done) else NA
  message(sprintf("[%s] %d / %d runs finished (%.1f%%) | %d failed | %d running | elapsed %s | ETA %s",
                  format(Sys.time(), "%H:%M:%S"), n.done, n.total, 100 * n.done / n.total,
                  n.fail, length(running), fmt_secs(elapsed), fmt_secs(eta)))
}

message(sprintf("experiment grid: %d runs (%d sites x %d settings), %d already done, %d workers",
                n.total, length(grid.sites), n.total / length(grid.sites), n.skip, grid.cores))
report()
while (length(todo) || length(running)) {
  while (length(running) < grid.cores && length(todo)) {
    i   <- todo[1]; todo <- todo[-1]
    job <- parallel::mcparallel(run_grid_row(i), silent = TRUE)
    running[[as.character(job$pid)]] <- list(job = job, row = i)
  }
  res <- parallel::mccollect(lapply(running, `[[`, "job"), wait = FALSE, timeout = 5)
  for (pid in names(res)) {
    i   <- running[[pid]]$row
    out <- res[[pid]]
    if (is.null(out)) out <- structure("worker died (no result returned)", class = "try-error")
    grid.out[[i]] <- out
    n.done <- n.done + 1
    if (inherits(out, "try-error")) {
      n.fail <- n.fail + 1
      message(sprintf("  FAILED: site %d, %s, distprod %s, tau %s: %s", exp.grid$site[i],
                      exp.grid$R[i], exp.grid$useDistProduct[i], exp.grid$tauPrior[i], out))
    }
    running[[pid]] <- NULL
  }
  if (length(res)) report()
}

failed <- which(sapply(grid.out, inherits, "try-error"))
if (length(failed)) print(cbind(exp.grid[failed, c("R", "useDistProduct", "tauPrior", "site")],
                                error = sapply(grid.out[failed], as.character)))
```

## analysis

```{r}
## Load a saved ORJointUpdate run. The site index and run settings are parsed from
## the file name (site_<idx>_distprod_<TF>_<source>_<model>_tau<prior>.rds). Runs
## store their own year axis ($year) and aligned observations ($obs); runs made
## before that (which put 1990 biomass at a 1985 timestep) are rejected.
load_hist <- function(file) {
  h     <- readRDS(file)
  tag   <- sub("\\.rds$", "", basename(file))
  site  <- as.integer(sub("^site_(\\d+)_.*$", "\\1", tag))
  if (is.null(h$year)) stop(file, ": no $year axis -- stale run from before the 1990 alignment fix; rerun it")
  h$site <- site
  h$tag  <- tag
  h
}

## shared look: one palette, light grid, no box
.pal <- list(mean = "#2a78d6", band = "#2a78d6", obs = "#0b0b0b", dist = "#eb6834",
             grid = "#e6e6e3", ink = "#52514e")
.base_par <- function() par(mar = c(4.2, 4.8, 3.2, 1), las = 1, bty = "n",
                            col.axis = .pal$ink, fg = .pal$ink, cex.axis = 0.9)

## Wood-pool analysis: mean (line), 95% interval mean +/- 1.96 sd (band), observations (points)
plot_analysis <- function(h, main = sprintf("Site %d: wood-pool analysis", h$site)) {
  op <- .base_par(); on.exit(par(op))
  x  <- h$year; sd <- sqrt(pmax(h$var, 0))
  lo <- h$mean - 1.96 * sd; hi <- h$mean + 1.96 * sd
  plot(NA, xlim = range(x), ylim = range(c(lo, hi, h$obs), na.rm = TRUE),
       xlab = "year", ylab = expression("wood carbon (kg C m"^-2*")"), main = main)
  abline(h = axTicks(2), col = .pal$grid)
  polygon(c(x, rev(x)), c(lo, rev(hi)), col = adjustcolor(.pal$band, 0.2), border = NA)
  lines(x, h$mean, col = .pal$mean, lwd = 2.5)
  points(x, h$obs, pch = 21, bg = .pal$obs, col = "white", cex = 1.1)
  legend("topright", bty = "n", cex = 0.85, text.col = .pal$ink,
         legend = c("analysis mean", "95% interval", "observation"),
         col = c(.pal$mean, adjustcolor(.pal$band, 0.35), .pal$obs),
         lwd = c(2.5, 10, NA), pch = c(NA, NA, 16))
  invisible(h)
}

## Posterior probability of disturbance per year (bars), with the 0.5 line
plot_pdist <- function(h, main = sprintf("Site %d: probability of disturbance", h$site)) {
  op <- .base_par(); on.exit(par(op))
  x <- h$year; p <- h$pDist
  plot(NA, xlim = range(x) + c(-0.6, 0.6), ylim = c(0, 1), xlab = "year",
       ylab = "P(disturbed)", main = main)
  abline(h = seq(0, 1, 0.25), col = .pal$grid)
  abline(h = 0.5, col = .pal$ink, lty = 2)
  rect(x - 0.38, 0, x + 0.38, p, col = .pal$dist, border = NA)
  big <- which(p >= 0.5)
  if (length(big)) text(x[big], p[big], sprintf("%.2f", p[big]), pos = 3, cex = 0.75,
                        col = .pal$ink, xpd = NA)
  invisible(h)
}
```

```{r, fig.width = 9, fig.height = 4.5}
h <- load_hist("experiments/jointUpdate1990/site_300_distprod_TRUE_GEDI_het_tauloguniform.rds")
plot_analysis(h)
plot_pdist(h)
```
