---
title: "RelabelDisturbances"
output: html_document
author: "Jacob Epstein"
date: 1/9/26
---

## Goal:

## - Find disturbance years as changepoints in biomass timeseries
## - Generate labels for each year based on disturbance class
## - Derive disturbance model priors based on the distribution of % biomass 
##    retained for each kind of disturbance
 
## Disturbance classes: 
```{r}
dist_name <- c(
  "0"  = "none",
  "10" = "other",
  "20" = "cut",
  "30" = "road",
  "40" = "fire",
  "50" = "biotic"
)
```

```{r}
knitr::opts_chunk$set(echo = TRUE)
source("disturbance.R")
source("multi.ens.adj.R")
source("multi.analytical.ens.adj.R")
source("StatsAuxiliary.R")
library(BayesianTools)
library(magrittr)
library(mclust)
```

```{r}
## Load disturbance rate, state&transition, disturbed biomass stats
load("STprior.RData")
load("extract.h03v04.RData")
dtime <- 1985:2012
colnames(dextract) <- paste0("t",dtime)
dextract[is.na(dextract)] <- 0
d.prior["0"] = d.prior["0"] + d.prior[6]                         ## change at runtime (removes <NA>)
d.prior = d.prior[-6]

## Load Calibration
load("fit.RData")
Qsoil = Qs
```

```{r}
## BIOMASS observations
B2C = 0.5 ##biomass to carbon ratio
Yw <- bextract*B2C*1000*1e-4 ## convert Mg/ha (Mg per hectacre)  to  kg C
```

```{r}
pdf("./experiments/exploration/disturbance-years-labelled.pdf")
nsite <- dim(Yw)[1]
nyear <- dim(dextract)[2]
disttypes <- seq(0, 50, by=10)
distlabels <- matrix(data = 0, nrow = nsite, ncol = nyear)
t04 <- which(dtime == 2004)

for (i in 1:nsite) {
  ## compute the disturbances in the time series for the site
  pctchanges <- as.numeric(Yw[i, 2:nyear] / Yw[i, 1:nyear-1])
  nchange <- length(pctchanges)
  changediffs <- c()
  for (t in 1:nchange) {
    # find average in 3 year window around t
    tmin <- max(1, t-3)
    tmax <- min(nchange, t+3)
    
    if (t < tmax) {
      changeavg <- mean(c(pctchanges[tmin:(t-1)], pctchanges[(t+1):tmax])) 
    } else {
      changeavg <- mean(pctchanges[tmin:(t-1)])
    }
    changediff <- abs(pctchanges[t] / changeavg)
    changediffs <- c(changediffs, changediff)
  }
  disidxs <- which(changediffs < 0.80) # get disturbance idxs
  disidxs <- Filter(function(idx) {
    yidx <- Yw[i, idx]
    dyidx <- Yw[i, idx + 1] - Yw[i, idx]
    return(dyidx < -0.75 & yidx > 0.75); # filter indices for which the decrease or starting biomass is too small
  }, disidxs)
  plot(Yw[i,], ylim=c(0, 20), type="l", main=paste("site", i))
  abline(v = disidxs, col = "red", lwd = 2, lty = 2)
  
  ## label the disturbances
  for (idx in disidxs) {
    # default to disturbance product's label if it labelled a disturbance
    # else search for a label in a 3 year window (6 years total) around the target year
    # else default to disturbance product label for target year (2004) if it was labelled a disturbance
    # else, assign to class "other"
    product.labels <- dextract[i,idx:min(idx+1,nyear)]
    product.labels.positive <- which(product.labels > 0)
    product.labels.window <- dextract[i,max(0, idx-3):min(idx+3, nyear)]
    window.positive <- which(product.labels.window > 0)
    if (length(product.labels.positive) > 0) {
      distlabels[i,idx] <- product.labels[product.labels.positive[1]]
    } else if (length(window.positive) > 0) {
      distlabels[i,idx] <- product.labels.window[window.positive[1]]
    } else if (dextract[i, t04] > 0) {
      distlabels[i,idx] <- dextract[i, t04]
    } else {
      distlabels[i, idx] <- 10 # "other"
    }
  }

}
dev.off()
```

## visualize distributions of percent biomass retained
```{r}
pdf("./experiments/exploration/disturbance-size-across-years-from-metric.pdf")
dist_name <- c(
  "0"  = "none",
  "10" = "other",
  "20" = "cut",
  "30" = "road",
  "40" = "fire",
  "50" = "biotic"
)
nyear <- dim(dextract)[2]
binwidth <- 0.025

## density plot for each disturbance type of percent biomass retained
for (disttype in seq(0, 50, by=10)) {
  dist.indices <- which(distlabels == disttype, arr.ind=TRUE)
  dist.pct.all <- c()
  nidx <- dim(dist.indices)[1]
  if (nidx > 0) {
    for (i in 1:nidx) {
      idx <- dist.indices[i,]
      if (idx[2] < nyear) {
        dist.pct <- as.numeric(Yw[idx[1], idx[2] + 1] / Yw[idx[1], idx[2]])
        if (!is.na(dist.pct) & 0 < dist.pct & dist.pct < 1.1) {
          dist.pct.all <- c(dist.pct.all, dist.pct) 
        }
      }
    }
  }
  hist(as.numeric(dist.pct.all), prob = TRUE, breaks=seq(0, 1.1, by=0.025), xlim=c(0, 1.1), ylim=c(0, 20), xlab="% Cw retained after decrease", main=paste(
    dist_name[as.character(disttype)]
  ))
}
dev.off()
```

## load a larger number of disturbed pixels
```{r}
# load("state-transition-saved.RData")

## sample disturbances for target year
nsamp = 15000
samp = raster::sampleStratified(d04,size = nsamp,exp=1000) # nsamp per class

## eliminate disturbances that did not apply to forests
c03 <- covS[[which(ctime==2003)]]
csamp <- c03[samp[,1]]
fsamp <- samp[csamp==42,]

## print total number of samples for each class
for (c in seq_along(dclass)) {
  nc <- length(which(fsamp[,2] == dclass[c]))
  print(paste("class", dclass[c], ":", nc, "samples"))  
}

# extract disturbance / biomass data
dextract.large = raster::extract(disR,fsamp[,1])
bextract.large = raster::extract(bioS,fsamp[,1])
B2C = 0.5
Yw.large <- bextract.large*B2C*1000*1e-4
```

## relabel the disturbances
```{r}
nsite <- nrow(dextract.large)
nyear <- ncol(dextract.large)
disttypes <- seq(0, 50, by=10)
distlabels.large <- matrix(data = 0, nrow = nsite, ncol = nyear)

for (i in 1:nsite) {
  ## compute the disturbances in the time series for the site
  pctchanges <- as.numeric(Yw.large[i, 2:nyear] / Yw.large[i, 1:nyear-1])
  nchange <- length(pctchanges)
  changediffs <- c()
  for (t in 1:nchange) {
    # find average in 3 year window around t
    tmin <- max(1, t-3)
    tmax <- min(nchange, t+3)
    
    if (t < tmax) {
      changeavg <- mean(c(pctchanges[tmin:(t-1)], pctchanges[(t+1):tmax])) 
    } else {
      changeavg <- mean(pctchanges[tmin:(t-1)])
    }
    changediff <- abs(pctchanges[t] / changeavg)
    changediffs <- c(changediffs, changediff)
  }
  disidxs <- which(changediffs < 0.80) # get disturbance idxs
  disidxs <- Filter(function(idx) {
    yidx <- Yw.large[i, idx]
    dyidx <- Yw.large[i, idx + 1] - Yw.large[i, idx]
    return(dyidx < -0.75 & yidx > 0.75); # filter indices for which the decrease or starting biomass is too small
  }, disidxs)
  
  ## label the disturbances
  for (idx in disidxs) {
    # default to disturbance product's label if it labelled a disturbance
    # else search for a label in a 3 year window (6 years total) around the target year
    # else default to disturbance product label for target year (2004) if it was labelled a disturbance
    # else, assign to class "other"
    product.labels <- dextract.large[i,idx:min(idx+1,nyear)]
    product.labels.positive <- which(product.labels > 0)
    product.labels.window <- dextract.large[i,max(0, idx-3):min(idx+3, nyear)]
    window.positive <- which(product.labels.window > 0)
    label04 <- dextract.large[i, t04]
    # save(val, file="debugging-or.RData")
    if (length(product.labels.positive) > 0) {
      distlabels.large[i,idx] <- product.labels[product.labels.positive[1]]
    } else if (length(window.positive) > 0) {
      distlabels.large[i,idx] <- product.labels.window[window.positive[1]]
    } else if (!is.na(label04) & label04 > 0) {
      distlabels.large[i,idx] <- dextract.large[i, t04]
    } else {
      distlabels.large[i, idx] <- 10 # "other"
    }
  }
}
```


## generate disturbance model priors for each disturbance type
```{r}
dist.priors <- list()

## fit a mixture of normals to each density plot
for (disttype in c(10, 20, 40, 50)) {
  # compute array of percent retained for that disturbance type
  dist.indices <- which(distlabels.large == disttype, arr.ind=TRUE)
  nidx <- dim(dist.indices)[1]
  dist.pct.all <- c()
  for (i in 1:nidx) {
    idx <- dist.indices[i,]
    if (idx[2] < nyear) {
      dist.pct <- as.numeric(Yw.large[idx[1], idx[2] + 1] / Yw.large[idx[1], idx[2]])
      if (!is.na(dist.pct) & 0 < dist.pct & dist.pct < 1.1) {
        dist.pct.all <- c(dist.pct.all, dist.pct) 
      }
    }
  }
  
  # fit a mixture of normals to the distribution of percent retained
  fit <- mclust::Mclust(dist.pct.all)
  
  # plot the fitted distribution / density plot of percent retained
  hist(
    as.numeric(dist.pct.all),
    prob = TRUE,
    breaks = seq(0, 1.1, by = 0.025),
    xlim = c(0, 1.1),
    ylim = c(0, 20),
    xlab = "% Cw retained after decrease",
    main = paste(dist_name[as.character(disttype)])
  )
  
  xgrid <- seq(0, 1.1, length.out = 500)
  
  # Print stats
  # print(fit$BIC)
  
  # Extract parameters
  weights <- fit$parameters$pro
  means   <- fit$parameters$mean
  sds     <- sqrt(fit$parameters$variance$sigmasq)
  
  if (length(sds) < length(weights)) {
    sds <- rep(sds[1], length(weights))
  }
  
  mix_density <- rowSums(
    sapply(seq_along(weights), function(k) {
      weights[k] * dnorm(xgrid, means[k], sds[k])
    })
  )
  
  # store parameters
  dist.priors[[as.character(disttype)]]$weights <- weights
  dist.priors[[as.character(disttype)]]$means <- means
  dist.priors[[as.character(disttype)]]$vars <- sds
  
  lines(xgrid, mix_density, col = "red", lwd = 2)
}

save(distlabels, dist.priors, file="dis.postprocessed.RData")
```





